---
title: "Lecture 14: Agreement on Agriculture"
subtitle: "ECO 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "July 25, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    footer: "ECO 2203 | International Trade and Policy in Agriculture"
    code-fold: true
    code-summary: "Show R code"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

## Recap: Lecture 13 {.smaller}

**Lecture 13** established the WTO as the institutional framework of international trade.

::: {.columns}
::: {.column width="55%"}
**Key ideas from Lecture 13:**

- GATT → WTO (1995): from provisional to permanent
- Five core principles: MFN, National Treatment, Bound Tariffs, Reciprocity, Transparency
- DSU: binding dispute settlement
- India's Peace Clause — protects MSP procurement
- Doha Round stalemate: agriculture at the core
:::
::: {.column width="45%"}
::: {.highlightbox}
**Today's Focus**

We zoom into the **Agreement on Agriculture (AoA)** — the specific WTO text that governs how much protection India can give its farmers, how much support the US/EU can give theirs, and how agricultural trade is managed globally.

Three pillars: **Market Access · Domestic Support · Export Competition**
:::
:::
:::

---

## Why Was the AoA Needed? {.smaller}

::: {.columns}
::: {.column width="55%"}
**Pre-Uruguay Round Agriculture: A Wild West**

Before 1995, agriculture was essentially **outside GATT discipline**:

- USA had a 1955 GATT waiver exempting its agricultural support programs
- EU's Common Agricultural Policy (CAP): massive production and export subsidies
- Result: **global overproduction** of wheat, butter, sugar; chronic dumping on world markets

::: {.methodbox}
**The dumping problem:**
In the 1980s, EU butter was sold on world markets at 20% of EU domestic prices. US wheat exports were subsidised by up to 40%. This depressed world prices and **hurt farmers in Australia, Argentina, Brazil, India, and the developing world**.
:::
:::
::: {.column width="45%"}
**Why This Hurt India**

::: {.keybox}
Artificially low world prices for wheat and rice:
- Made Indian exports uncompetitive
- Provided cheap food imports that undercut domestic prices
- Reduced incentives for agricultural investment globally

The AoA's promise to the developing world:
- **Discipline** US/EU subsidies
- **Reduce** import barriers
- **Eliminate** export subsidies

Reality: progress has been slow — but the framework now exists.
:::
:::
:::

---

## AoA Structure: The Three Pillars + S&DT {.smaller}

::: {.keybox}
**Agreement on Agriculture (1995) — 21 Articles + Annexes**
In force: January 1, 1995 | Implementation period: 1995–2004 (developed), 1995–2005 (developing)
:::

::: {.columns}
::: {.column width="50%"}
**Agreement on Agriculture (1995)**

- **Pillar 1: Market Access**
  - Tariffication of non-tariff barriers
  - Bound tariff schedules
  - Tariff reduction commitments (average 36%)
  - Minimum access provisions (3–5% of domestic consumption)
  - Special Safeguard (SSG) clause

- **Pillar 2: Domestic Support**
  - Aggregate Measurement of Support (AMS) reduction (20% for developed, 13.3% for developing)
  - Green Box: non-distorting (exempt)
  - Blue Box: production-limiting (exempt)
  - Amber Box: trade-distorting (limited)
  - De minimis provisions

- **Pillar 3: Export Competition**
  - Export subsidy reduction/elimination
  - Export credit disciplines
:::
::: {.column width="50%"}
**Special and Differential Treatment (S&DT)**

::: {.methodbox}
Developing countries get:
- Longer implementation periods (10 vs. 6 years)
- Smaller reduction commitments (24% vs. 36% tariff cuts; 13% vs. 24% domestic support cuts)
- Higher de minimis thresholds (10% vs. 5%)
- Continued input subsidies for poor farmers
- Continued investment subsidies for agricultural development

**India fully utilises all S&DT provisions**
:::
:::
:::

---

# Pillar 1 — Market Access

---

## Tariffication: Converting NTBs to Tariffs {.smaller}

::: {.columns}
::: {.column width="55%"}
**What is Tariffication?**

Before 1995: Countries protected agriculture through:
- Import quotas (how much can come in)
- Variable levies (duty adjusts to keep domestic prices stable)
- Minimum import prices
- Voluntary export restraints

AoA Article 4: **All these NTBs converted to tariff equivalents** (bound tariff rates in WTO schedules) by 1995.

**The Reduction Commitment:**

| Country | Average Cut | Minimum/Product | Period |
|---------|-------------|-----------------|--------|
| Developed | 36% | 15% | 6 years |
| Developing | 24% | 10% | 10 years |
| LDCs | None | None | — |
:::
::: {.column width="45%"}
::: {.highlightbox}
**"Dirty Tariffication"**

Many developed countries set artificially **high tariff equivalents** when converting their NTBs — overstating the protection level to preserve policy space.

Example: Canada's tariff on butter: **298%**; on chicken: **238%**

This gave rich countries huge "water in the tariff" even after the required cuts.

India negotiated high bound rates (113% average) for similar reasons — policy space for future flexibility.
:::
:::
:::

---

## AoA Tariff Reduction Commitments {.center}

```{r}
#| label: fig-aoa-tariff-reduction
#| fig-cap: "AoA Tariff Reduction Commitments: Developed vs Developing\nSource: WTO, Agreement on Agriculture; UNCTAD."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(tidyr)

df_tar <- data.frame(
  round_num  = c(0, 1, 2, 3),
  round_lbl  = c("Base\n(1986-88)", "Uruguay Round\n(1995)",
                 "Doha\n(stalled)", "Target\n(proposed)"),
  Developed  = c(0, 36, 36, 54),
  Developing = c(0, 24, 24, 36)
)

df_long <- pivot_longer(df_tar, cols = c(Developed, Developing),
                        names_to = "Group", values_to = "Reduction")

ggplot(df_long, aes(x = round_num, y = Reduction,
                    color = Group, group = Group)) +
  geom_step(linewidth = 1.2) +
  geom_point(size = 3) +
  annotate("rect", xmin = 1.55, xmax = 2.45, ymin = -1, ymax = 58,
           fill = "#C84B31", alpha = 0.08) +
  annotate("text", x = 2, y = 54,
           label = "Doha stalled\n(no progress)",
           color = "#C84B31", size = 3, fontface = "italic") +
  scale_color_manual(values = c("Developed" = "#012169", "Developing" = "#B9975B")) +
  scale_x_continuous(
    breaks = 0:3,
    labels = df_tar$round_lbl
  ) +
  scale_y_continuous(limits = c(-2, 62)) +
  labs(
    title = "AoA Tariff Reduction Commitments: Developed vs Developing",
    subtitle = "Average percentage cut from Uruguay Round base tariffs",
    x = NULL, y = "Average Tariff Cut (%)", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## Tariff Rate Quotas (TRQs) {.smaller}

::: {.columns}
::: {.column width="55%"}
**How TRQs Work**

Where post-tariffication tariffs were extremely high, **TRQs** were created to ensure minimum market access:

- **In-quota tariff:** lower rate (e.g., 15%) applied to a specified quantity
- **Out-of-quota tariff:** higher rate (e.g., 60%) applied to quantities above the quota

```
Imports
  ↑
  │ Out-of-quota tariff = 60%
  │ ───────────────────────────
  │ In-quota tariff = 15%     ← TRQ fills here
  └─────────────────────────→ Volume
        Quota quantity
```

**Minimum access commitment:** 5% of domestic consumption by end of implementation period
:::
::: {.column width="45%"}
**India's TRQ Commitments**

::: {.methodbox}
India has TRQs for:

| Product | In-quota rate | Out-of-quota |
|---------|--------------|--------------|
| Wheat | 50% | 100% |
| Rice (non-PDS) | 70% | 80% |
| Maize | 15% | 60% |
| Sunflower oil | 100% | 100% |
| Cotton | 5% | 10% |
:::

::: {.keybox}
**Most Indian TRQ quotas are never filled** — because even the in-quota tariff is high enough to limit imports below the quota. India's TRQs are symbolic: the real protection is the bound tariff, not the TRQ.
:::
:::
:::

---

## India's Bound Tariffs: Agricultural Products {.smaller}

::: {.columns}
::: {.column width="50%"}
**"Water in the Tariff" — India's Policy Flexibility**

| Product | Bound (%) | Applied (2024, %) | Water (%) |
|---------|-----------|-------------------|-----------|
| Rice | 80 | 25 | **55** |
| Wheat | 100 | 40 | **60** |
| Maize | 60 | 50 | **10** |
| Sugar | 150 | 100 | **50** |
| Crude palm oil | 300 | 100 | **200** |
| Refined palm oil | 300 | 100 | **200** |
| Chickpeas | 100 | 50 | **50** |
| Milk powder (SMP) | 60 | 30 | **30** |
| Onions | 100 | 0 | **100** |
:::
::: {.column width="50%"}
::: {.keybox}
**How India Used This Space — Palm Oil**

In 2018–19, domestic oilseed farmers (soybean, groundnut, sunflower) were hurt by cheap imported palm oil from Malaysia and Indonesia.

India **raised the applied tariff** on crude palm oil from 30% to 100% (still below bound 300%).

This protected domestic edible oil industry without violating WTO — no dispute case possible.

Then in 2021, India **lowered it again** to reduce consumer food prices during inflation. Flexible policy instrument.
:::

::: {.highlightbox}
The large tariff "water" is India's most important defensive trade policy instrument for agriculture
:::
:::
:::

---

## Special Safeguard (SSG) vs. SSM {.smaller}

::: {.columns}
::: {.column width="55%"}
**Special Safeguard (SSG) — AoA Article 5**

Allows additional tariff duties **automatically** (no injury investigation needed) when:
- **Volume trigger:** imports exceed 125% of average imports in previous 3 years; OR
- **Price trigger:** import price falls more than 10% below average 1986–88 price

**Only available to countries that tariffied their NTBs** — mainly developed countries and a few developing countries that had NTBs pre-1995.

::: {.methodbox}
India cannot invoke the SSG on most products because India used high bound tariffs, not tariffication of all NTBs. India has SSG access for only ~6 products.
:::
:::
::: {.column width="45%"}
::: {.keybox}
**Special Safeguard Mechanism (SSM)**

**What developing countries want (Doha)**

An SSM would allow all developing countries to temporarily raise tariffs beyond bound levels when imports surge or prices fall — without the SSG's historical eligibility restriction.

Key disagreement (2008 collapse): At what threshold can the SSM be triggered? USA wanted 140% of import volume; India demanded 110%.

**SSM not agreed** — one of the main reasons the Doha Round collapsed.
:::
:::
:::

---

# Pillar 2 — Domestic Support

---

## Aggregate Measurement of Support (AMS) {.smaller}

::: {.columns}
::: {.column width="55%"}
**What is AMS?**

A measure of the total trade-distorting domestic support given to agriculture by a government.

**Formula for product-specific AMS:**

$$\text{AMS} = (\text{Administered Price} - \text{FERP}) \times \text{Eligible Production}$$

where:
- **Administered Price** = MSP or procurement price
- **FERP** = Fixed External Reference Price (average world price, 1986–88)
- **Eligible Production** = quantity actually purchased at administered price

**Non-product-specific AMS:** Sum of general input subsidies (fertiliser, power, irrigation, credit) across all crops
:::
::: {.column width="45%"}
::: {.highlightbox}
**Why the FERP is controversial**

The FERP is fixed at **1986–88 world prices** — when US and EU subsidies were at their peak, artificially depressing world prices.

Rice FERP: **USD 116/tonne** (1986–88 world price)

Current world price (2024): ~**USD 450–500/tonne**

So the "ceiling" against which India's MSP is measured is a 40-year-old, distorted baseline. India argues this systematically overestimates India's support and underestimates rich country support.
:::

::: {.methodbox}
Developing country de minimis: **10%** of production value (5% for developed countries)
:::
:::
:::

---

## The Three Boxes: Traffic Light Analogy {.smaller}

::: {.columns}
::: {.column width="33%"}
::: {.keybox}
**🟢 GREEN BOX**

*Allowed — no limit*

Criteria:
- Government-funded
- No price support
- Minimal trade distortion

Includes:
- Agricultural research
- Extension services
- Pest/disease control
- Infrastructure
- Food security stockholding (with conditions)
- Decoupled income support
- Environmental programs
:::
:::
::: {.column width="33%"}
::: {.methodbox}
**🔵 BLUE BOX**

*Permitted conditionally*

Payments under **production-limiting programs**

Used primarily by EU (old area payments) and USA (counter-cyclical payments tied to historical acreage, not current production)

Less relevant for India — India doesn't formally claim Blue Box, though some argue certain schemes could qualify
:::
:::
::: {.column width="33%"}
::: {.highlightbox}
**🟡 AMBER BOX**

*Restricted — must not exceed AMS commitment*

All trade-distorting support: MSP procurement, input price subsidies tied to specific products

Subject to de minimis threshold:
- Product-specific: 10% of that product's value
- Non-product-specific: 10% of total agri output value

India's AMS commitment: India has no scheduled AMS reduction commitment — it must only stay within de minimis (10%)
:::
:::
:::

---

## India's Agricultural Support by WTO Box {.center}

```{r}
#| label: fig-india-box-subsidies
#| fig-cap: "India's Agricultural Support by WTO Box (FY2020)\nSource: India's WTO Domestic Support Notifications; Ministry of Agriculture, GoI."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)

df_box <- data.frame(
  box   = c("Green Box\n(Exempt)",
            "Amber Box\n(Capped)",
            "Blue Box\n(N/A for India)"),
  value = c(32.5, 5.8, 0),
  color = c("#2E7D32", "#B9975B", "#90CAF9")
)

df_box$box <- factor(df_box$box, levels = df_box$box)

ggplot(df_box, aes(x = box, y = value, fill = box)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(
    aes(label = ifelse(value > 0, paste0("$", value, "B"), "N/A")),
    vjust = -0.5, size = 4, fontface = "bold"
  ) +
  annotate("segment", x = 1.65, xend = 2.35, y = 9.0, yend = 9.0,
           color = "#C84B31", linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = 2, y = 10.5,
           label = "De Minimis limit ~$8.6B\n(10% of agri output value)",
           color = "#C84B31", size = 3.2) +
  scale_fill_manual(values = c(
    "Green Box\n(Exempt)"          = "#2E7D32",
    "Amber Box\n(Capped)"          = "#B9975B",
    "Blue Box\n(N/A for India)"    = "#90CAF9"
  )) +
  scale_y_continuous(limits = c(0, 38), expand = expansion(mult = c(0, 0.06))) +
  labs(
    title    = "India's Agricultural Support by WTO Box (FY2020)",
    subtitle = "Green Box: Research, Extension, PSH Storage  |  Amber Box: Input subsidies (MSP)",
    x = NULL, y = "USD Billion"
  ) +
  theme_minimal(base_size = 11)
```

---

## Green Box: What India Can and Cannot Claim {.smaller}

::: {.columns}
::: {.column width="55%"}
**India's Legitimate Green Box Programs**

::: {.keybox}
Programs India can legitimately classify as Green Box:

- **ICAR research programs** — rice varieties, disease resistance, HYV seeds
- **Krishi Vigyan Kendras (KVKs)** — 731 farm science centres across India
- **Agricultural extension services** — ATMA, ARYA programs
- **National Food Security Act** — *food security stockholding* (subject to conditions)
- **Pradhan Mantri Krishi Sinchayee Yojana** — irrigation infrastructure
- **PM-KISAN** — ₹6,000/year decoupled income transfer to 110 million farmers
- **Crop insurance (PMFBY)** — part may qualify as decoupled
:::
:::
::: {.column width="45%"}
**What India CANNOT Claim as Green Box**

::: {.methodbox}
**MSP procurement system** is NOT Green Box because:

1. It involves price support (FCI buys at above-market MSP)
2. It is linked to the quantity produced and sold to government
3. It distorts production decisions — farmers produce more rice/wheat for FCI procurement than they would at market prices

→ MSP procurement falls in the **Amber Box**
→ Subject to de minimis limit
→ Protected only by Peace Clause
:::

::: {.highlightbox}
PM-KISAN, if truly decoupled (paid regardless of what crop is grown or how much), is a legitimate Green Box payment. India has been expanding this.
:::
:::
:::

---

## India's Amber Box Challenge: The Numbers {.smaller}

::: {.columns}
::: {.column width="55%"}
**Calculating India's Rice AMS (FY2023–24)**

| Item | Value |
|------|-------|
| MSP for paddy (common) | ₹2,183/quintal ≈ **USD 26.3/quintal** |
| FERP (1986–88 world price) | USD 116/tonne = **USD 11.6/quintal** |
| Support per quintal | USD 26.3 − USD 11.6 = **USD 14.7/quintal** |
| Total procurement by FCI | ~**42 million tonnes** |
| **Total rice AMS** | USD 14.7 × 420,000 quintals per tonne... |

Calculation: USD 14.7/quintal × 420 million quintals = **~USD 6.2 billion**

::: {.methodbox}
De minimis limit (10% of rice production value):
Rice output ~120 MT × ~USD 250/tonne = USD 30B × 10% = **~USD 3 billion threshold**

India's rice AMS (**USD 6.2B**) **exceeds** the de minimis limit (**USD 3B**)
:::
:::
::: {.column width="45%"}
::: {.keybox}
**Why India Is Technically in Violation**

India's rice AMS exceeds its 10% de minimis limit. Without the Peace Clause:

→ Any WTO member could bring a dispute case
→ Panel could find India in violation
→ India would need to either cut MSP or reduce procurement

This is why the **Peace Clause is not optional for India** — it is existential for India's food security architecture.

**India's counter-argument:** Update the FERP to current world prices. AMS would then be **negative** (MSP < current world price for rice and wheat).
:::
:::
:::

---

## Peace Clause: Technical Details {.smaller}

::: {.columns}
::: {.column width="55%"}
**Bali Ministerial Decision on Public Stockholding**
*(MC9, December 2013 — India's biggest recent WTO win)*

**Conditions for Peace Clause protection:**

1. Program **existed before December 31, 2013** (India's FCI procurement qualifies)
2. **Annual notification** to WTO Committee on Agriculture with detailed data on:
   - Stocks procured, average prices, distribution
   - Assessment of trade distortion impact
3. Stocks **not exported below acquisition cost** — cannot be used to dump food on world markets
4. Program **targets the poor** — National Food Security Act 2013 explicitly does this (75% rural, 50% urban population)

::: {.keybox}
India has notified its public stockholding program **every year** since 2013
:::
:::
::: {.column width="45%"}
**The Permanent Solution Problem**

::: {.methodbox}
MC10 Nairobi (2015): "Ministerial Decision affirms that the Peace Clause will remain in force until a permanent solution is found."

Committed to permanent solution by **MC11 (2017)** — missed.

Each subsequent MC: reaffirmed commitment — no solution.

India's demands for permanent solution:
1. Update FERP to current world prices OR base year of 2014–16
2. Include food inflation adjustment
3. No conditions on which crops qualify
:::

::: {.highlightbox}
The FERP update would essentially solve India's problem — at current world prices, India's MSP is often below market, making AMS negative. USA/EU resist this change as it would expose their own support programs.
:::
:::
:::

---

## Non-Product-Specific Support: India's Fertiliser Subsidy {.smaller}

::: {.columns}
::: {.column width="55%"}
**The Fertiliser Subsidy Problem**

India's fertiliser subsidy (FY2023–24): **₹2.25 lakh crore (~USD 27 billion)**

This is a **non-product-specific** subsidy — available to all crops, not tied to any specific commodity.

Goes into the **non-product-specific AMS pool:**

| Subsidy | Approximate Value (FY24) |
|---------|-------------------------|
| Fertiliser | ₹2,25,000 crore |
| Power subsidy to agriculture | ₹80,000–90,000 crore |
| Irrigation water subsidy | ₹30,000 crore |
| Agri credit subvention | ₹20,000 crore |
| **Total NPS AMS** | **~₹3.5–4.5 lakh crore** |
:::
::: {.column width="45%"}
::: {.keybox}
**India's WTO Argument**

Non-product-specific AMS de minimis: **10% of total agricultural output value**

India's total agri GVA (2023–24): ~₹24 lakh crore (~USD 288B)

De minimis threshold: **USD 28.8 billion**

India's total NPS support: ~USD 42–55 billion

→ India likely exceeds NPS de minimis too

**India's defence:** 
- Most subsidies benefit small/marginal farmers (eligible for S&DT input subsidies)
- Investment subsidies for agriculture exempt per AoA Annex 2
- S&DT permits input subsidies for low-income resource-poor farmers
:::

::: {.methodbox}
India has never submitted a formal AMS notification to WTO — technically still in "due restraint" (pre-Bali practice). This opacity is itself a WTO violation risk.
:::
:::
:::

---

# Pillar 3 — Export Competition

---

## Export Subsidies: History and Elimination {.smaller}

::: {.columns}
::: {.column width="55%"}
**AoA Article 9 — Export Subsidy Disciplines**

Original Uruguay Round commitments (developed countries):
- Reduce export subsidy **value** by **36%** (from 1986–90 base)
- Reduce subsidised export **volumes** by **21%**
- Over 6-year implementation period

Developing countries: 24% value cut, 14% volume cut over 10 years

**What counted as export subsidies:** Direct payments, government stock selling below cost, export credit at below-market rates, food aid as export dumping, payments to marketing boards for exports

::: {.keybox}
**Nairobi MC10 (December 2015):** Developed countries committed to **eliminate** export subsidies immediately for most products (by end-2018 for almost all). Developing countries: 2023 for most, 2018 for some.
:::
:::
::: {.column width="45%"}
**Who Were the Big Offenders?**

::: {.methodbox}
**EU Common Agricultural Policy:**
Peak export subsidy bill (1992): **€10 billion/year**
Used to export surplus wheat, butter, cheese, beef, and sugar at 40–60% below domestic prices.

Hurt: Argentina (wheat), Australia (dairy), Brazil (sugar), India (rice, cotton)

**USA:** Export Enhancement Program (EEP) subsidised wheat exports to Eastern Europe and developing countries

Both programs dramatically wound down after AoA and Nairobi — a genuine WTO success story
:::

::: {.highlightbox}
The elimination of EU and US agricultural export subsidies (largely achieved by 2020) is one of the AoA's real achievements — reduced price depression in world markets
:::
:::
:::

---

## India's Export Support: WTO Legal? {.smaller}

::: {.columns}
::: {.column width="55%"}
**India Does Not Have Scheduled Export Subsidy Commitments**

India never committed to reduce or eliminate export subsidies in its WTO schedule — because India historically didn't use them. This gives flexibility.

**But India has used export support programs:**

::: {.methodbox}
**MEIS (Merchandise Exports from India Scheme):**
- Duty credit scrips for exporters
- USA challenged it in DS541 (2019)
- WTO Panel ruled MEIS **violated** AoA export subsidy disciplines
- India discontinued MEIS
- Replaced by **RoDTEP** (Remission of Duties and Taxes on Export Products)
:::
:::
::: {.column width="45%"}
**RoDTEP: India's Current Export Support**

::: {.keybox}
India's argument: RoDTEP is a **legitimate tax refund** — it refunds embedded taxes that cannot be claimed back otherwise (state levies, mandi taxes, electricity duty). This is NOT a subsidy but a correction of tax cascading.

WTO principle: Exported goods should not bear domestic taxes. Duty drawback is permitted.

If RoDTEP refunds only actual embedded taxes (no more, no less), it is WTO-consistent.

Risk: If rates are set above actual embedded taxes, the excess becomes a subsidy.
:::

::: {.highlightbox}
**APEDA, MAI, TMA** schemes (marketing assistance) — India argues these are Green Box general services (market promotion, not direct export subsidies)
:::
:::
:::

---

## Export Restrictions: AoA Article 12 {.smaller}

::: {.columns}
::: {.column width="55%"}
**When India Bans Agricultural Exports**

AoA Article 12: When imposing food export restrictions or prohibitions for food security:
1. **Notify** the WTO in advance
2. **Consider the effects** on importing countries' food security
3. **Consult** if any member requests

**India's August 2023 Export Restrictions (major case):**
- **Non-basmati white rice:** Complete export ban (July 20, 2023)
- **Parboiled rice:** 20% export duty
- **Broken rice:** Earlier ban (September 2022) continued
- Reason: Domestic food inflation, El Niño fears, low buffer stocks

India notified WTO but didn't consult affected countries in advance.
:::
::: {.column width="45%"}
::: {.keybox}
**Who Protested India's Rice Export Restrictions?**

- **Bangladesh:** ~35% of rice imports from India; scrambled for alternatives
- **Philippines:** Faced sudden shortfall in rice supplies
- **West African countries** (Senegal, Cameroon, Ivory Coast): dependent on Indian broken rice
- **Indonesia, Malaysia:** Sought alternative supplies

Several countries formally requested consultations under AoA Art. 12 — **first major use of Art. 12 consultation requests against India**.

India's response: domestic food security is paramount; exported food is subject to sovereign decision.
:::

::: {.methodbox}
The episode highlighted the global dependency on Indian rice exports and India's responsibility as the world's largest rice exporter (~40% of global rice trade in 2022)
:::
:::
:::

---

## Special and Differential Treatment for Developing Countries {.smaller}

::: {.columns}
::: {.column width="50%"}
**S&DT Provisions in AoA**

::: {.keybox}
**Market Access:**
- Smaller tariff cuts: 24% average (vs. 36%)
- Minimum per product: 10% (vs. 15%)
- Longer implementation: 10 years (vs. 6)

**Domestic Support:**
- De minimis: 10% (vs. 5%) of product value
- Input subsidies for low-income/resource-poor farmers — **not counted in AMS** (AoA Annex 2, Para 4)
- Investment subsidies for agriculture — **not counted**

**Export Competition:**
- Developing countries retain right to use export subsidies for transport, marketing to reduce marketing costs
- Longer phase-out periods
:::
:::
::: {.column width="50%"}
**India's Use of S&DT**

::: {.methodbox}
India fully utilises every S&DT provision:

1. **High bound tariffs** — used the 10-year period to bind at high rates
2. **10% de minimis** — gives India more Amber Box space
3. **Input subsidy exemption** — fertiliser subsidies targeted to small/marginal farmers (85% of India's 140 million farm households are small/marginal)
4. **Investment subsidy exemption** — RKVY, PMKSY, agricultural infrastructure investments
:::

::: {.highlightbox}
India's 2017 WTO notification: India claimed the input subsidy exemption for fertiliser subsidies benefiting **low-income, resource-poor farmers** — this is the strongest legal shield available.
:::
:::
:::

---

## The Unfinished Doha Agriculture Agenda {.smaller}

::: {.columns}
::: {.column width="55%"}
**What Doha Promised on Agriculture**

::: {.methodbox}
The Doha Development Agenda (2001) mandated "substantial improvements" in:

1. **Market access** — "substantial reductions in tariffs" including elimination of all tariff peaks and escalation
2. **Domestic support** — "substantial reductions in trade-distorting support" especially Amber Box
3. **Export competition** — "elimination with a view to phasing out all forms of export subsidies" ✅ (Done at Nairobi 2015)
:::

**The Sticking Points:**

- USA: refuses to cut overall trade-distorting support (OTDS) significantly; Farm Bill politics
- EU: defensive on sugar, dairy, poultry
- India: demands strong SSM; resists deep tariff cuts on sensitive products
- China: joined WTO 2001; now major power with complex interests
:::
::: {.column width="45%"}
::: {.keybox}
**Where We Are in 2026**

The Doha Round as a "single undertaking" is dead.

**What has happened:**
✅ Export subsidy elimination (Nairobi 2015)
✅ Trade Facilitation Agreement (Bali 2013 / 2017)
✅ Peace Clause (Bali 2013)
✅ Fisheries subsidies (MC12 2022)

**What has NOT happened:**
❌ Substantial Amber Box cuts in US/EU
❌ Special Safeguard Mechanism (SSM)
❌ Permanent solution for public stockholding
❌ Market access improvements (tariff cuts on both sides)

After **30 years**, agriculture remains the most politically contested sector in trade negotiations
:::
:::
:::

---

## AoA and Indian Farmer Welfare: An Assessment {.smaller}

::: {.columns}
::: {.column width="55%"}
**Arguments that AoA helps Indian farmers**

- Disciplines US/EU subsidies that depressed world prices for 40 years
- Eliminated EU/US export subsidies — raised world prices for wheat, sugar, dairy
- Forced transparency in domestic support programs
- Created a framework where India can challenge unfair competition

**Arguments that AoA constrains India**

- Constrains India's ability to provide MSP support beyond de minimis
- FERP methodology unfair — systematically overestimates India's "distortion"
- Tariff reduction pressure (even if gradual) limits India's protection for sensitive crops
- Disciplines India's export support while US subsidies continue under "Green Box" labels
:::
::: {.column width="45%"}
::: {.keybox}
**Overall Assessment**

India has navigated the AoA remarkably well:

1. **High bound tariffs** = protection without WTO violation
2. **Peace Clause** = MSP procurement protected
3. **S&DT** = input subsidies, larger de minimis
4. **PM-KISAN** = Green Box-compliant income support
5. **Active coalitions** (G-33, G-20) = blocking unfavourable outcomes

India's domestic support policy is primarily driven by **food security and farmer welfare** — WTO constraints are real but India has worked around them creatively.

The key challenge: **permanent solution** for public stockholding — if this fails, India's MSP system remains legally vulnerable beyond the Peace Clause's diplomatic protection
:::
:::
:::

---

## India Agricultural Trade Balance {.center}

```{r}
#| label: fig-india-agri-trade
#| fig-cap: "India Agricultural Trade Balance (FY2019-FY2024, USD Billion)\nSource: APEDA / DGCI&S."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(tidyr)

df_trade <- data.frame(
  year    = c("FY2019","FY2020","FY2021","FY2022","FY2023","FY2024"),
  Exports = c(32.8, 35.2, 41.3, 50.2, 53.2, 43.7),
  Imports = c(19.1, 20.3, 19.6, 26.8, 29.3, 27.4)
)

df_trade$balance <- df_trade$Exports - df_trade$Imports
df_trade$year    <- factor(df_trade$year, levels = df_trade$year)

df_long_t <- pivot_longer(df_trade, cols = c(Exports, Imports),
                          names_to = "Type", values_to = "Value")

scale_s <- max(df_long_t$Value) / max(df_trade$balance)

ggplot() +
  geom_col(data = df_long_t,
           aes(x = year, y = Value, fill = Type),
           position = "dodge", width = 0.65) +
  geom_line(data = df_trade,
            aes(x = as.numeric(year), y = balance * scale_s, group = 1),
            color = "#C84B31", linewidth = 1.3) +
  geom_point(data = df_trade,
             aes(x = as.numeric(year), y = balance * scale_s),
             color = "#C84B31", size = 2.5) +
  scale_fill_manual(values = c("Exports" = "#012169", "Imports" = "#B9975B")) +
  scale_y_continuous(
    name     = "USD Billion",
    sec.axis = sec_axis(~ . / scale_s, name = "Trade Balance (USD Billion)")
  ) +
  labs(
    title   = "India Agricultural Trade Balance (FY2019-FY2024, USD Billion)",
    x = NULL, fill = NULL,
    caption = "Red line = Trade Balance (right axis)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## Key Takeaways: Agreement on Agriculture {.smaller}

::: {.keybox}
**Seven Things to Remember from Lecture 14**

1. **AoA has three pillars:** Market Access (tariffication + TRQs + reductions), Domestic Support (Green/Blue/Amber Boxes + de minimis), Export Competition (subsidy disciplines)

2. **India's bound tariffs average 113.5%** — the "water" in the tariff is India's primary agricultural protection instrument

3. **Three Boxes:** Green = unlimited (research, extension, decoupled payments); Blue = conditional (production-limiting); Amber = limited by de minimis (MSP procurement falls here)

4. **India's MSP procurement for rice likely exceeds 10% de minimis** — protected only by the Peace Clause (Bali 2013)

5. **FERP is the core injustice:** Using 1986–88 world prices (depressed by US/EU subsidies) overstates India's trade distortion. Updating FERP to current prices would solve India's problem

6. **Nairobi 2015 eliminated export subsidies** — a genuine WTO success; India's RoDTEP is a legitimate tax refund scheme (if correctly calibrated)

7. **Doha Round dead** — but AoA framework remains; India must push for permanent public stockholding solution at every MC
:::

---

## Next Lecture: SPS Measures and Codex Alimentarius {.smaller}

::: {.columns}
::: {.column width="60%"}
**Lecture 15 (August 4, 2026):**

**SPS Measures and Codex Alimentarius**

Why Indian spices were pulled from Singapore supermarkets.Why India's shrimp face detention in the USA. Why EU pesticide standards block Indian rice exports.

We will cover:

- **SPS Agreement** — scientific basis, Codex/OIE/IPPC
- **Codex Alimentarius** — the food code that governs market access
- **India's export rejection challenges:** spices, shrimp, basmati rice
- **Case studies:** MDH/Everest spice scandal, Indian aquaculture
- **India's strategy** for SPS compliance
:::
::: {.column width="40%"}
::: {.highlightbox}
**Think About This**

In April 2023, **MDH and Everest** — two of India's most iconic spice brands — had their products banned in Singapore and Hong Kong.

The reason: ethylene oxide (a fumigant) detected above permitted limits.

*What standard did they violate? Who sets these standards? Is this genuine food safety — or disguised protectionism?*

We will answer all of these in Lecture 15.
:::

**Reading:**
- WTO SPS Agreement (full text)
- FSSAI Annual Report 2023-24
- Spices Board of India Export Statistics
- MPEDA: Aquaculture Export Performance Report
:::
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 12)
- *International Economics* — Appleyard & Field (Ch. 12)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
