---
title: "Lecture 9: Balance of Trade & Balance of Payments"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "June 20, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(patchwork)
options(scipen = 999)
theme_set(theme_minimal(base_size = 11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## Introduction: Why the BoP Matters

**Key questions for today:**

- How does a country "pay" for its imports?
- What happens when a country consistently imports more than it exports?
- How do capital flows balance the external accounts?
- When does a BoP *crisis* occur?

::: {.highlightbox}
**BoP accounting identity:**

$$\underbrace{\text{CA}}_{\text{trade + income}} + \underbrace{\text{KA}}_{\text{transfers}} + \underbrace{\text{FA}}_{\text{capital flows}} + \underbrace{\Delta R}_{\text{reserve change}} = 0$$

A current account deficit *must* be financed by capital/financial inflows or by running down foreign exchange reserves. If neither is available → **BoP crisis** (India 1991).
:::

:::{.notes}
**India's external sector at a glance (FY2024):**

- Merchandise exports: **$437 billion** (goods)
- Merchandise imports: **$677 billion** (goods)
- Trade deficit: **-$240 billion** (goods only)
- Services surplus: **+$163B**
- Remittances: **+$120B** (world's largest recipient)
- Current account: **-$23B (-0.7% of GDP)**
:::

---

## Balance of Trade (BoT): the idea

$$\text{BoT} = X_{goods} - M_{goods}$$

- BoT > 0: trade surplus
- BoT < 0: trade deficit
- India typically has a **goods deficit** (oil + gold + electronics)

---

## India FY2024: why the goods deficit persists

- Goods exports ≈ **$437B**
- Goods imports ≈ **$677B**
- Goods trade balance ≈ **-$240B**

::: {.keybox}
Structural drivers: crude oil dependence + gold demand + electronics import intensity.
:::

---

## Current Account identity (absorption approach)

$$Y = C + I + G + (X - M)$$

Rearrange:

$$\boxed{\text{CA} \approx (X-M) = Y - A}\quad\text{where }A=C+I+G$$

- If **$Y>A$**: produce more than spend → net lender
- If **$Y<A$**: spend more than produce → net borrower

---

## Current Account identity (saving–investment form)

$$\boxed{\text{CA} = (S - I) + (T - G)}$$

- CA deficit can reflect **high investment** (good) or **fiscal deficit** (risk)
- Devaluation alone won’t fix CA if absorption stays high

---

## India FY2024: why the CA deficit was small

- Goods trade deficit was large (≈ **-$240B**)
- Services trade surplus was large (≈ **+$163B**)
- Remittances were huge (≈ **+$120B**)
- Net result: CA ≈ **-$23B (-0.7% of GDP)**

::: {.keybox}
In India, services + remittances are the “cushion” behind the external sector.
:::

---

## BoP accounts: Current vs Capital

- **Current Account (CA):** goods, services, income, transfers
- **Capital Account (KA):** capital transfers (usually small)
- **Financial Account (FA):** FDI, FPI, loans, deposits, reserves

---

## Double-entry bookkeeping (2 examples)

- Import machines worth $10M: CA **-10** and FA **+10**
- Export software worth $5M: CA **+5** and FA **-5** (forex asset)

::: {.keybox}
Every transaction has equal-and-opposite entries → accounts always add up.
:::

## Financial Account and BoP Identity

## The Balance of Payments (BoP): Accounts + Identity

- **Current Account (CA):** goods, services, income, transfers
- **Capital Account (KA):** capital transfers (typically small)
- **Financial Account (FA):** FDI, FPI, loans, and reserve changes
- **Accounting identity:** $\text{CA} + \text{KA} + \text{FA} + \Delta R = 0$
- If a CA deficit cannot be financed → **BoP crisis** (India 1991)

---

## India: BoP snapshot (FY2024)

| Component | USD B |
|---|---|
| Current Account | -23 |
| Financial Account | +20 |
| Δ Reserves | +3 |
| **Total** | **≈ 0** |

:::{.notes}
The CA deficit is financed primarily through FDI and portfolio inflows, making FA stability critical.
:::

---

## India's BoP Components: FY2020–FY2024

```{r}
#| label: fig-bop-india
#| fig-cap: "India's Balance of Payments Components, FY2020–FY2024 (USD billion)\nSource: RBI, Balance of Payments Statistics."
#| fig-width: 9
#| fig-height: 5

bop_data <- data.frame(
  year           = rep(c("FY20","FY21","FY22","FY23","FY24"), each = 6),
  component      = rep(c("Merchandise\nTrade Balance", "Services\nSurplus",
                         "Remittances", "Investment\nIncome (net)",
                         "FDI (net)", "FPI (net)"), times = 5),
  value          = c(
    # FY20
    -157,  85,  83, -26,  43, -14,
    # FY21
     -98,  89,  83, -29,  44,  36,
    # FY22
    -190, 107,  89, -33,  39, -16,
    # FY23
    -264, 143, 112, -39,  28, -16,
    # FY24
    -238, 163, 120, -68,  22,  41
  )
)

bop_data <- bop_data |>
  mutate(
    account = ifelse(component %in% c("Merchandise\nTrade Balance",
                                       "Services\nSurplus",
                                       "Remittances",
                                       "Investment\nIncome (net)"),
                     "Current Account", "Financial Account"),
    component = factor(component,
                       levels = c("Merchandise\nTrade Balance",
                                  "Investment\nIncome (net)",
                                  "FPI (net)",
                                  "FDI (net)",
                                  "Services\nSurplus",
                                  "Remittances"))
  )

ggplot(bop_data, aes(x = year, y = value, fill = component)) +
  geom_col(position = "dodge", width = 0.85) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "grey30") +
  scale_fill_manual(values = c(
    "Merchandise\nTrade Balance" = "#8B0000",
    "Investment\nIncome (net)"  = "#cc4444",
    "FPI (net)"                  = "#B9975B",
    "FDI (net)"                  = "#7B9F35",
    "Services\nSurplus"          = "#012169",
    "Remittances"                = "#5B8DB8"
  )) +
  labs(
    title    = "India's Balance of Payments Components, FY2020–FY2024",
    subtitle = "Merchandise deficit offset by services surplus, remittances, and (usually) financial inflows",
    x = NULL, y = "USD Billion", fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(size = 8),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2))
```

---

## India's Current Account: FY2024 Decomposition

```{r}
#| label: fig-ca-waterfall
#| fig-cap: "India's Current Account, FY2024: Component Contributions (USD Billion)\nSource: RBI, Balance of Payments Statistics."
#| fig-width: 7.5
#| fig-height: 4.5

# CA = merch balance + services + remittances + investment income
# = -238 + 163 + 120 - 68 = -23 (approx)
ca_items <- data.frame(
  item      = factor(
    c("Merch.\nExports", "Merch.\nImports", "Services\nSurplus",
      "Remittances", "Inv. Income\n(net)", "CA Balance"),
    levels = c("Merch.\nExports", "Merch.\nImports", "Services\nSurplus",
               "Remittances", "Inv. Income\n(net)", "CA Balance")
  ),
  value     = c( 437, -677,  163,  120, -68, -23),
  fill_col  = c("#012169", "#8B0000", "#012169", "#5B8DB8", "#8B0000", "grey30"),
  label_pos = c( 437, -677,  163,  120, -68, -23)
)

ggplot(ca_items, aes(x = item, y = value, fill = fill_col)) +
  geom_col(width = 0.65) +
  scale_fill_identity() +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  geom_text(
    aes(
      label = paste0(ifelse(value >= 0, "+", ""), value, "B"),
      vjust = ifelse(value >= 0, -0.4, 1.3)
    ),
    size = 3.5, fontface = "bold"
  ) +
  scale_y_continuous(
    limits = c(-720, 520),
    labels = label_dollar(prefix = "$", suffix = "B")
  ) +
  labs(
    title    = "India's Current Account Components, FY2024 (USD Billion)",
    subtitle = "Gross exports shown separately from imports for clarity\nCA deficit = -$23B (-0.7% of GDP) — manageable",
    x = NULL, y = "USD Billion"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

---

## Financial Account (FA): India FY2024 (headline)

- FA inflows were positive (FDI + portfolio + loans + deposits)
- This helped finance the CA deficit and support reserves

$$\text{BoP identity: } CA + KA + FA + \Delta R = 0$$

---

## FDI vs FPI (why it matters)

- **FDI:** long-term ownership, more stable
- **FPI:** liquid “hot money”, more volatile
- BoP vulnerability often comes from **sudden stops** in FPI/short-term flows

---

## Forex reserves: what they are

- Foreign currency assets (USD/EUR/GBP/JPY, etc.)
- Gold
- SDRs + IMF reserve tranche
- Used for liquidity in crises and to smooth extreme volatility

::: {.keybox}
A common adequacy benchmark is “months of import cover”.
:::

---

## RBI intervention under a managed float

- **Buy USD / sell INR** to resist sharp appreciation
- **Sell USD / buy INR** to resist sharp depreciation
- Goal is usually to **smooth volatility**, not to fix a peg
- Intervention affects reserves and domestic liquidity

---

## BoP adjustment mechanisms (3 channels)

- **Price (exchange rate):** depreciation switches demand toward domestic goods
- **Income/absorption:** lower spending reduces imports
- **Monetary/financial:** reserve loss tightens liquidity; rates rise; capital flows respond

---

## Typical policy instruments

- **Devaluation / depreciation** (expenditure-switching)
- **Fiscal/monetary tightening** (expenditure-reducing)
- **Interest rate policy** to stabilise capital flows
- **IMF financing** in reserve crises

::: {.keybox}
India 1991 used a mix: devaluation + tightening + reforms.
:::

---

## Marshall–Lerner condition

$$\boxed{|e_X| + |e_M| > 1}$$

- Depreciation makes exports cheaper and imports costlier
- Trade balance improves only if **quantities respond enough**
- Short run: contracts + habits → weak quantity response

---

## Elasticities: short run vs long run

- **Long run:** $|e_X|$ and $|e_M|$ tend to be higher → M–L more likely to hold
- **Short run:** elasticities are low → trade balance can worsen first

::: {.keybox}
This is the logic behind the J-curve.
:::

---

## The J-Curve Effect

```{r}
#| label: fig-jcurve
#| fig-cap: "J-Curve: Trade Balance Initially Worsens After Currency Depreciation\nSource: Author's illustration."
#| fig-width: 8
#| fig-height: 4.5

set.seed(42)
t <- seq(-5, 14, by = 0.2)

# J-curve shape: initial deterioration then improvement
# TB(t<0) = 0 (baseline)
# After depreciation at t=0:
#   short run (0-3 months): TB falls (pre-set contracts, inelastic demand)
#   medium (3-12 months):   TB recovers as quantities adjust
#   long run (12+ months):  TB improves above baseline if M-L holds

tb_j <- ifelse(
  t < 0, 0,
  ifelse(
    t < 3,
    -4 * (t/3),                                    # deterioration phase
    ifelse(
      t < 8,
      -4 + 5 * ((t - 3)/5),                        # recovery phase
      1 + 0.3 * (t - 8)                             # long-run improvement
    )
  )
)

# Add some noise for realism
tb_j <- tb_j + c(rep(0, sum(t < 0)), rnorm(sum(t >= 0), 0, 0.15))

df <- data.frame(time = t, tb = tb_j)

ggplot(df, aes(x = time, y = tb)) +
  geom_line(color = "#012169", linewidth = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1.2) +
  # Shaded deterioration region
  geom_ribbon(
    data = df |> filter(time >= 0 & time <= 4),
    aes(ymin = pmin(tb, 0), ymax = 0),
    fill = "#cc0000", alpha = 0.2
  ) +
  # Shaded improvement region
  geom_ribbon(
    data = df |> filter(time > 4),
    aes(ymin = 0, ymax = pmax(tb, 0)),
    fill = "#012169", alpha = 0.15
  ) +
  # Annotations
  annotate("text", x = 0.3, y = 3.8,
           label = "Depreciation\noccurs (t=0)", color = "red", size = 3.5, hjust = 0) +
  annotate("text", x = -3.5, y = 0.4,
           label = "Pre-depreciation\nbaseline", size = 3, color = "grey40") +
  annotate("text", x = 1.5, y = -2.8,
           label = "Short-run\ndeterioration\n(contracts pre-set,\nM-L not satisfied)",
           size = 3, color = "#cc0000", hjust = 0.5) +
  annotate("text", x = 11, y = 2.2,
           label = "Long-run\nimprovement\n(M-L holds,\nquantities adjust)",
           size = 3, color = "#012169", hjust = 0.5) +
  annotate("segment",
           x = 1.5, y = -2.3, xend = 1.5, yend = -1,
           arrow = arrow(length = unit(0.25, "cm")), color = "#cc0000") +
  annotate("segment",
           x = 11, y = 1.8, xend = 11, yend = 0.6,
           arrow = arrow(length = unit(0.25, "cm")), color = "#012169") +
  # Phase labels
  annotate("text", x = 1.5, y = -4.3,
           label = "Phase I:\n0-3 months", size = 2.8, color = "#cc0000") +
  annotate("text", x = 5.5, y = -4.3,
           label = "Phase II:\n3-12 months\n(recovery)", size = 2.8, color = "#B9975B") +
  annotate("text", x = 11,  y = -4.3,
           label = "Phase III:\n12+ months\n(improvement)", size = 2.8, color = "#012169") +
  scale_x_continuous(breaks = seq(-4, 14, 2),
                     labels = paste0(seq(-4, 14, 2), "m")) +
  scale_y_continuous(limits = c(-5.5, 5)) +
  labs(
    title    = "The J-Curve Effect",
    subtitle = "Trade balance worsens before improving after currency depreciation\nM-L condition (|eX|+|eM|>1) is satisfied only in the long run",
    x = "Time (months after depreciation)", y = "Trade Balance (change from baseline)"
  ) +
  theme_minimal(base_size = 11)
```

---

## J-curve: mechanisms (3 phases)

- **Phase I (0–3 months):** contracts fixed → values move before quantities
- Import bill can rise even if volumes don’t
- **Phase II (3–12 months):** firms adjust sourcing and sales
- **Phase III (12+ months):** quantities respond fully
- Long run: M–L more likely to hold → TB improves

---

## India: what we see in practice

- 1991: large devaluation; adjustment took time (rigidities)
- 2013: TB improved faster as trade was more integrated
- 2022: oil prices dominated → weaker TB response
- Agriculture adjusts faster than many manufactured goods (more spot pricing)

---

## Absorption approach: policy implication

$$\boxed{CA = Y - A}\quad (A=C+I+G)$$

- Improve CA by **raising output** ($Y$) and/or **reducing absorption** ($A$)
- Devaluation helps only if it changes quantities and spending patterns
- Inflation can offset gains if real absorption doesn’t fall

---

## India 1991: adjustment mix

- Fiscal tightening reduced absorption ($\Delta G<0$)
- Large devaluation switched demand toward domestic goods
- Reforms improved long-run output capacity
- CA improved over time, but growth fell temporarily
- Inflation spiked during adjustment

---

## India’s 1991 BoP crisis: what happened

- Oil prices spiked and external financing dried up (1990–91)
- Forex reserves fell to **$5.8B** (~3 weeks imports), then **$1.2B** (~2 weeks)
- India pledged **67 tonnes of gold** to raise emergency foreign exchange
- IMF support followed (Aug 1991)
- Adjustment package triggered liberalisation and macro stabilisation

---

## 1991 crisis: root causes and policy response

- **Fiscal deficit + short-term external borrowing** (fragile financing)
- **Oil shock** (Kuwait crisis) worsened import bill
- **Remittance/NRI deposit outflows** amplified the squeeze
- **Trade liberalisation + industrial delicensing** reduced distortions
- **FDI opening** (automatic route) improved long-run financing

:::{.notes}
Legacy: the crisis was the proximate cause of India’s 1991 reforms.
:::

---

## India FY2024 BoP Assessment {.smaller}

**Current Account:**

- CA deficit: $23.2B (-0.7% of GDP) — **comfortable**
- Merchandise trade deficit: $238B (oil, gold, electronics)
- Services surplus: $163B (IT/BPO dominance)
- Remittances: $120B — world's largest recipient (World Bank data)

::: {.methodbox}
**Risks and concerns:** (1) Oil price vulnerability: $10/barrel rise adds ~$15B to import bill; (2) FPI volatility: $41B inflow in FY24 could reverse quickly; (3) FDI decline from $55B (FY22) to $22B (FY24) — concerning for long-run financing. India's CA/GDP (-0.7%) << GDP growth (7%) — external position is **sustainable**.
:::

:::{.notes}
**Financial Account:**

- FDI net inflows: $22B (declined from $83B peak in FY22)
- FPI: +$41B (strong equity market inflows)
- ECB + banking capital: +$25B

**Reserve adequacy:**

- Forex reserves: $646B (March 2024)
- Import cover: ~11 months (**well above IMF 3-month threshold**)
- Short-term debt/reserves: 19% (**well below 100% danger threshold**)

$$\text{CA/GDP} < \text{long-run growth rate} \times \frac{\text{Net foreign debt}}{\text{GDP}}$$
:::

---

## BoP and agriculture: key linkages

- MSP and productivity can reduce **food imports** (import substitution)
- Export restrictions (rice/onion/sugar) affect **credibility** and export earnings
- Input policies affect **cost competitiveness**
- Edible oil dependence is a major **import bill** risk
- Migration/remittances matter for rural income and the CA

---

## India’s agri trade balance: surplus, with one big weakness

- Net agri balance is typically **positive** (order of ~$20B)
- Rice and spices are major surplus contributors
- **Edible oils** are the dominant structural deficit item (≈ -$20B)
- Pulses can be a smaller deficit item in some years

::: {.keybox}
Policy takeaway: edible oils are the macro vulnerability; SPS/processing are the export upside.
:::

---

## India’s services exports: the external-sector cushion

- Services exports create a large surplus (IT/ITES is central)
- They are often more stable than goods exports
- This surplus helps offset the goods trade deficit

---

## Why it matters for policy

- BoP resilience depends heavily on the IT/services engine
- Disruptions to services exports would widen the CA deficit quickly
- Services trade negotiations (Mode 4) are a key Indian interest

---

## Remittances: India’s largest CA credit (FY2024)

- Remittances ≈ **$120B** (World Bank)
- Major sources: **USA, UAE, Saudi Arabia, UK** (+ other Gulf)
- Often rise when INR is weaker (more rupees per dollar sent)

---

## Why remittances matter for agriculture

- Smooth consumption during crop shocks
- Finance farm investment (irrigation, equipment, land)
- Enable shift toward higher-value crops (risk capacity)
- Outmigration raises rural wages (labour market effects)
- BoP stabiliser: remittances are large and relatively stable

---

## India's Forex Reserves: RBI Intervention Chart

```{r}
#| label: fig-forex-reserves
#| fig-cap: "India's Forex Reserves and INR/USD Exchange Rate, FY2015–FY2024\nSource: RBI, Database on Indian Economy (DBIE)."
#| fig-width: 8.5
#| fig-height: 4.5

reserves_data <- data.frame(
  year      = 2015:2024,
  reserves  = c(341, 360, 370, 424, 412, 477, 590, 633, 578, 646),
  inr_usd   = c(62.1, 65.5, 67.1, 64.5, 69.9, 70.4, 74.2, 75.0, 82.7, 83.3)
)

# Dual-axis plot using secondary axis
coeff <- 10   # scale factor for INR (reserves ~100x INR so scale accordingly

p_res <- ggplot(reserves_data, aes(x = year)) +
  # Reserves bars
  geom_col(aes(y = reserves), fill = "#012169", alpha = 0.75, width = 0.6) +
  # INR/USD line (scaled)
  geom_line(aes(y = inr_usd * coeff), color = "#cc0000", linewidth = 1.5) +
  geom_point(aes(y = inr_usd * coeff), color = "#cc0000", size = 2.5) +
  # Mark 2022 taper tantrum equivalent
  annotate("text", x = 2023, y = 920,
           label = "Post-Ukraine\nrupee weakness\n(₹83/USD)", size = 3, color = "#cc0000") +
  # Mark 2020 reserve build
  annotate("text", x = 2021, y = 650,
           label = "$590B\n(COVID\nreserve build)", size = 3, color = "#012169") +
  scale_x_continuous(breaks = 2015:2024,
                     labels = paste0("FY", substr(2015:2024, 3, 4))) +
  scale_y_continuous(
    name     = "Forex Reserves (USD billion)",
    limits   = c(0, 1000),
    sec.axis = sec_axis(~ . / coeff,
                        name   = "INR per USD",
                        breaks = seq(60, 90, 5))
  ) +
  labs(
    title    = "India's Forex Reserves (bars) and INR/USD Rate (line), FY2015–FY2024",
    subtitle = "RBI builds reserves during capital inflow surges; deploys during stress periods",
    x = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.title.y.right = element_text(color = "#cc0000"),
    axis.text.y.right  = element_text(color = "#cc0000"),
    axis.title.y.left  = element_text(color = "#012169")
  )

p_res
```

---

## Key Takeaways: Lecture 9

::: {.keybox}
**1. National income identity:** $\text{CA} = Y - A = (S - I) + (T - G)$. A CA deficit means the country borrows from abroad — financed by FA inflows or reserve drawdown.

**2. BoP accounting:** $\text{CA} + \text{KA} + \text{FA} + \Delta R = 0$. The BoP always balances — surpluses and deficits are in *individual accounts*, not the overall BoP.

**3. India's external position (FY2024):** CA = -0.7% GDP (comfortable); remittances = $120B (world's largest); services surplus = $163B; forex reserves = $646B (11 months imports). External position is **strong**.

**4. Marshall-Lerner condition:** $|e_X| + |e_M| > 1$ must hold for depreciation to improve TB. In the short run, M-L is typically NOT satisfied → J-curve: TB worsens before improving.

**5. India 1991 crisis:** Reserves fell to $1.2B (2 weeks imports); India pledged 67 tonnes gold to the BoE. IMF bailout triggered liberalisation — delicensing, trade reform, FDI opening. Crisis → structural reform.
:::

:::{.notes}
**6. Agricultural BoP:** India runs a trade surplus in agriculture (~$20B), but edible oil imports (-$21B) are the single largest structural weakness. MSP policy affects both food security and export competitiveness.
:::

---

## Next Lecture Preview

::: {.methodbox}
**Lecture 10 — Exchange Rates and Agricultural Trade**
*June 27, 2026*

- Fixed vs flexible exchange rate systems
- Purchasing Power Parity (PPP): absolute and relative
- Real exchange rate: $q = eP^*/P$ and agricultural competitiveness
- How rupee depreciation affects Indian rice and wheat exports
- Dutch disease: can commodity booms harm agricultural exporters?
:::

:::{.notes}
- India's managed float: RBI intervention and the REER

**Preparation:** Review Salvatore Ch. 14–15 (Exchange Rates). Note that India operates a **managed float** — this means understanding both PPP theory AND RBI's intervention behaviour. We will use regression analysis of real exchange rate and export volume data.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

**Further reading**

- Salvatore, *International Economics* (relevant chapters)
- Appleyard & Field, *International Economics* (relevant chapters)

**Key data sources**

- DGCI\&S: merchandise trade
- RBI: balance of payments
- APEDA: agricultural export statistics
- WTO: tariff + trade databases
