---
title: "Lecture 16: Export Procedures and Documentation"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "August 8, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

# {.section}

::: {style="text-align:center; margin-top: 2em;"}
**Lecture 16**

[Export Procedures and Documentation]{.hi style="font-size:1.8em;"}

From Karnataka Farm to Leicester Kitchen
:::

---

## Opening: The Journey of a Kilo of Tur Dal

::: {.columns}
::: {.column width="58%"}
A kilo of **Kalaburagi tur dal** travels from a Karnataka farmer to a diaspora kitchen in **Leicester, UK**.

What stands between the farmer and the foreign consumer?

- Registrations with four government agencies
- Twelve documents crossing seven desks
- Two financial instruments spanning two continents
- One customs portal used by 1.4 million exporters

**Today's lecture answers: how does this actually work?**
:::
::: {.column width="42%"}
::: {.highlightbox}
**India's agri export snapshot (FY2024):**

- Total agri exports: **₹3.61 lakh crore (~$43.7B)**
- Agricultural exporters registered: **~85,000**
- Ports handling agri cargo: **13 major + 200 minor**
- Average documents per shipment: **8–12**

Every export begins with a registration and ends with a bank certificate.
:::
:::
:::

---

## Step 1 — Pre-Export Registrations: The Four Pillars

::: {.columns}
::: {.column width="55%"}
Before any shipment, an agricultural exporter must hold:

**1. Import Export Code (IEC)**
- Issued by DGFT (Director General of Foreign Trade)
- 10-digit PAN-based code — **mandatory** for all exports
- One-time registration; lifelong validity
- Apply online at dgft.gov.in — fee ₹500

**2. RCMC (Registration-cum-Membership Certificate)**
- Issued by the relevant commodity body:
  - [APEDA]{.hi} — processed food, fresh produce, meat, dairy, cereals
  - [MPEDA]{.hi} — marine products
  - [Spices Board]{.hi} — all spices
  - [Coffee/Tea/Rubber Board]{.hi} — respective commodities
:::
::: {.column width="45%"}
::: {.methodbox}
**3. GST Registration**

GSTIN is mandatory for tax input credit on export supplies. Exports are **zero-rated** under GST — exporter can claim refund of input taxes.

**4. AD Bank Account**

Authorised Dealer (AD) Category-I bank account for receiving foreign exchange. FEMA requires all export proceeds to be realised through an AD bank within **9 months** of shipment.
:::
:::
:::

::: {.keybox}
**Sequence matters:** IEC → GST → RCMC → AD Bank account. No RCMC without IEC. No export incentive claim without RCMC.
:::

---

## Export Contracts and Incoterms 2020

::: {.columns}
::: {.column width="55%"}
An **export contract** specifies: price, quantity, quality, delivery terms, payment mode, and dispute resolution.

**Incoterms 2020** (ICC, Paris) — 11 standardised trade terms that define exactly **where risk and cost transfer** from seller to buyer.

Key terms for Indian agri exports:

| Term | Risk Transfers At | Used For |
|---|---|---|
| [EXW]{.hi} | Seller's premises | Rare in agri |
| [FCA]{.hi} | Named place to carrier | Air/rail shipments |
| [FOB]{.hi-gold} | On board vessel | **Most Indian agri** |
| [CIF]{.hi} | Destination port | Buyer-preferred |
| [DDP]{.hi} | Buyer's premises | E-commerce niche |
:::
::: {.column width="45%"}
::: {.highlightbox}
**FOB vs CIF for India:**

Under **FOB**, the Indian exporter's responsibility ends when goods cross the ship's rail. Freight and insurance arranged by the buyer.

Under **CIF**, the exporter arranges (and prices) freight and insurance to the destination port.

**~65% of India's agri exports are on FOB terms.** CIF is preferred by Gulf and African buyers who have captive freight arrangements.

**Negotiation tip:** FOB is simpler for small exporters; CIF offers a higher invoice value and more price control.
:::
:::
:::

---

## Export Documents: Proof of Transaction

- **Commercial Invoice** — product, quantity, price, HSN/IEC; basis for customs valuation
- **Packing List** — box-by-box weight and dimensions
- **Transport document** — **B/L** (sea) or **AWB** (air)
- **Certificate of Origin (CoO)** — used for FTA preferential tariffs
- **Key idea:** **B/L is negotiable** (can be endorsed/transferred)

:::{.notes}
Detailed descriptions:

1. **Commercial Invoice** — price, quantity, HSN code, IEC, buyer/seller details; basis for customs valuation and bank payment verification under LC.
2. **Packing List** — box-by-box itemisation of weight, dimensions, marks and numbers; used by customs and buyer for checking receipt.
3. **Bill of Lading (B/L)** — shipping company's receipt and title document; original needed to claim cargo at destination port; negotiable instrument that can be endorsed and transferred.
4. **Airway Bill (AWB)** — non-negotiable air freight equivalent of B/L; issued by airline; consignee named directly.
5. **Certificate of Origin (CoO)** — issued by FIEO, chamber of commerce, or APEDA; proves country of origin; required for preferential duty rate under FTAs.
:::

---

## Export Documents: Compliance and Regulatory

- **Phytosanitary certificate (PPQS)** — plant products are pest-free
- **Health/inspection certificate (EIC)** — meat/marine/organic/dairy
- **Fumigation certificate** — pulses/oilseeds; treatment details
- **Shipping Bill (ICEGATE)** — customs clearance + **LEO**
- **EDPMS (RBI)** — links shipment to forex realization

:::{.notes}
Detailed descriptions:

6. **Phytosanitary Certificate** — issued by Directorate of Plant Protection, Quarantine & Storage (PPQS) under DAPFW; certifies products are pest-free; mandatory for all plant and plant-product exports.
7. **Health / Inspection Certificate** — issued by Export Inspection Council (EIC) for meat, marine, organic, and dairy exports; confirms product meets importing country food safety standards.
8. **EDPMS declaration** (Export Data Processing and Monitoring System) — RBI's online system; all export shipments must be registered; links shipment data to forex realization within 9 months.
9. **Fumigation Certificate** — mandatory for pulses, oilseeds, and many cereals; issued by certified fumigators; many countries require methyl bromide or phosphine treatment; certificate must specify chemical, dosage, duration.
10. **Shipping Bill** — the key customs clearance document filed on ICEGATE; must be approved by Customs before loading; distinct types (Free Shipping Bill, Dutiable SB, SB under claim for Drawback) for different incentive claims.
:::

---

## Document flow: a typical agri export

1. Contract signed + invoice prepared
2. Buyer arranges payment (often via **LC**)
3. Shipment booked → **B/L or AWB** issued
4. Certificates obtained + Shipping Bill filed → **LEO**
5. Documents presented to bank → payment + **EDPMS** closure

---

## Letter of Credit: Reducing Counterparty Risk

::: {.columns}
::: {.column width="55%"}
**Why LC?** In international trade, buyer and seller do not know each other. The LC inserts two banks as guarantors.

**Step-by-step LC mechanism:**

1. Buyer and seller agree on price and LC terms
2. **Buyer's bank (Issuing Bank)** issues LC in favour of exporter
3. **Advising Bank** (exporter's bank in India) notifies exporter; may also confirm
4. Exporter ships goods; presents documents to advising bank
5. Documents checked against LC terms; bank pays exporter
6. Documents forwarded to issuing bank; buyer pays and gets documents
7. Buyer presents B/L to shipping line → collects cargo
:::
::: {.column width="45%"}
::: {.methodbox}
**Types of LC:**

- **Sight LC:** Exporter paid immediately on document presentation
- **Usance/DA LC:** Payment deferred (30, 60, 90 days) — extends credit to buyer
- **Revolving LC:** Renews automatically for repeat shipments
- **Back-to-Back LC:** Small exporters use master LC as collateral to open LC for their suppliers

**India's EXIM Bank:** Pre-shipment and post-shipment credit; refinances AD banks for agri exporters.

**ECGC:** Export Credit Guarantee Corporation — insures against buyer default and country risk. Premium: ~0.5–1% of invoice value.
:::
:::
:::

---

## Export Credit: Pre vs Post Shipment Trend

```{r}
#| label: fig-export-credit
#| fig-cap: "Pre-shipment credit dipped during COVID (FY2021) and recovered alongside export growth; post-shipment credit tracks pre-shipment with a structural lag.\nSource: RBI, Handbook of Statistics on Indian Economy."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(dplyr)
library(tidyr)

df_credit <- data.frame(
  Year = factor(
    c("FY2018","FY2019","FY2020","FY2021","FY2022","FY2023","FY2024"),
    levels = c("FY2018","FY2019","FY2020","FY2021","FY2022","FY2023","FY2024")),
  Pre_Shipment_B  = c(1.2, 1.3, 1.1, 0.9, 1.4, 1.7, 1.9),
  Post_Shipment_B = c(0.8, 0.9, 0.7, 0.6, 0.9, 1.1, 1.2)
)

df_long <- df_credit |>
  tidyr::pivot_longer(cols = c(Pre_Shipment_B, Post_Shipment_B),
                      names_to = "Type", values_to = "Value_B") |>
  dplyr::mutate(Type = ifelse(Type == "Pre_Shipment_B",
               "Pre-Shipment Credit", "Post-Shipment Credit"))

ggplot(df_long, aes(x = Year, y = Value_B, colour = Type, fill = Type, group = Type)) +
  geom_area(alpha = 0.15, position = "identity") +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  annotate("rect", xmin = 3.5, xmax = 4.5, ymin = 0, ymax = 2.4, fill = "grey90", alpha = 0.4) +
  annotate("text", x = 4, y = 2.2, label = "COVID\ndip", size = 2.9, colour = "grey45", hjust = 0.5) +
  scale_colour_manual(values = c("Pre-Shipment Credit" = "#012169", "Post-Shipment Credit" = "#B9975B")) +
  scale_fill_manual(values = c("Pre-Shipment Credit" = "#012169", "Post-Shipment Credit" = "#B9975B")) +
  scale_y_continuous(limits = c(0, 2.4), labels = function(x) paste0("$", x, "B")) +
  labs(title = "India's Export Credit: Pre vs Post Shipment (USD Billion)",
       subtitle = "Scheduled commercial bank credit outstanding for agricultural export finance",
       x = NULL, y = "USD Billion", colour = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"),
        axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Customs Clearance: ICEGATE and the Shipping Bill

::: {.columns}
::: {.column width="55%"}
**ICEGATE** (Indian Customs EDI Gateway) — the single portal for all Indian customs transactions.

**Shipping Bill types:**

| Type | When Used |
|---|---|
| **Drawback SB** | Claiming duty drawback on inputs |
| **DEEC/Advance Auth SB** | Against advance authorisation |
| **EP Copy SB** | Ordinary exports |
| **Free Shipping Bill** | Personal effects, gifts |

**Process flow:**
1. Exporter files Shipping Bill online with all document details
2. Customs Risk Management System (RMS) assesses risk
3. Low-risk: **Green Channel** → LEO (Let Export Order) in 2–4 hours
4. Medium-risk: **Yellow Channel** → document examination
5. High-risk: **Red Channel** → physical examination of cargo
:::
::: {.column width="45%"}
::: {.highlightbox}
**LEO — Let Export Order:**

The LEO is the customs officer's green light for loading. Cargo **cannot board** without LEO. Once LEO is given:

- Exporter has 24 hours to load at port
- Date of LEO = official date of export for FEMA and incentive purposes

**AEO Status (Authorised Economic Operator):**

Exporters with strong compliance history can apply for AEO status (Tier I, II, III). Benefits:
- Fast-track customs clearance
- Reduced physical examination
- Priority in port queue
- Deferred duty payment

In FY2024, ~12,000 entities hold AEO status in India.
:::
:::
:::

---

## India's Merchandise Exports by Mode of Transport

```{r}
#| label: fig-export-mode
#| fig-cap: "Sea freight dominates at ~68% of India's merchandise exports; air (~15%) and land (~12%) are secondary modes. Share composition is stable across both years.\nSource: DGCI&S / Ministry of Commerce and Industry, GoI."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(dplyr)

df_mode <- data.frame(
  Mode    = rep(c("Sea", "Air", "Land", "Other"), 2),
  Year    = rep(c("FY2022-23", "FY2023-24"), each = 4),
  Value_B = c(297, 67, 52, 21, 310, 71, 56, 25)
) |>
  dplyr::group_by(Year) |>
  dplyr::mutate(
    Share = round(Value_B / sum(Value_B) * 100, 1),
    Mode  = factor(Mode, levels = c("Other", "Land", "Air", "Sea"))
  ) |>
  dplyr::ungroup()

ggplot(df_mode, aes(x = Share, y = Mode, fill = Year)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  geom_text(aes(label = paste0(Share, "% ($", Value_B, "B)")),
    position = position_dodge(width = 0.72), hjust = -0.07, size = 3.1) +
  scale_fill_manual(values = c("FY2022-23" = "#4a7fd4", "FY2023-24" = "#012169")) +
  scale_x_continuous(limits = c(0, 86), labels = function(x) paste0(x, "%")) +
  labs(title = "India's Merchandise Exports by Mode of Transport (FY2023)",
       subtitle = "Total exports FY2022-23 = USD 437 Billion  |  FY2023-24 = USD 462 Billion",
       x = "Share of Total Merchandise Exports (%)", y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

---

## Direct vs Indirect Export Channels

::: {.columns}
::: {.column width="50%"}
**Direct Export:**
- Exporter deals directly with foreign buyer
- Higher margin, full control over branding/pricing
- Higher risk (counterparty, forex, compliance)
- Requires: IEC, RCMC, AD bank, full documentation
- Suited to: large processors, established exporters

**Examples:** Kohinoor Foods (basmati direct to UAE retail), Venkateshwara Hatcheries (poultry to Middle East)
:::
::: {.column width="50%"}
**Indirect Export:**
- Sell to a **Merchant Exporter (ME)** who manages the export
- No direct forex exposure; simpler for small farmer/processor
- Lower price realisation (merchant takes margin)
- ME provides GST exemption on purchase from manufacturer at 0.1% CGST+SGST

**Export Management Company (EMC):** manages exports on commission without taking title to goods.

**FPO Route:**
::: {.keybox}
**Farmer Producer Organisations (FPOs)** aggregate smallholders (avg. 1.5 ha in Karnataka) to meet minimum export quantity. APEDA's FPO Connect links 1,200+ FPOs to export markets. Tur dal FPOs in Gulbarga district now export directly to UK Indian grocery wholesalers.
:::
:::
:::

---

## Export Houses: India's Trade Titans

::: {.columns}
::: {.column width="55%"}
The Government of India recognises export performance tiers with **Star Export House** status, conferring priority clearances, bank credit, and EPCG benefits:

| Status | Min. Export (3-yr avg) | Benefits |
|---|---|---|
| Export House | ₹100 crore | Priority treatment |
| Trading House | ₹500 crore | Self-certification of CoO |
| Star Trading House | ₹2,000 crore | Fast-track customs, AEO |
| Premier Trading House | ₹10,000 crore | All of above + policy voice |
:::
::: {.column width="45%"}
::: {.highlightbox}
**Major agri export houses (India):**

- **ITC Agribusiness:** wheat, spices, processed food; exports to 60+ countries; $1B+ annual turnover
- **Olam International:** cashew, spices, coffee; integrated from farm to retail
- **Adani Agri Fresh:** apple, potato, processed; cold chain integration
- **KRBL Ltd:** world's largest basmati rice brand (India Gate); exports to 90+ countries
- **Apex Frozen Foods:** shrimp to USA/EU; vertically integrated aquaculture

These firms account for **~35%** of India's total agricultural export value.
:::
:::
:::

---

## Export Incentive Schemes

::: {.columns}
::: {.column width="50%"}
**1. RoDTEP (Remission of Duties and Taxes on Exported Products)**
- Replaced MEIS in January 2021 (WTO-compliant)
- Refunds embedded taxes not otherwise refunded: SGST, mandi tax, electricity duty, stamp duty
- Agri rates: **0.5% to 4.3% of FOB value**
- Issued as transferable scrips on ICEGATE

**2. Advance Authorisation (AA)**
- Duty-free import of inputs used in export production
- Standard Input Output Norms (SION) define input ratios
- Example: basmati rice exporter imports paddy from Nepal duty-free under AA

**3. EPCG (Export Promotion Capital Goods)**
- Import capital goods (cold storage, processing equipment) at **0–3% duty**
- In return, commit to export 6× the duty saved over 6 years
:::
::: {.column width="50%"}
**4. Transport and Marketing Assistance (TMA)**
- Freight subsidy for agri exports to notified markets (Africa, EU, USA)
- Cover: 50% of sea freight for small exporters; 25% for large
- Products: fresh fruits/vegetables, processed food, organic, floriculture

**5. MAI Scheme (Market Access Initiative)**
- APEDA funds participation in international trade fairs (Gulfood Dubai, SIAL Paris, Annapoorna Mumbai)
- Buyer-seller meets, market studies, reverse buyer visits
- Budget: ₹250 crore/year for all export promotion councils

::: {.keybox}
**WTO compliance note:** RoDTEP is WTO-consistent (remission of domestic taxes, not export subsidy). AA and EPCG are permissible under WTO SCM Agreement as tax exemptions on inputs. TMA freight subsidy is under scrutiny — USA has raised concerns at WTO DSB.
:::
:::
:::

---

## FSSAI and Food Safety Compliance for Exports

::: {.columns}
::: {.column width="55%"}
**FSSAI (Food Safety and Standards Authority of India)**

- Central licensing authority for all food businesses including exporters
- An **FSSAI Central Licence** is mandatory for manufacturers/processors exporting food products
- Annual turnover > ₹20 crore or multi-state operation → Central Licence

**FSSAI's role in exports:**

1. **Standards alignment:** FSSAI standards increasingly aligned with Codex Alimentarius (facilitates market access)
2. **Organic certification:** FSSAI-accredited certification bodies issue Indian Organic Certificate (recognises NPOP)
3. **Bilateral equivalence:** FSSAI has equivalence arrangements with USDA NOP (organic) — US organic certifiers can certify Indian organic products
:::
::: {.column width="45%"}
::: {.methodbox}
**EIC accreditation for key markets:**

The **Export Inspection Council (EIC)** issues health certificates that importing countries accept:

| Market | Certificate | Requirement |
|---|---|---|
| European Union | EIC Health Cert. | Meat, fish, dairy |
| USA (FDA) | EIC + prior notice | Marine, processed |
| Japan | EIC + residue test | Marine, processed |
| Middle East | Halal + EIC | Meat, poultry |

EIC operates 5 regional offices (Mumbai, Chennai, Kochi, Kolkata, Delhi) and 40 sub-offices. In FY2024, EIC certified **₹1.1 lakh crore** worth of exports.
:::
:::
:::

---

# Case Study {.section}

::: {style="text-align:center; margin-top:1em;"}
**Exporting Basmati Rice from Punjab to Saudi Arabia**

*Tracing every document, agency, and payment from paddy field to Riyadh shelf*
:::

---

## Case Study: Punjab Basmati to Saudi Arabia — Part I

::: {.columns}
::: {.column width="55%"}
**The exporter:** Amritsar-based rice miller; 10,000 tonne annual capacity. Target: 500 MT to Saudi retail chain.

**Step 1 — Registrations:**
- IEC from DGFT portal (already held)
- APEDA RCMC (renewed annually; fee ₹7,500)
- FSSAI Central Licence
- AD Bank: State Bank of India, Amritsar branch

**Step 2 — Contract:**
- Incoterm: **CIF Jeddah**
- Price: $1,100/MT; Total: $550,000
- Payment: **Sight LC** from Riyad Bank (Saudi), advised through SBI

**Step 3 — Procurement:**
- Procures paddy from Amritsar mandi at MSP + premium
- Milling, grading, drying; APEDA quality check (aroma, grain length)
:::
::: {.column width="45%"}
**Step 4 — Documents prepared:**

| # | Document | Issuing Agency |
|---|---|---|
| 1 | Commercial Invoice | Exporter |
| 2 | Packing List | Exporter |
| 3 | Certificate of Origin | FIEO/Chamber |
| 4 | Phytosanitary Certificate | PPQS Punjab |
| 5 | Fumigation Certificate | Licensed fumigator |
| 6 | Quality/Weight Certificate | EIC / SGS |
| 7 | APEDA Certificate | APEDA |
| 8 | B/L | Shipping line (NSICT) |
| 9 | EDPMS export declaration | RBI system (auto) |
| 10 | Shipping Bill | ICEGATE (Customs) |
:::
:::

---

## Case Study: Punjab Basmati to Saudi Arabia — Part II

::: {.columns}
::: {.column width="55%"}
**Step 5 — Customs clearance (JNPT, Mumbai):**
- Exporter files Shipping Bill on ICEGATE: **Drawback type** (to claim duty drawback on GST paid on packaging)
- RMS assigns **Green Channel** — LEO granted within 3 hours
- Container loaded; B/L issued by Maersk

**Step 6 — Banking:**
- Exporter presents all 10 documents to SBI within **21 days of B/L date** (LC requirement)
- SBI verifies documents against LC terms: ✓
- SBI claims payment from Riyad Bank
- **Sight LC:** SBI receives USD 550,000 within 5 working days
- SBI converts USD to INR; credits exporter account

**Step 7 — Post-shipment:**
- RBI's EDPMS marks export as "realised"
- Exporter files **e-BRC** (electronic Bank Realisation Certificate) on DGFT portal
- RoDTEP scrip of ~₹11 lakh (2% of FOB) issued
:::
::: {.column width="45%"}
::: {.highlightbox}
**Saudi Arabia end:**

- Saudi SFDA (food authority) inspects on arrival
- Certificate of Origin checked against GCC preferential tariff rules
- Phytosanitary certificate accepted by Saudi Plant Protection
- If fumigation certificate absent → shipment quarantined

**Typical timeline (farm to Riyadh shelf):**

| Stage | Days |
|---|---|
| Procurement + milling | 15 |
| Documentation | 5 |
| Port + customs | 3 |
| Sea transit (JNPT–Jeddah) | 8 |
| Saudi customs clearance | 5 |
| Distribution to shelf | 7 |
| **Total** | **~43 days** |
:::
:::
:::

---

## Import Procedures: The Mirror Image

::: {.columns}
::: {.column width="55%"}
Import procedures mirror export procedures on the buyer side.

**Key documents for agricultural imports to India:**

1. **Bill of Entry** — filed on ICEGATE by importer (equivalent of Shipping Bill)
2. **Commercial Invoice and Packing List**
3. **Bill of Lading / Airway Bill**
4. **Certificate of Origin** — for FTA preferential duty
5. **SPS/Phytosanitary Certificate** — from exporting country's authority

**Customs assessment:**
- RMS assigns risk category
- Basic Customs Duty + IGST (at GST rate) assessed
- IGST paid by importer; can claim ITC if registered
:::
::: {.column width="45%"}
::: {.methodbox}
**SPS inspection for agricultural imports:**

Under the **Plant Quarantine Order 2003** and **Meat Food Products Order**, all agri imports are subject to:

- Document verification (phytosanitary cert)
- Physical sampling and lab testing
- For pests: fumigation before release

**Restricted/Prohibited imports:**

- **STE channel:** wheat, rice, sugar, fertilisers — importable only through state trading enterprises (FCI, NAFED, MMTC) under quantitative restrictions
- **License required:** some oilseeds, livestock products
- **Prohibited:** live pigs (several ports), certain livestock breeds without quarantine

**Bill of Entry types:** Home Consumption (B/E), Warehousing (W/E), Ex-Bond (E/B).
:::
:::
:::

---

## Logistics and Infrastructure Challenges

::: {.columns}
::: {.column width="55%"}
India's **Logistics Performance Index (LPI)** rank: **38 out of 139** (World Bank, 2023) — up from rank 44 in 2018.

**Key bottlenecks for agricultural exports:**

**1. Port congestion:**
- JNPT handles 55% of India's container traffic; average dwell time: **3.2 days** (global best: 1.5 days)
- Mundra, Chennai, Kochi also congested during harvest seasons

**2. Cold chain deficit:**
- India's cold storage capacity: **~35 million MT** (FY2024)
- Requirement for horticulture + marine: **~90 million MT**
- Gap = 55 million MT → 25–40% post-harvest losses in fruits and vegetables

**3. Last-mile connectivity:**
- Farm-to-port road quality: NH coverage improving, but mandi-to-cold store roads often unpaved in Karnataka/Maharashtra
:::
::: {.column width="45%"}
::: {.highlightbox}
**Cost of logistics as % of GDP:**

| Country | Logistics Cost / GDP |
|---|---|
| Germany | 8% |
| USA | 8% |
| China | 14% |
| **India** | **~14%** |

High logistics costs reduce India's price competitiveness despite low labour costs.

**PM GatiShakti:** National Master Plan integrating roads, railways, ports, warehouses. Target: reduce logistics cost to **8% of GDP by 2030**.

**Dedicated Freight Corridors (DFC):** Eastern (Ludhiana–Kolkata) and Western (JNPT–Dadri) operational 2024 — will cut rail freight time by 40% for Punjab basmati, Maharashtra grapes.
:::
:::
:::

---

## Digital Initiatives in Export Trade

::: {.columns}
::: {.column width="55%"}
**DGFT Digital Modernisation (2021–ongoing):**

- Fully online IEC issuance: **24 hours** (was 3 weeks paper-based)
- Online RCMC by APEDA: auto-linked to IEC
- e-BRC (electronic Bank Realisation Certificate): exporter's bank submits directly to DGFT portal; no paper filing
- **DARPAN platform:** dashboard for tracking all incentive scheme status in real time

**SWIFT (Single Window Interface for Facilitating Trade):**
- One portal for customs, FSSAI, PPQS, Drug Controller, textile ministry clearances
- Launched 2016; covers 26 regulatory agencies
- 2024 expansion: includes MPEDA, APEDA certificates integrated
:::
::: {.column width="45%"}
::: {.methodbox}
**ICEGATE 2.0 (launched 2023):**

- AI-based risk management system → 70% of shipments now on Green Channel (was 55%)
- e-Sanchit: paperless document upload — all 10 export documents can be uploaded digitally; customs officer reviews on-screen
- Integration with GSTN: automatic input tax credit verification for RoDTEP computation

**APEDA-DGFT integration:**

APEDA's NHB system auto-generates RCMC data to DGFT; exporters need not re-enter. Certificate of Origin auto-populated from ICEGATE data for FTA claims.

::: {.keybox}
India's trade facilitation score (WTO TFA): **90.59%** (2023) — above developing-country average of 74%.
:::
:::
:::
:::

---

## Summary: The Export Ecosystem

::: {.columns}
::: {.column width="55%"}
**The full export chain for Indian agri:**

```
Farmer/Processor
      ↓ FPO aggregation / merchant exporter
IEC + RCMC + FSSAI + AD Bank
      ↓ pre-shipment
Export Contract (Incoterm negotiated)
      ↓ finance
Letter of Credit (bank guarantee)
      ↓ documentation
10 documents (Invoice, B/L, CoO, Phyto, EIC...)
      ↓ customs
ICEGATE Shipping Bill → LEO
      ↓ logistics
Port → Ship → Destination customs
      ↓ payment
Bank realises forex → e-BRC filed
      ↓ incentives
RoDTEP scrip + drawback + TMA
```
:::
::: {.column width="45%"}
::: {.keybox}
**Key Takeaways:**

1. **IEC is the master key** — no export without it
2. **The Shipping Bill + LEO** is the customs clearance milestone — nothing loads without it
3. **The Bill of Lading** is the only negotiable title document — it unlocks payment under LC
4. **RoDTEP** has replaced MEIS as the primary export incentive — WTO-compliant, duty refund basis
5. **Cold chain and logistics** remain India's primary structural bottleneck — world-class documentation systems exist, but physical infrastructure lags
:::

:::{.notes}
6. **Digital India** is working: ICEGATE, SWIFT, e-BRC have dramatically reduced transaction costs
:::
:::
:::

---

## Next Lecture

::: {.methodbox}
**Lecture 17 — Agricultural Export Promotion Organizations**  
*August 15, 2026*

- APEDA: mandate, functions, financial assistance, market access achievements
- MPEDA and India's $7.6 billion seafood export ecosystem
- The five Commodity Boards: Tea, Coffee, Rubber, Spices, Tobacco
- Export Inspection Council and quality certification
- NAFED: buffer stocks, MSP procurement, export implications
:::

:::{.notes}
- Farmer Producer Organizations as export units
- India's 2030 export target: $60 billion in agri and food products

**Preparation:** Review APEDA Annual Report 2023–24 (available at apeda.gov.in). Note the top 10 products by export value and the top 5 destination countries.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- Lecture notes and APEDA/WTO official documents
- DGFT: Foreign Trade Policy 2023-28
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
