---
title: "Lecture 10: India's External Sector"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "June 27, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
    code-fold: true
    code-summary: "Show R code"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(patchwork)
options(scipen = 999)
theme_set(theme_minimal(base_size = 11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## India's External Sector {background-gradient="linear-gradient(135deg, #1a3c5e 0%, #2d6a9f 100%)"}

::: {style="color: white; font-size: 1.1em; margin-top: 1em;"}
A deep dive into India's trade structure, agricultural trade, and balance of payments
:::

## Overview of India's External Sector

::: {.columns}
::: {.column width="60%"}
India is a **major trading nation** — ranked among the top 20 exporters and importers globally.

Key FY2024 aggregates:

- **Merchandise exports:** $437 billion
- **Merchandise imports:** $677 billion
- **Trade deficit:** −$240 billion
- **Services exports:** $338 billion
- **Remittances:** $120 billion

:::
::: {.column width="40%"}
::: {.keybox}
**Why does this matter for agriculture?**

India's external sector shapes:

- Price of edible oils (imports)
- Farmer income (export demand)
- Food inflation (rupee/import costs)
- BoP sustainability
:::
:::
:::

## Merchandise Exports: Structure FY2024

::: {.methodbox}
**Top 5 Merchandise Export Categories — FY2024**
:::

| Category | Value (USD B) | Share of Total |
|---|---|---|
| Engineering goods | $109 | 14.0% |
| Petroleum products | $96 | 12.3% |
| **Agriculture & allied** | **$43.7** | **5.6%** |
| Gems & jewellery | $37 | 4.8% |
| Chemicals | $32 | 4.1% |
| **Total merchandise** | **$437** | |

::: {.fragment}
::: {.hi}
Agriculture contributes ~5.6% of total merchandise exports — a modest but strategically important share.
:::
:::

## Merchandise Export Composition: FY2024

```{r}
#| label: fig-export-composition
#| fig-cap: "India's Merchandise Export Composition, FY2024 (USD billion)\nSource: DGCI&S / Ministry of Commerce and Industry, GoI."
#| fig-width: 9
#| fig-height: 5

export_data <- data.frame(
  sector   = c("Engineering Goods", "Petroleum Products", "Gems & Jewellery",
               "Chemicals & Pharma", "Agriculture & Allied",
               "Textiles & Apparel", "Electronics", "Other"),
  value    = c(109.3, 78.5, 35.4, 65.2, 43.7, 33.4, 23.6, 17.9),
  category = c("Manufacturing", "Petroleum", "Manufacturing", "Manufacturing",
               "Agriculture", "Manufacturing", "Manufacturing", "Other")
)

export_data <- export_data |>
  dplyr::arrange(desc(value)) |>
  dplyr::mutate(sector = factor(sector, levels = sector))

ggplot(export_data, aes(x = sector, y = value, fill = category)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = paste0("$", value, "B")), vjust = -0.3, size = 3.2) +
  scale_fill_manual(
    values = c("Manufacturing" = "#012169", "Petroleum" = "#B9975B",
               "Agriculture"   = "darkgreen", "Other"   = "grey60")) +
  labs(title = "India's Merchandise Exports by Sector, FY2024 (Total: ~$437B)",
       x = NULL, y = "USD Billion", fill = "Sector") +
  scale_y_continuous(limits = c(0, 125)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

::: {.fragment}
::: {.hi-gold}
Engineering goods ($109B) dominate — India's shift toward capital goods exports. Agriculture ($44B) is the 5th largest category, but holds outsized strategic importance for rural incomes and food security diplomacy.
:::
:::

## Merchandise Imports: Structure FY2024

::: {.columns}
::: {.column width="55%"}

**Top Import Categories:**

| Category | USD B |
|---|---|
| Petroleum & crude oil | $232 |
| Engineering goods | $91 |
| Electronic goods | $80 |
| Gold | $47 |
| **Edible oils** | **$14** |
| Pulses | $2.8 |

:::
::: {.column width="45%"}
::: {.keybox}
**Agricultural import vulnerability**

Edible oils at $14B represent a structural dependency — India imports ~60–65% of its edible oil needs.

This is India's single largest agricultural import burden.
:::
:::
:::

## Merchandise Import Composition: FY2024

```{r}
#| label: fig-import-composition
#| fig-cap: "India's Merchandise Import Composition, FY2024 (USD billion)\nSource: DGCI&S / Ministry of Commerce and Industry, GoI."
#| fig-width: 8
#| fig-height: 4.5

import_data <- data.frame(
  sector    = c("Crude Oil &\nPetroleum", "Gold & Silver", "Edible Oil",
                "Electronics &\nMachinery", "Chemicals", "Coal",
                "Fertilizers", "Other"),
  value     = c(227.8, 47.8, 14.2, 138.4, 38.6, 34.2, 18.9, 72.1),
  vulnerable = c(TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE, FALSE)
)

import_data <- import_data |>
  dplyr::arrange(desc(value)) |>
  dplyr::mutate(sector = factor(sector, levels = sector))

ggplot(import_data, aes(x = sector, y = value, fill = vulnerable)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = paste0("$", value, "B")), vjust = -0.3, size = 3.2) +
  scale_fill_manual(
    values = c("TRUE" = "#cc0000", "FALSE" = "#012169"),
    labels = c("TRUE" = "Commodity vulnerability", "FALSE" = "Strategic import")) +
  labs(title    = "India's Merchandise Imports by Sector, FY2024 (Total: ~$692B)",
       subtitle = "Red = commodity import vulnerabilities (oil, edible oil, coal, fertilizers)",
       x = NULL, y = "USD Billion", fill = NULL) +
  scale_y_continuous(limits = c(0, 250)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

::: {.fragment}
::: {.hi}
**Vulnerability arithmetic:** Crude oil ($228B) + Coal ($34B) + Fertilizers ($19B) + Edible oil ($14B) = **~$295B in commodity imports** — 43% of total imports are subject to global commodity price shocks beyond India's control.
:::
:::

## Agricultural Exports FY2024: Breakdown

::: {.highlightbox}
**India's total agricultural exports FY2024: ~$43.7 billion**
:::

::: {.columns}
::: {.column width="50%"}
| Commodity | USD B |
|---|---|
| Marine products | $7.6 |
| **Non-basmati rice** | **$6.4** |
| Basmati rice | $5.8 |
| Processed food | $4.8 |
:::
::: {.column width="50%"}
| Commodity | USD B |
|---|---|
| Spices | $3.7 |
| Cotton | $2.7 |
| Coffee | $1.2 |
| Others | ~$11.5 |
:::
:::

::: {.fragment}
::: {.hi-gold}
⚠️ Non-basmati rice exports were restricted from August 2023, significantly curtailing export earnings.
:::
:::

## Agricultural Imports FY2024: Breakdown

**Total agricultural imports: ~$28.2 billion**

::: {.columns}
::: {.column width="55%"}
**Edible Oils — $14 billion (largest component)**

| Oil | USD B |
|---|---|
| Palm oil | $9.5 |
| Soybean oil | $2.5 |
| Sunflower oil | $2.0 |

:::
::: {.column width="45%"}
**Other major imports:**

- Pulses: **$2.8B**  
  *(Canada, Australia key sources)*
- Fresh fruits: **$1.8B**
- Cashew (raw): ~$1.5B
- Spices (for re-export): ~$0.8B

:::
:::

## India's Net Agricultural Trade Position

::: {.methodbox}
**Agricultural Trade Balance FY2024**

Exports $43.7B − Imports $28.2B = **Surplus +$15.5B**
:::

**Five-Year Trend in Agricultural Trade Surplus:**

| FY | Surplus (USD B) |
|---|---|
| FY2020 | $17.0 |
| FY2021 | $20.0 |
| FY2022 | $25.0 |
| FY2023 | $15.8 |
| FY2024 | $15.5 |

::: {.fragment}
::: {.hi}
The sharp drop from FY2022 to FY2023 reflects global commodity price corrections and India's rice export restrictions.
:::
:::

## India's Key Merchandise Trading Partners

::: {.columns}
::: {.column width="50%"}
**Top Export Destinations**

1. 🇺🇸 USA — $77B
2. 🇦🇪 UAE — $35B
3. 🇳🇱 Netherlands — $23B
4. 🇬🇧 UK — $19B
5. 🇧🇩 Bangladesh — $14B

:::
::: {.column width="50%"}
**Top Import Sources**

1. 🇨🇳 China — $101B
2. 🇷🇺 Russia — $61B
3. 🇦🇪 UAE — $48B
4. 🇺🇸 USA — $42B
5. 🇮🇶 Iraq — $22B
:::
:::

::: {.keybox}
**China dominates imports** — especially electronics, machinery, chemicals. India's trade deficit with China alone is ~$85B. Reducing this is a stated policy goal.
:::

## India's Trade Partners: FY2024

```{r}
#| label: fig-trade-partners
#| fig-cap: "India's Top 5 Export Destinations and Import Sources, FY2024\nSource: DGCI&S / Ministry of Commerce and Industry, GoI."
#| fig-width: 9
#| fig-height: 5

exports_dest <- data.frame(
  country = c("USA", "UAE", "Netherlands", "China", "UK"),
  value   = c(77.5, 35.5, 18.2, 16.7, 11.1),
  share   = c(17.8,  8.1,  4.2,  3.8,  2.5)
)

imports_src <- data.frame(
  country = c("China", "UAE", "Russia", "USA", "Saudi Arabia"),
  value   = c(101.7, 53.2, 44.3, 40.8, 33.5),
  share   = c(14.7,   7.7,  6.4,  5.9,  4.8)
)

p1 <- ggplot(
  exports_dest |> dplyr::arrange(value) |>
    dplyr::mutate(country = factor(country, levels = country)),
  aes(x = country, y = value)) +
  geom_col(fill = "#012169", width = 0.7) +
  geom_text(aes(label = paste0("$", value, "B\n(", share, "%)")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  labs(title = "Top Export Destinations", x = NULL, y = "USD Billion") +
  scale_y_continuous(limits = c(0, 92)) +
  theme_minimal(base_size = 10)

p2 <- ggplot(
  imports_src |> dplyr::arrange(value) |>
    dplyr::mutate(country = factor(country, levels = country)),
  aes(x = country, y = value)) +
  geom_col(fill = "#B9975B", width = 0.7) +
  geom_text(aes(label = paste0("$", value, "B\n(", share, "%)")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  labs(title = "Top Import Sources", x = NULL, y = "USD Billion") +
  scale_y_continuous(limits = c(0, 125)) +
  theme_minimal(base_size = 10)

p1 + p2 +
  plot_annotation(
    title    = "India's Trade Partners, FY2024",
    subtitle = "USA = top export destination; China = top import source — strategic asymmetry"
  )
```

::: {.fragment}
::: {.hi-gold}
**Strategic asymmetry:** India exports most to the USA ($78B) but imports most from China ($102B). The bilateral trade deficit with China alone (~$85B) exceeds India's entire trade surplus with the USA.
:::
:::

## Agricultural Trade Partners

::: {.columns}
::: {.column width="50%"}
**Top Agri Export Markets**

1. USA — marine, spices, processed
2. Bangladesh — rice, food grains
3. UAE — rice, vegetables, spices
4. China — cotton, marine, soybean
5. Saudi Arabia — rice, spices

*Rice dominates exports to South/SE Asia*
:::
::: {.column width="50%"}
**Top Agri Import Sources**

1. Indonesia — palm oil
2. Malaysia — palm oil
3. Canada — pulses, lentils
4. Australia — pulses, wheat (intermittent)
5. Ukraine/Argentina — sunflower/soybean oil

:::
:::

::: {.hi-gold}
India's agricultural trade geography reflects its structural reliance on SE Asia for edible oils and the Anglosphere for pulses.
:::

## Services Exports: The Other Engine

::: {.highlightbox}
**Services exports FY2024: $338 billion** — India's invisible export superpower
:::

::: {.columns}
::: {.column width="55%"}
| Services Category | USD B |
|---|---|
| IT/Software services | $220 |
| Business process mgmt | $38 |
| Financial services | $20 |
| Travel & tourism | $18 |
| Other services | $42 |

:::
::: {.column width="45%"}
::: {.methodbox}
**Why this matters for agri trade:**

IT export earnings provide forex that finances agricultural imports (edible oils, pulses).

Without IT services surplus, India's BoP position would be far more stressed.
:::
:::
:::

## Services Trade Composition: FY2024

```{r}
#| label: fig-services-trade
#| fig-cap: "India's Services Trade Composition, FY2024 (USD billion)\nSource: RBI, Balance of Payments Statistics."
#| fig-width: 8
#| fig-height: 4.5

services_data <- data.frame(
  category = c("Software &\nIT", "Business\nServices", "Travel\n(Tourism)",
               "Transport", "Financial\nServices", "Education\n& Health"),
  exports  = c(194.0, 31.2, 28.1, 14.6, 11.5, 8.3),
  imports  = c( 11.5, 20.4, 20.2, 27.8,  9.1, 5.3)
)

svc_long <- tidyr::pivot_longer(services_data,
                                cols      = c(exports, imports),
                                names_to  = "type",
                                values_to = "value")

ggplot(svc_long, aes(x = category, y = value, fill = type)) +
  geom_col(position = "dodge", width = 0.75) +
  scale_fill_manual(
    values = c("exports" = "#012169", "imports" = "#B9975B"),
    labels = c("exports" = "Service Exports", "imports" = "Service Imports")) +
  labs(title    = "India's Services Trade, FY2024 (USD Billion)",
       subtitle = "Software/IT ($194B) dominates — India's 'comparative advantage' in skilled services",
       x = NULL, y = "USD Billion", fill = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

::: {.fragment}
::: {.hi}
**Revealed comparative advantage:** India's RCA in Software/IT services is extremely high — the $194B export vs $12B import reflects a structural specialisation that offsets the entire merchandise trade deficit ($76B) by itself.
:::
:::

## Remittances: India's Hidden BoP Pillar

::: {.columns}
::: {.column width="60%"}
**India is the world's largest recipient of remittances**

- FY2024 inflows: **$120 billion**
- Share of GDP: ~3.4%
- Primary sources: Gulf countries (UAE, Saudi Arabia, Kuwait), USA, UK, Canada

**Key sending states:** Kerala, UP, Bihar, Rajasthan, Punjab
:::
::: {.column width="40%"}
::: {.keybox}
**Remittances vs Trade Deficit**

Trade deficit: −$76B  
Remittances: +$120B  

Remittances alone more than offset the merchandise trade deficit — a critical BoP stabilizer.
:::
:::
:::

## Remittances: Global Comparison

```{r}
#| label: fig-remittances
#| fig-cap: "India: World's Largest Remittance Recipient (USD billion, 2010–2024)\nSource: World Bank, Migration and Remittances Data."
#| fig-width: 8
#| fig-height: 4.5

remit_data <- data.frame(
  year        = c(2010, 2012, 2014, 2016, 2018, 2019, 2020, 2021, 2022, 2023, 2024),
  india       = c(  53,   69,   70,   63,   79,   83,   83,   89,  100,  112,  120),
  china       = c(  53,   58,   64,   61,   67,   70,   60,   53,   51,   50,   48),
  mexico      = c(  22,   23,   24,   27,   36,   39,   41,   54,   61,   67,   66),
  philippines = c(  21,   24,   28,   30,   33,   35,   34,   37,   38,   40,   40)
)

remit_long <- tidyr::pivot_longer(remit_data, -year,
                                  names_to  = "country",
                                  values_to = "remit")

remit_long$country <- factor(
  remit_long$country,
  levels = c("india", "china", "mexico", "philippines"),
  labels = c("India", "China", "Mexico", "Philippines")
)

ggplot(remit_long, aes(x = year, y = remit, colour = country, linewidth = country)) +
  geom_line() +
  geom_point(data = subset(remit_long, country == "India"), size = 3) +
  scale_colour_manual(
    values = c("India" = "#012169", "China" = "#B9975B",
               "Mexico" = "grey50", "Philippines" = "grey70")) +
  scale_linewidth_manual(values = c("India" = 2, "China" = 1,
                                    "Mexico" = 1, "Philippines" = 1)) +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  labs(title    = "India: World's Largest Remittance Recipient ($120B in FY2024)",
       subtitle = "Remittances = largest single positive item in India's Current Account",
       x = NULL, y = "USD Billion",
       colour = NULL, linewidth = NULL) +
  theme(legend.position = "bottom")
```

::: {.fragment}
::: {.hi-gold}
**Gulf concentration risk:** ~55% of India's remittances come from Gulf Cooperation Council (GCC) countries. Oil price slumps → Gulf construction slowdowns → remittance drops → India's CAD widens. The 1991 crisis featured exactly this channel.
:::
:::

## Foreign Direct Investment (FDI) Flows

**FDI inflows FY2024: ~$71 billion (gross)**

::: {.columns}
::: {.column width="55%"}
**Top recipient sectors:**

- IT & software: ~$18B
- Financial services: ~$9B
- Pharmaceuticals: ~$7B
- Telecom: ~$6B
- **Agriculture & food processing: ~$3.6B**

:::
::: {.column width="45%"}
::: {.keybox}
**Agri FDI: $3.6B and growing**

Focus areas:
- Cold chain infrastructure
- Food processing (100% FDI allowed)
- AgriTech startups
- Contract farming/supply chains
:::
:::
:::

::: {.fragment}
::: {.hi}
FDI in agri-food is growing but remains a small fraction — India's farm sector still heavily domestic-capital-driven.
:::
:::

## Forex Reserves: India's Financial Buffer

::: {.methodbox}
**India's Foreign Exchange Reserves — FY2024: $645 billion**
:::

**Components of forex reserves:**

| Component | USD B |
|---|---|
| Foreign currency assets (FCA) | $571 |
| Gold | $53 |
| SDR holdings (IMF) | $18 |
| Reserve tranche position (IMF) | $3 |

::: {.fragment}
::: {.highlightbox}
**Import cover: ~11 months** — well above the conventional safe threshold of 3 months. This gives India significant resilience against external shocks.
:::
:::

## India's 1991 Balance of Payments Crisis

::: {.columns}
::: {.column width="60%"}
**The Crisis:**

- By early 1991, reserves fell to **$1 billion** (< 2 weeks of imports)
- Iraq-Kuwait war raised oil import bill
- Gulf remittances fell sharply
- Fiscal deficit had grown unsustainably
- Credit rating downgrades; capital flight

**India's Response:**

- IMF bailout loan ($2.2B)
- Gold pledged as collateral (47 tonnes)
- Two-step rupee devaluation (25%)
- Structural adjustment → liberalization reforms
:::
::: {.column width="40%"}
::: {.keybox}
**1991 as a turning point**

The crisis forced India to:

- Open up to FDI
- Dismantle import licensing
- Devalue rupee
- Phase out industrial licensing

Agricultural trade also liberalized — QRs on agri imports phased out by 2001 (WTO bound).
:::
:::
:::

## Current Account Sustainability

::: {.highlightbox}
**Current Account Deficit (CAD) FY2024: approximately −0.7% of GDP**
:::

**India's CAD trend:**

| Year | CAD (% GDP) | Assessment |
|---|---|---|
| FY2012 | −4.8% | Dangerously high |
| FY2019 | −2.1% | Elevated |
| FY2022 | −1.7% | Moderate |
| FY2023 | −2.0% | Moderate |
| **FY2024** | **−0.7%** | **Comfortable** |

::: {.fragment}
::: {.hi}
A CAD of −0.7% GDP is considered sustainable. The "safe zone" is generally under −3% for an economy of India's size and growth rate.
:::
:::

## The Edible Oil Vulnerability: A Policy Case Study

::: {.columns}
::: {.column width="55%"}
**The structural problem:**

India produces ~10 million tonnes of edible oil domestically but consumes ~24 million tonnes.

The **14 million tonne import gap** = $14B foreign exchange outflow annually.

**Source concentration risk:**
- 90% of palm oil from Malaysia & Indonesia
- Subject to export bans, price manipulation, climate shocks
:::
::: {.column width="45%"}
::: {.methodbox}
**Policy responses:**

1. **PM KUSUM** — promote oilseed cultivation
2. **National Mission on Edible Oils (Oilpalm)** — expand domestic oil palm to NE states
3. **Import duty manipulation** — lower duties during inflation
4. **Buffer stock policy** — proposed but not implemented
:::
:::
:::

## Rice Export Restrictions: A Policy Dilemma

::: {.columns}
::: {.column width="60%"}
**August 2023: India bans non-basmati white rice exports**

**Rationale:**
- Domestic rice prices rising
- Food security concerns ahead of state elections
- Erratic monsoon raised supply uncertainty

**Impact:**
- India's rice exports fell sharply (India = 40% of global rice trade)
- Global rice prices surged 30–40%
- Indonesia, Philippines, West Africa faced food price spikes
- India's agri export earnings fell by billions
:::
::: {.column width="40%"}
::: {.keybox}
**The trade-off:**

| Restrict exports | Let exports flow |
|---|---|
| Lower domestic prices | Higher farmer income |
| Better food security | Better forex earnings |
| Harms trading partners | Global price stability |

*Classic food security vs. trade earnings dilemma — central to agri trade policy*
:::
:::
:::

## India's Balance of Payments: Summary Picture

**BoP Current Account FY2024 (approximate):**

| Item | USD B |
|---|---|
| Merchandise exports | +437 |
| Merchandise imports | −677 |
| **Trade balance** | **−240** |
| Services exports | +338 |
| Services imports | −167 |
| **Services balance** | **+171** |
| Primary income (net) | −32 |
| Secondary income (remittances) | +120 |
| **Current Account Balance** | **~−17** |

::: {.fragment}
::: {.hi-gold}
CAD of ~$17B on a GDP of ~$3.4 trillion = about −0.5% to −0.7% GDP. Healthy and sustainable.
:::
:::

## India's Global Trade Rankings

::: {.methodbox}
**India's position in global trade (WTO 2024):**
:::

::: {.columns}
::: {.column width="50%"}
**Merchandise trade:**
- Exports: **17th largest globally**
- Imports: **10th largest globally**
- Share of world trade: ~1.8%

**Services trade:**
- Services exports: **7th globally**
- IT services: **Largest global exporter**
:::
::: {.column width="50%"}
**Agriculture specifically:**
- Rice: **#1 global exporter** (40% share)
- Spices: **Top 3 globally**
- Cotton: **Top 3 globally**
- Seafood: **Top 10 globally**

::: {.hi-gold}
India punches above its weight in agricultural exports relative to its overall merchandise trade share.
:::
:::
:::

## Structural Challenges in India's External Sector {.smaller}

**Key vulnerabilities:**

1. **Energy import dependence** — crude oil $232B; geopolitical risk

2. **Edible oil import dependence** — $14B; food inflation risk

3. **Electronic goods deficit** — $80B; reflects low domestic manufacturing

4. **Gold demand** — $47B; compresses CAD but reflects domestic saving patterns

5. **China trade deficit** — ~$85B; supply chain dependency

:::{.notes}
**Structural strengths:**

1. IT services surplus — $220B
2. Remittances — $120B
3. Pharmaceutical exports — $27B
4. Agricultural surplus — +$15.5B
5. Large, diversified economy — natural hedge
:::

## Agricultural Sector's Role in India's External Sector

::: {.highlightbox}
**Agriculture in India's external sector: 3 key roles**
:::

**1. Forex earner:** $43.7B in exports — supports BoP, provides farmer income linkage to global prices

**2. Forex user:** $28.2B in imports — edible oils and pulses; food security imperative but forex cost

**3. Policy instrument:** Government uses export/import restrictions to manage domestic food prices — with global spillover effects (India = swing supplier in rice, sugar, onions)

::: {.fragment}
::: {.hi}
The net surplus of $15.5B means agriculture is a **net contributor** to India's BoP — a valuable buffer.
:::
:::

## Looking Ahead: India's External Sector Goals {.smaller}

**India's trade policy objectives (2025–2030):**

- Merchandise exports target: **$1 trillion by FY2030**
- Agricultural export target: **$100 billion by 2030**
- Increase oilseed domestic production (reduce $14B import bill)
- Value-added processed food exports
- Agri-GI tags to premium global markets

:::{.notes}
**Structural shifts underway:**

- Electronics exports: $300B (replace imports)
- RCEP exit (2019) — protected domestic agri
- FTA with UAE (2022) — agri market access
- FTA with Australia (2022) — pulses imports
- FTA with UK (pending) — key agri market
- EU FTA (ongoing) — GI products, SPS issues
:::

## Key Takeaways — Lecture 10

::: {.keybox}
**Five things to remember:**
:::

1. **India's merchandise trade deficit is $76B** but is more than offset by services surplus ($171B) and remittances ($120B)

2. **Agriculture contributes a net surplus of $15.5B** — making it a positive contributor to the BoP

3. **Edible oil imports ($14B) are India's single largest agricultural vulnerability** — a structural food security and forex risk

4. **India is the world's largest rice exporter** — and unilateral export restrictions have significant global price consequences

5. **Forex reserves at $645B (11 months import cover)** provide substantial external sector resilience; a far cry from the 1991 crisis

## Next Lecture: Foreign Exchange Rates and Determination

**Lecture 11 (July 7, 2026):** *Foreign Exchange — Rates and Determination*

We will cover:

- What is the exchange rate and why does it matter?
- Nominal vs. real exchange rate
- PPP theory and why the rupee "should" depreciate
- Covered and uncovered interest parity
- How exchange rates affect agricultural exporters and importers

:::{.notes}
- History of India's rupee (1947 → today)

Today's external sector data (the flows — exports, imports, remittances) determine **tomorrow's exchange rate.** Understanding the rate is the next step to understanding how agri trade prices are actually set.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 13-14)
- *International Economics* — Appleyard & Field (Ch. 13-14)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
