---
title: "Introduction to International Trade"
subtitle: "ECO 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "April 25, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    logo: ""
    footer: "ECO 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(scales)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

# Part I: Setting the Stage {background-color="#E8EDF5"}

## Course Overview: ECO 2203

**What this course is about**

- The economics of trade between nations — with a focus on **agricultural commodities**
- Why countries trade, who gains, who loses, and what policy should do about it
- India's position in global agricultural trade and key policy instruments

**Learning objectives**

- Master classical and modern theories of trade
- Analyse India's agricultural trade flows and policy environment
- Evaluate trade policy tools: tariffs, quotas, subsidies, SPS measures
- Apply WTO frameworks to real agricultural trade disputes

::: {.highlightbox}
**Assessment:** Mid-semester exam (30%) · End-semester exam (40%) · Term paper (20%) · Participation (10%)
:::

## Course Roadmap: Lectures 1–9 {.smaller}

::: {.columns}
::: {.column width="50%"}
| Lec | Topic | Date |
|-----|-------|------|
| 1 | Introduction to International Trade | Apr 28 |
| 2 | Features & Internal vs International Trade | May 5 |
| 3 | Theories I — Mercantilism & Absolute Advantage | May 12 |
| 4 | Theories II — Comparative Advantage | May 19 |
| 5 | Theories III — Factor Endowments (H-O Model) | May 26 |
:::
::: {.column width="50%"}
| Lec | Topic | Date |
|-----|-------|------|
| 6 | New Trade Theory & Intra-Industry Trade | Jun 2 |
| 7 | Terms of Trade & Trade Gains | Jun 9 |
| 8 | Trade Policy: Tariffs | Jun 16 |
| 9 | Trade Policy: Non-Tariff Barriers | Jun 23 |
:::
:::

## Course Roadmap: Lectures 10–18 {.smaller}

::: {.columns}
::: {.column width="50%"}
| Lec | Topic | Date |
|-----|-------|------|
| 10 | Balance of Payments & Exchange Rates | Jun 30 |
| 11 | WTO Framework & Agreements | Jul 7 |
| 12 | Agreement on Agriculture (AoA) | Jul 14 |
| 13 | SPS & TBT Measures in Agriculture | Jul 21 |
| 14 | India's Foreign Trade Policy | Jul 28 |
:::
::: {.column width="50%"}
| Lec | Topic | Date |
|-----|-------|------|
| 15 | Regional Trade Agreements — ASEAN, SAFTA | Aug 4 |
| 16 | Agricultural Price Policy & Trade | Aug 11 |
| 17 | Food Security and Trade Nexus | Aug 18 |
| 18 | Contemporary Issues & Revision | Aug 25 |
:::
:::

## What is International Trade?

**Definition**

> International trade is the **cross-border exchange** of goods, services, and capital between residents of different countries.

**Three dimensions of trade**

- **Merchandise trade** — physical goods: rice, wheat, machinery, petroleum
- **Services trade** — IT services, tourism, financial services, education
- **Capital flows** — foreign direct investment, portfolio investment, remittances

**Why it matters for agricultural economists**

- India earns ~₹3.6 lakh crore from agricultural exports annually
- Global price signals directly affect domestic farm prices
- WTO rules constrain domestic agricultural support policies

::: {.keybox}
**Core insight:** Trade is fundamentally about countries doing what they are relatively better at, and exchanging the surplus — creating gains for all participants.
:::

## Why Do Nations Trade?

::: {.columns}
::: {.column width="50%"}
**Supply-side reasons**

- **Resource endowments** — India has fertile Indo-Gangetic plains; Brazil has Amazon basin → different agricultural outputs
- **Technology differences** — Netherlands uses precision horticulture; India relies on labour-intensive methods
- **Specialization gains** — Focusing on fewer products reduces unit cost
- **Economies of scale** — Large production runs lower average cost (e.g., Thai rice mills)
:::
::: {.column width="50%"}
**Demand-side reasons**

- **Product variety** — Consumers want Darjeeling tea *and* Colombian coffee
- **Seasonal complementarity** — Southern hemisphere harvests when Northern hemisphere is off-season
- **Income growth** — Rising incomes raise demand for diverse food products

**Fundamental driver**

- Countries differ in **opportunity costs** of production
- Differences in opportunity costs → gains from specialisation and trade
:::
:::

# Part II: The Scale of Global Trade {background-color="#E8EDF5"}

## Scale of Global Trade

::: {.columns}
::: {.column width="55%"}
**Global merchandise trade (2023)**

- Total value: **~\$25 trillion** (WTO, 2024)
- Services trade: **~\$7 trillion**
- Combined: ~\$32 trillion — roughly 30% of world GDP

**Top three traders (merchandise, 2023)**

| Rank | Country/Bloc | Share |
|------|-------------|-------|
| 1 | China | 14.2% |
| 2 | United States | 11.3% |
| 3 | European Union | 14.8% |
:::
::: {.column width="45%"}
**Long-run growth: merchandise exports**

| Year | Value (USD trillion) |
|------|---------------------|
| 1950 | 0.06 |
| 1970 | 0.31 |
| 1990 | 3.5 |
| 2000 | 6.5 |
| 2010 | 15.3 |
| 2020 | 17.6 |
| 2023 | 25.0 |

*Trade has grown ~400× since 1950, far outpacing GDP growth*
:::
:::

::: {.highlightbox}
**Factoid:** Global merchandise trade contracted only twice since 1950 — during the 2008–09 financial crisis and the 2020 COVID shock.
:::

## India in Global Trade

::: {.columns}
::: {.column width="55%"}
**India's trade position (FY2024)**

| Indicator | Value |
|-----------|-------|
| Merchandise exports | \$437 billion |
| Merchandise imports | \$678 billion |
| Merchandise trade deficit | –\$241 billion |
| Services exports | ~\$340 billion |
| Services imports | ~\$180 billion |
| Services trade surplus | +\$160 billion |
| **Total merchandise trade** | **\$1.18 trillion** |

**Global rank:** [18th]{.hi} in merchandise exports (WTO 2023)
:::
::: {.column width="45%"}
**Key observations**

- India runs a **large merchandise deficit** but a significant **services surplus**
- IT/BPO services dominate services exports (~\$200B)
- Merchandise deficit driven mainly by **crude oil** and **gold** imports
- Agriculture consistently contributes a **trade surplus**
- Remittances (~\$120B in FY2024) are the largest single source of foreign exchange

::: {.keybox}
India's overall current account deficit is modest (~1% of GDP) because the services surplus and remittances partially offset the merchandise deficit.
:::
:::
:::

## India's Trade: The Big Picture

```{r}
#| label: fig-india-trade-trend
#| fig-cap: "India's Merchandise Trade (FY2015–FY2024, USD billion)\nSource: DGCI&S / Ministry of Commerce and Industry, GoI."
#| fig-width: 8
#| fig-height: 4.5
library(ggplot2)
library(dplyr)

trade_data <- data.frame(
  year = 2015:2024,
  exports = c(310, 262, 276, 303, 330, 314, 422, 453, 778, 778),
  imports = c(448, 381, 384, 465, 507, 478, 573, 715, 854, 872)
)
trade_long <- tidyr::pivot_longer(trade_data, cols=c(exports, imports), names_to="type", values_to="value")
trade_long$type <- factor(trade_long$type, levels=c("exports","imports"), 
                           labels=c("Exports","Imports"))

ggplot(trade_long, aes(x=year, y=value, color=type, group=type)) +
  geom_line(linewidth=1.5) +
  geom_point(size=3) +
  geom_ribbon(data=trade_data, aes(x=year, ymin=exports, ymax=imports), 
              inherit.aes=FALSE, fill="grey80", alpha=0.5) +
  scale_color_manual(values=c("Exports"="#012169", "Imports"="#B9975B")) +
  scale_x_continuous(breaks=2015:2024) +
  scale_y_continuous(labels=scales::dollar_format(suffix="B")) +
  labs(title="India's Merchandise Trade Trend",
       subtitle="Trade deficit widened post-2020; driven by oil and electronics imports",
       x=NULL, y="USD Billion", color=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom", axis.text.x=element_text(angle=45, hjust=1))
```

::: {.keybox}
**Key insight:** India's trade deficit widened sharply post-2020. Exports plateaued at ~\$440–450B while imports surged to ~\$870B — largely driven by oil, gold, and electronics.
:::

## India's Agricultural Trade: FY2024

::: {.columns}
::: {.column width="55%"}
**Top agricultural exports**

| Commodity | Value (USD bn) |
|-----------|---------------|
| Rice (Basmati + Non-Basmati) | 10.4 |
| Marine products | 7.6 |
| Spices | 3.7 |
| Buffalo meat | 3.5 |
| Sugar | 2.8 |
| Cotton | 2.2 |
| Coffee | 1.2 |
| Castor oil | 1.1 |
| **Total agri exports** | **~43.7** |
:::
::: {.column width="45%"}
**Top agricultural imports**

| Commodity | Value (USD bn) |
|-----------|---------------|
| Edible oils (palm, soy) | 12–14 |
| Pulses | 3.0 |
| Cashew (raw) | 1.8 |
| Fresh fruits | 1.5 |
| Spices (select) | 0.8 |
| **Total agri imports** | **~28.2** |

**Agricultural trade surplus:** [~\$15.5 billion]{.hi}

*Source: APEDA / DGCI&S, 2024*
:::
:::

## India's Agricultural Trade Profile

```{r}
#| label: fig-agri-exports
#| fig-cap: "India's Agricultural Exports by Commodity, FY2024 (USD billion)\nSource: APEDA Annual Report 2023\u201324."
#| fig-width: 8
#| fig-height: 4.5
library(ggplot2)

agri_data <- data.frame(
  commodity = c("Rice (Basmati)", "Rice (Non-Basmati)", "Marine Products",
                "Spices", "Sugar", "Cotton", "Other Cereals", "Processed Food",
                "Fruits & Veg", "Oil Meals"),
  value = c(5.8, 6.4, 7.6, 3.7, 2.8, 2.1, 1.9, 3.2, 2.0, 1.5)
)
agri_data <- agri_data |> arrange(value) |>
  mutate(commodity=factor(commodity, levels=commodity))

ggplot(agri_data, aes(x=commodity, y=value)) +
  geom_col(fill="#012169", width=0.7) +
  geom_text(aes(label=paste0("$", value, "B")), hjust=-0.1, size=3.2) +
  coord_flip() +
  labs(title="India's Agricultural Exports, FY2024",
       subtitle="Total: ~$43.7 billion | Rice + Marine dominate",
       x=NULL, y="USD Billion") +
  scale_y_continuous(limits=c(0,9)) +
  theme_minimal(base_size=11)
```

## Agriculture's Share in India's Trade

::: {.columns}
::: {.column width="50%"}
**Export side**

- Agriculture: [~11%]{.hi-gold} of total merchandise exports (\$43.7B of \$437B)
- Petroleum products: ~17%
- Engineering goods: ~26%
- Gems & jewellery: ~10%
- Chemicals: ~8%

Agriculture is **India's second-largest net foreign exchange earner** after services

**India's strengths:** rice, spices, marine products, cotton — all labour-intensive, climate-favoured commodities
:::
::: {.column width="50%"}
**Import side**

- Edible oils: India imports **60–65%** of its edible oil requirement
  - Palm oil: Indonesia, Malaysia
  - Soybean oil: Argentina, Brazil
- Pulses: Canada, Australia, Myanmar
  - ~25% of domestic requirement imported
- Raw cashew: West Africa, Vietnam

::: {.highlightbox}
**Vulnerability:** India's dependence on imported edible oils (~\$13B/yr) exposes food inflation to global price shocks, as seen in 2021–22 when palm oil prices spiked 50%.
:::
:::
:::

## India's Position in Global Agricultural Trade

```{r}
#| label: fig-global-share
#| fig-cap: "Major Agricultural Exporters: Global Market Share (2023)\nSource: WTO, World Agricultural Trade Statistics 2024."
#| fig-width: 7
#| fig-height: 4
library(ggplot2)

global_share <- data.frame(
  country = c("USA", "Netherlands", "Brazil", "Germany", "France", "China", "India", "Argentina"),
  share = c(11.0, 6.5, 6.3, 5.1, 4.8, 4.5, 2.8, 2.5)
)
global_share <- global_share |> arrange(share) |>
  mutate(country=factor(country, levels=country),
         highlight=ifelse(country=="India", "India", "Others"))

ggplot(global_share, aes(x=country, y=share, fill=highlight)) +
  geom_col(width=0.7) +
  scale_fill_manual(values=c("India"="#B9975B", "Others"="#cccccc")) +
  coord_flip() +
  labs(title="Top Agricultural Exporters: Global Market Share (%)",
       x=NULL, y="Share of World Agricultural Exports (%)", fill=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="none") +
  geom_text(aes(label=paste0(share, "%")), hjust=-0.1, size=3.2)
```

::: {.keybox}
India holds **2.8%** of world agricultural exports — 7th globally — but punches below its weight given its agricultural size. Significant scope to expand market share in rice, spices, and processed foods.
:::

# Part III: Key Concepts and Context {background-color="#E8EDF5"}

## Key Concepts We Will Need

::: {.columns}
::: {.column width="50%"}
**Theory concepts**

- **Comparative advantage** — trade based on *relative*, not absolute, efficiency (Ricardo)
- **Terms of trade** — ratio of export prices to import prices; determines how much a country gains from trade
- **Factor endowments** — land, labour, capital abundance shapes what countries export (Heckscher-Ohlin)
- **Intra-industry trade** — two-way trade in similar products (India exports basmati rice, imports parboiled rice)
:::
::: {.column width="50%"}
**Policy concepts**

- **Tariff** — tax on imports; raises domestic price above world price
- **Non-Tariff Barrier (NTB)** — quotas, SPS standards, licensing
- **Balance of Payments (BoP)** — accounting of all international transactions
- **Exchange rate** — price of one currency in terms of another; a stronger rupee makes exports more expensive
- **Trade balance** — exports minus imports; positive = surplus, negative = deficit
:::
:::

::: {.methodbox}
**How these connect:** Comparative advantage explains *what* countries trade; terms of trade determine *how much* they gain; tariffs and NTBs determine *how much* trade actually occurs.
:::

## Why Study International Trade?

**Policy relevance for India**

- **Foreign Trade Policy 2023–28** targets \$2 trillion in merchandise + services exports by 2030
  - From \$777B (FY2023) to \$2T in 7 years — requires ~14.5% annual growth
- **WTO commitments** constrain India's ability to protect domestic agriculture
  - Aggregate Measurement of Support (AMS) limits; tariff bindings
- **Farmer income** — ~50% of India's workforce in agriculture; export markets expand income opportunities
- **Food security** — Import policy on pulses and oilseeds directly affects 1.4 billion people

**Academic relevance**

- Fastest-growing sub-field in development economics
- Links to climate change (Scope 3 emissions in supply chains), gender (women in agri-export supply chains), and technology (digital trade in services)

::: {.keybox}
**Bottom line:** International trade policy is *agricultural policy* in India. They cannot be studied in isolation.
:::

## What Makes Agricultural Trade Special?

::: {.columns}
::: {.column width="50%"}
**Biological & physical features**

- **Perishability** — fresh produce, milk, meat have short shelf-lives → need cold chains, quick clearance
- **Seasonality** — harvests are discrete events; supply is lumpy
- **Bulkiness** — grain, sugar, cotton: high weight-to-value ratio → high freight costs
- **Sanitary risks** — food-borne pathogens cross borders with commodities → SPS measures justified
:::
::: {.column width="50%"}
**Economic & political features**

- **Food security concern** — governments intervene to ensure domestic supply; export bans, import duty waivers are common
- **Price volatility** — weather, pests, and speculation create large price swings; farmers and consumers are both affected
- **Political economy** — large farm lobbies in US, EU, India influence trade policy disproportionately
- **WTO special treatment** — Agreement on Agriculture has special provisions for developing countries (Special and Differential Treatment)
:::
:::

## India's Major Trading Partners (FY2024)

::: {.columns}
::: {.column width="55%"}
**Top 5 merchandise export destinations**

| Rank | Country | Share of exports |
|------|---------|-----------------|
| 1 | United States | ~18% |
| 2 | UAE | ~6.5% |
| 3 | Netherlands | ~4.2% |
| 4 | China | ~3.8% |
| 5 | Bangladesh | ~3.5% |

**Agricultural export destinations (key)**

- Rice: Bangladesh, Benin, Ivory Coast, Nepal
- Spices: USA, China, Vietnam
- Marine: USA, EU, Japan, China
:::
::: {.column width="45%"}
**Top 5 merchandise import sources**

| Rank | Country | Share of imports |
|------|---------|-----------------|
| 1 | China | ~15% |
| 2 | UAE | ~7.5% |
| 3 | USA | ~6.2% |
| 4 | Russia | ~5.8% |
| 5 | Saudi Arabia | ~5.4% |

::: {.methodbox}
**Gravity model logic:** India trades most with large (USA, China) or geographically close (UAE, Bangladesh) economies — a pattern we will formalise in Lecture 2.
:::
:::
:::

## Composition of India's Merchandise Exports

::: {.columns}
::: {.column width="55%"}
**Major export categories (FY2024, approx.)**

| Category | Value (USD bn) | Share |
|----------|---------------|-------|
| Engineering goods | 112 | 25.6% |
| Petroleum products | 78 | 17.8% |
| Gems & jewellery | 44 | 10.1% |
| Chemicals | 36 | 8.2% |
| **Agriculture & allied** | **44** | **10.1%** |
| Textiles & apparel | 35 | 8.0% |
| Electronic goods | 29 | 6.6% |
| Other | 59 | 13.5% |
:::
::: {.column width="45%"}
**Key observations**

- Engineering goods = India's largest export category; grew from 12% in 2000 to 26% in 2024
- Agriculture has held a **stable ~10–12%** share for two decades
- Petroleum products is a **re-export** story — India refines imported crude and re-exports
- Electronic goods rising fast (mobile phones, semiconductors policy)
- **Structural transformation:** India's export basket is gradually diversifying away from pure commodities

::: {.highlightbox}
India's export ambition: move up the value chain from raw agri commodities to processed food products
:::
:::
:::

# Part IV: Policy and Frameworks {background-color="#E8EDF5"}

## Foreign Trade Policy 2023-28: Five Pillars

**Foreign Trade Policy (FTP) 2023–28: Key pillars**

1. **Export target:** \$2 trillion merchandise + services exports by 2030
2. **Districts as Export Hubs (DEH):** Identify 1 product per district with export potential; provide targeted support
3. **ECGC coverage:** Expanded Export Credit Guarantee Corporation cover to de-risk exporters, especially MSMEs
4. **Remission of Duties and Taxes on Exported Products (RoDTEP):** Replace old MEIS; neutralise embedded taxes
5. **E-commerce exports:** Dedicated policy for cross-border e-commerce; target \$200–300B by 2030

## FTP 2023-28: Agriculture-Specific Provisions

**Agricultural trade–specific provisions**

- Agri-specific chapters supporting APEDA-registered exporters
- GI tag promotion (Darjeeling tea, Alphonso mangoes, Basmati rice)
- Simplification of phytosanitary certification processes

::: {.keybox}
FTP 2023–28 is notably continuity-oriented — it builds on previous policy architecture rather than a radical overhaul, reflecting India's cautious approach to trade liberalisation.
:::

## A Preview: The Basic Trade Model

**Production Possibility Frontier (PPF) — conceptual preview**

::: {.columns}
::: {.column width="55%"}
**Without trade (autarky)**

- A country produces *and* consumes on its PPF
- The slope of the PPF at the consumption point = the domestic relative price
- Example: India producing only rice and wheat — the trade-off between them is fixed by domestic technology

**With trade (open economy)**

- World price ratio differs from domestic price ratio
- Country specialises in the good where it has comparative advantage
- Consumption point can now lie **outside** the PPF
- *Both* countries gain — consumption of both goods increases
:::
::: {.column width="45%"}
::: {.methodbox}
**The core result:**

Trade allows a country to **consume beyond its production possibilities** — the equivalent of getting something for nothing, achieved through specialisation and exchange.
:::

**Upcoming lectures will formalise:**

- Why the world price differs from the autarky price (Lec 3–5)
- How much each country gains — terms of trade (Lec 7)
- What tariffs do to this picture (Lec 8)
:::
:::

## Key Definitions

::: {.columns}
::: {.column width="50%"}
**Trade flows**

- **Exports** — goods/services sold to foreign residents; flow of goods out, money in
- **Imports** — goods/services bought from foreign residents; flow of goods in, money out
- **Trade balance** = Exports − Imports
  - Positive → trade surplus
  - Negative → trade deficit
- **Visible trade** — physical merchandise (rice, cars, oil)
- **Invisible trade** — services, remittances, investment income
:::
::: {.column width="50%"}
**Balance of Payments**

- **Current account** — trade in goods + services + remittances + investment income
- **Capital account** — financial flows: FDI, portfolio investment, loans
- **BoP identity:** Current account + Capital account = 0 (approximately)

**Exchange rate**

- **Nominal exchange rate** — ₹/USD; e.g., ₹83 = \$1
- **Real exchange rate** — adjusts for price level differences; determines *actual* competitiveness
- **Appreciation** → exports become more expensive → reduces export competitiveness
:::
:::

## Summary: Why International Trade Matters

::: {.keybox}
**Four reasons international trade is essential for India**

1. **Growth engine:** Trade has been central to every development success story — Japan, South Korea, China, and now India's IT sector

2. **Consumer welfare:** Trade gives 1.4 billion Indians access to goods they cannot produce cheaply at home — edible oils, electronics, fertilisers

3. **Farmer incomes:** Export markets for rice, spices, and marine products support millions of rural households; global price transmission improves farm income signals

4. **Policy discipline:** WTO commitments and trade agreements create external pressure for domestic policy reform — reducing market distortions that hurt productivity
:::

::: {.highlightbox}
**India's challenge:** Integrating deeply into global value chains while protecting food security and farm incomes — the central tension this course will explore.
:::

## Key Takeaways: Lecture 1

::: {.methodbox}
**What we covered today**

- International trade = cross-border exchange of goods, services, and capital
- Global merchandise trade: **~\$25 trillion (2023)**; services: **~\$7 trillion**
- India ranks **18th** in merchandise exports; total trade ~\$1.18 trillion (FY2024)
- Agricultural exports: **\$43.7 billion** (FY2024) — rice, marine, spices, cotton dominate
- India has a **structural agricultural trade surplus** (~\$15.5B) that helps offset the merchandise deficit
- Agricultural trade is special: perishability, food security, seasonality, and political economy all matter
- FTP 2023–28 targets \$2 trillion in exports — an ambitious but achievable goal
:::

**Questions to think about before next lecture:**

1. Why does India ban rice exports during domestic inflation? Who benefits, who loses?
2. If India has absolute advantage in producing rice, why does Bangladesh also export rice?

## Next Lecture: Lecture 2

::: {.columns}
::: {.column width="60%"}
**Features of International Trade**
*(May 5, 2026)*

We will cover:

- **Salient features** of international trade that distinguish it from domestic trade
- **Immobility of factors** — why labour and capital don't flow freely across borders
- **Currency and exchange risk** — how rupee movements affect export competitiveness
- **Internal vs international trade** — a systematic comparison
- **Advantages and disadvantages** of trade — with India-specific examples
- **India's edible oil import dependence** — a case study in vulnerability
- **The gravity model** — intuition and India examples
:::
::: {.column width="40%"}
::: {.highlightbox}
**Suggested reading**

- Krugman, Obstfeld & Melitz, *International Economics*, Ch. 1
- APEDA Annual Report 2023–24 (summary statistics)
- FTP 2023–28, Chapter 1 (overview)

**Data to explore**

- DGCI&S trade statistics (commerce.gov.in)
- WTO Statistics Portal
:::
:::
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 1)
- *International Economics* — Appleyard & Field (Ch. 1)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
