---
title: "Lecture 11: Foreign Exchange — Rates and Determination"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "July 4, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
    code-fold: true
    code-summary: "Show R code"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---
```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(scales)
library(patchwork)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## Foreign Exchange: Rates and Determination {background-gradient="linear-gradient(135deg, #1a3c5e 0%, #2d6a9f 100%)"}

::: {style="color: white; font-size: 1.1em; margin-top: 1em;"}
How the rupee's price against the dollar shapes India's agricultural exports, food prices, and farmer incomes — and what economic theory tells us about why exchange rates move.
:::

## What is an Exchange Rate?

**The nominal exchange rate** is the price of one currency in terms of another.

$$e = \text{units of domestic currency per unit of foreign currency}$$

$$e_{INR/USD} = 83.5 \Rightarrow \text{1 USD costs }₹83.5$$

- **Direct quote** (India convention): ₹ per USD — how many rupees to buy 1 dollar
- **Indirect quote**: USD per ₹ — how many dollars 1 rupee buys (= 1/83.5 = $0.012)

::: {.keybox}
**India 2024:** The RBI reference rate hovered near ₹83–84/USD throughout 2024. A software exporter invoicing $1M receives ₹83.5 crore at spot. A food importer buying $1M of edible oil pays ₹83.5 crore. The same rate creates winners and losers simultaneously.
:::

## Spot vs Forward Rates

| Feature | Spot Rate | Forward Rate |
|---------|-----------|--------------|
| Settlement | Immediate (T+2) | Agreed today, settled at future date |
| Certainty | Current market price | Locked-in price for future delivery |
| India example | ₹83.5/$ | 3-month forward: ₹84.2/$ |
| Use | Immediate trade settlement | Hedging future currency exposure |

**Forward premium calculation:**

$$\text{Forward Premium (p.a.)} = \frac{F - E}{E} \times \frac{12}{n} \times 100 = \frac{84.2 - 83.5}{83.5} \times 4 \times 100 = 3.35\%$$

::: {.highlightbox}
**Why forward rates matter for agri exporters:** An Indian basmati exporter signing a 3-month USD contract today faces uncertainty — if INR appreciates by delivery, they receive fewer rupees. By selling USD forward at ₹84.2/$, they lock in their revenue. The forward premium (3.35%) reflects market expectations of INR depreciation.
:::

## Nominal vs Real Exchange Rate

The **nominal** rate tells us the currency price. The **real** rate tells us purchasing power competitiveness.

$$q = e \cdot \frac{P^*}{P}$$

where $q$ = real exchange rate, $e$ = nominal rate (₹/$), $P^*$ = US price level, $P$ = India price level.

**Numerical example — Basmati Rice:**

- India price: ₹5,010/quintal; US equivalent: $60/quintal; $e$ = ₹83.5/$
- $q = 83.5 \times (60/5{,}010) = 83.5 \times 0.01197 = 0.999 \approx 1$
- If India inflation pushes price to ₹5,600 with same nominal $e$: $q = 83.5 \times (60/5{,}600) = 0.894$ → INR **real appreciation** → Indian rice less competitive even though nominal rate unchanged

::: {.methodbox}
**Why the real rate is what matters for trade competitiveness:** India's nominal exchange rate may be stable, but if domestic inflation outpaces foreign inflation, Indian goods become relatively expensive. Only a **real depreciation** — either a nominal depreciation or lower relative inflation — genuinely improves export competitiveness.
:::

**Real Effective Exchange Rate (REER):** Trade-weighted average across all trading partners:

$$\text{REER} = \prod_i \left(e_i \cdot \frac{P_i^*}{P}\right)^{w_i}$$

## Appreciation vs Depreciation {.smaller}

**The key rule:** If $e$ rises (more ₹ per $) → INR **depreciates**. If $e$ falls → INR **appreciates**.

::: {.columns}
::: {.column width="50%"}
### INR Depreciation (e ↑, e.g. ₹75→₹85)

- ✅ **Cotton exporters:** same USD price = more ₹ revenue
- ✅ **Rice exporters (world's No.1):** Indian rice cheaper for foreign buyers
- ✅ **Spice/marine exporters:** competitive advantage
- ❌ **Edible oil importers:** same $ price = more ₹ cost → higher cooking oil prices
:::
::: {.column width="50%"}
### INR Appreciation (e ↓, e.g. ₹85→₹75)

- ✅ **Edible oil importers:** cheaper palm, soybean, sunflower oil
- ✅ **Pulses importers:** cheaper lentils from Canada/Australia
- ✅ **Fertiliser importers:** urea, DAP cheaper in ₹
- ❌ **Rice exporters:** Indian rice more expensive globally → volume loss
:::
:::

:::{.notes}
Additional depreciation effects:

- ❌ **Fertiliser importers:** input costs rise → squeezes farmer margins
- ❌ **Food inflation:** imported food basket becomes expensive

Additional appreciation effects:

- ✅ **Consumers:** lower food inflation
- ❌ **Cotton exporters:** margins squeezed
- ❌ **Spice/marine sector:** competitiveness falls
:::

## Demand for Foreign Exchange

**Who demands USD (buys USD, sells ₹)?**

- 🛢️ **Importers** (crude oil, edible oil, electronics, machinery)
- 💸 **Capital outflows** (Indian firms investing abroad, FPI selling)
- ✈️ **Tourism abroad** (Indians travelling overseas)
- 🏦 **Debt repayments** (external commercial borrowings in USD)
- 📉 **FPI outflows** (foreign portfolio investors exiting India)

**Demand function:** As $e$ rises (INR depreciates), imports become expensive → quantity of USD demanded falls.

$$Q_D = 216 - 2e$$

where $e$ is ₹ per USD and $Q$ is billions of USD. Downward-sloping demand curve.

**Agricultural angle:** India's edible oil imports (~$14B/year) are the largest agricultural component of USD demand — palm oil from Indonesia/Malaysia, soybean oil from Argentina/Brazil, sunflower oil from Ukraine. A weaker rupee makes every litre more expensive.

## Supply of Foreign Exchange

**Who supplies USD (sells USD, buys ₹)?**

- 📦 **Exporters** (IT services, rice, spices, cotton, pharmaceuticals) — earn USD, convert to ₹
- 🏭 **FDI inflows** (foreign firms investing in India)
- 📈 **FPI inflows** (foreign portfolio investors buying Indian stocks/bonds)
- 💰 **Remittances** — largest single source at ~$100B/year (> FDI of $71B in FY2023)
- 🏦 **External borrowings** (ECBs, NRI deposits)

**Supply function:** As $e$ rises (INR weakens), Indian exports are cheaper globally → more exported → more USD earned and supplied.

$$Q_S = -116 + 2e$$

Upward-sloping supply curve. When $e$ is high, each dollar of export revenue translates to more rupees — incentivising more exports.

**Agricultural note:** Rice ($10B+), spices ($4B+), cotton ($2B+), marine products ($8B+) are key agri supply sources of USD. A weak rupee boosts these exports.

## Foreign Exchange Market Equilibrium

```{r}
#| label: fig-forex-eq
#| fig-cap: "Foreign Exchange Market: Equilibrium Exchange Rate Determination\nSource: Author's illustration."
#| fig-width: 7
#| fig-height: 4

e_seq <- seq(70, 100, by=0.5)
Q_d2 <- 216 - 2*e_seq
Q_s2 <- -116 + 2*e_seq

df_forex <- data.frame(
  e = e_seq,
  Q_d = pmax(0, Q_d2),
  Q_s = pmax(0, Q_s2)
)

df_d_shift <- data.frame(
  Q = pmax(0, 226 - 2*e_seq),
  e = e_seq
)

ggplot(df_forex) +
  geom_line(aes(x=Q_d, y=e, color="Demand for USD"), linewidth=1.5) +
  geom_line(aes(x=Q_s, y=e, color="Supply of USD"), linewidth=1.5) +
  geom_point(aes(x=50, y=83), size=4, color="black") +
  annotate("text", x=52, y=83.8, label="E* (Rs83/$, $50B)", size=3.5, hjust=0) +
  geom_hline(yintercept=83, linetype="dotted", color="grey50") +
  geom_vline(xintercept=50, linetype="dotted", color="grey50") +
  geom_line(data=df_d_shift, aes(x=Q, y=e), color="#012169", linetype="dashed", linewidth=1.1) +
  annotate("text", x=60, y=96.5, label="D' (import surge\nshifts D right)", size=3, color="#012169") +
  annotate("segment", x=55, y=86, xend=61, yend=89,
           arrow=arrow(length=unit(0.25,"cm")), color="#012169") +
  annotate("text", x=54, y=84.5, label="INR\ndepreciates", size=3, color="#012169") +
  scale_color_manual(values=c("Demand for USD"="#012169", "Supply of USD"="#B9975B")) +
  labs(title="Foreign Exchange Market Equilibrium",
       subtitle="Equilibrium: e* = Rs83/USD, Q* = $50B. Import surge -> demand shifts right -> INR depreciates.",
       x="Quantity of USD (billions)", y="Exchange Rate e (Rs per USD)",
       color=NULL) +
  theme(legend.position="bottom")
```

**Solving for equilibrium:** Set $Q_D = Q_S$:

$$216 - 2e = -116 + 2e \Rightarrow 332 = 4e \Rightarrow e^* = 83, \quad Q^* = 50$$

## Factors Shifting Demand and Supply

| Factor | Shift | Effect on e (₹/$) | INR |
|--------|-------|-------------------|-----|
| India inflation ↑ | D shifts right | e ↑ | Depreciates |
| India growth ↑ → imports ↑ | D shifts right | e ↑ | Depreciates |
| FDI/FPI outflows ↑ | D shifts right | e ↑ | Depreciates |
| Oil prices ↑ | D shifts right | e ↑ | Depreciates |
| Export earnings ↑ | S shifts right | e ↓ | Appreciates |
| FDI/FPI inflows ↑ | S shifts right | e ↓ | Appreciates |
| Remittances ↑ | S shifts right | e ↓ | Appreciates |
| RBI raises rates | S shifts right (capital inflows) | e ↓ | Appreciates |

:::{.notes}
**Reading the table:** Every factor works through either the **current account** (trade in goods/services) or the **capital/financial account** (investment flows). The 2013 "Taper Tantrum" combined FPI outflows (D↑) with reduced remittances to trigger the sharpest post-2008 INR depreciation.
:::

## India INR/USD Trend (2008–2024)

```{r}
#| label: fig-inr-trend
#| fig-cap: "INR/USD Exchange Rate: Key Depreciation Episodes (2008–2024)\nSource: RBI, Database on Indian Economy (DBIE)."
#| fig-width: 9
#| fig-height: 4.5

inr_events <- data.frame(
  year = c(2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,
           2018,2019,2020,2021,2022,2023,2024),
  rate = c(43.5,48.4,45.7,46.7,53.4,60.1,61.0,65.5,67.1,65.1,
           69.9,70.4,74.1,73.9,79.8,82.6,83.5)
)

events <- data.frame(
  year  = c(2008,  2013,       2016,                2018,               2022),
  rate  = c(43.5,  60.1,       67.1,                70.4,               79.8),
  label = c("GFC\n2008","Taper\nTantrum","Demonetisation\n+Trump","Oil crisis\n+capital flight","Ukraine war\n+USD surge")
)

ggplot(inr_events, aes(x=year, y=rate)) +
  geom_line(color="#012169", linewidth=1.8) +
  geom_point(color="#012169", size=2) +
  geom_point(data=events, aes(x=year, y=rate), color="#B9975B", size=4) +
  geom_text(data=events, aes(x=year, y=rate+2.5, label=label),
            size=2.8, color="#B9975B", vjust=0, lineheight=0.9) +
  scale_x_continuous(breaks=seq(2008,2024,2)) +
  labs(title="India's INR/USD Exchange Rate (2008–2024)",
       subtitle="INR depreciated from Rs43 to Rs83/USD over 16 years — inflation differentials + global shocks",
       x=NULL, y="Rs per USD (higher = weaker INR)") +
  theme_minimal(base_size=11)
```

- **2008 GFC:** Global risk-off → capital flight from EMs. Rice/agri export demand fell; edible oil imports also dropped.
- **2013 Taper Tantrum:** Fed signalled rate hike → FPI outflows. Cotton, rice exporters saw ₹ revenue spike but volatile.
- **2018:** Iran sanctions + crude oil surge. India's $150B crude import bill inflated → INR hit ₹75.
- **2022 Ukraine:** Sunflower oil import disruption + USD surge → edible oil crisis in India.

## Purchasing Power Parity: Absolute PPP

**Foundation — Law of One Price:** In competitive markets with free trade, identical goods sell at the same price internationally (once converted at the exchange rate).

**Absolute PPP:** If a representative basket costs ₹4,175 in India and $50 in the USA:

$$e_{PPP} = \frac{P_{India}}{P_{USA}} = \frac{4175}{50} = ₹83.5 \text{ per USD}$$

::: {.methodbox}
**Absolute PPP says:** the exchange rate should equal the ratio of price levels. If the actual rate differs from PPP, there is a real misalignment.
:::

**Big Mac Index 2024 — Reality Check:**

| Item | India | USA |
|------|-------|-----|
| Big Mac price | ₹259 | $5.69 |
| Implied PPP | ₹259/$5.69 = **₹45.5/$** | — |
| Actual rate | ₹83.5/$ | — |
| Verdict | INR **undervalued by ~45%** vs PPP | — |

This does not mean INR is "wrong" — it reflects real structural differences (Balassa-Samuelson effect, see Slide 13).

## Relative PPP

**Relative PPP:** Exchange rates change in proportion to inflation differentials.

$$\% \Delta e \approx \pi_{India} - \pi_{USA}$$

**Numerical example:**

If India's inflation $\pi_{India} = 5\%$ and USA's $\pi_{USA} = 2\%$:

$$\% \Delta e = 5\% - 2\% = +3\% \Rightarrow \text{INR depreciates 3\% per year}$$

**Long-run empirical test for India:**

- ₹45/USD (2000) → ₹83.5/USD (2024): **85.6% depreciation** over 24 years → ~**3.5% per year**
- India avg CPI inflation (2000–2024): ~5.5%; US avg: ~2.2% → **differential ≈ 3.3%**
- ✓ **Broadly consistent with Relative PPP!**

::: {.keybox}
**Agricultural implication:** Relative PPP predicts that India's ~3–4% annual INR depreciation roughly offsets the ~3% domestic cost inflation facing Indian farmers. This means Indian agri exports maintain their global price competitiveness in the long run — inflation at home is offset by depreciation in the exchange rate. Short-run deviations, however, create real competitive swings.
:::

## India: PPP-Implied vs Actual Rate

```{r}
#| label: fig-ppp-india
#| fig-cap: "India: Actual Exchange Rate vs PPP-Implied Rate (INR/USD, 2000–2024)\nSource: RBI; World Bank, International Comparison Program (ICP)."
#| fig-width: 8
#| fig-height: 4.5

inr_data <- data.frame(
  year = 2000:2024,
  actual = c(44.9,47.2,48.6,46.6,45.3,44.1,45.3,41.3,43.5,48.4,
             45.7,46.7,53.4,60.1,61.0,65.5,67.1,65.1,69.9,70.4,
             74.1,73.9,76.2,82.6,83.5),
  ppp_implied = c(16,17,18,19,21,22,24,25,27,30,
                  32,34,37,39,42,44,46,48,51,53,
                  55,57,59,62,64)
)

ggplot(inr_data, aes(x=year)) +
  geom_line(aes(y=actual, color="Actual INR/USD"), linewidth=1.5) +
  geom_line(aes(y=ppp_implied, color="PPP-Implied Rate"), linewidth=1.5, linetype="dashed") +
  geom_ribbon(aes(ymin=ppp_implied, ymax=actual), fill="grey70", alpha=0.3) +
  annotate("text", x=2010, y=38, label="Gap = INR structurally\nundervalued vs PPP\n(Balassa-Samuelson effect)",
           size=3, color="grey40") +
  scale_color_manual(values=c("Actual INR/USD"="#012169","PPP-Implied Rate"="#B9975B")) +
  scale_x_continuous(breaks=seq(2000,2024,4)) +
  labs(title="India: INR/USD — Actual vs PPP-Implied (2000–2024)",
       subtitle="Actual rate persistently above PPP-implied: INR appears undervalued by PPP measure",
       x=NULL, y="Rs per USD", color=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```

**Balassa-Samuelson Effect:** Fast-growing economies (like India) have high productivity growth in tradables (IT, manufacturing) but not in non-tradables (haircuts, local food). This raises wages overall, making non-tradables expensive. PPP calculations using full CPI baskets therefore understate the "true" exchange rate for a developing economy — the gap between actual and PPP-implied is structurally expected, not a sign of manipulation.

## Limitations of PPP

PPP is a powerful long-run benchmark but has important practical limitations:

1. **Non-tradable goods** — haircuts, education, land, local restaurants don't arbitrage across borders. Their prices differ permanently.
2. **Quality differences** — a "Big Mac" in Tokyo ≠ a Big Mac in Delhi in quality perception.
3. **Trade barriers** — tariffs, quotas, and transport costs prevent price equalisation even for tradables.
4. **Balassa-Samuelson effect** — richer/faster-growing countries systematically have currencies above PPP because productivity gains in tradables spill over to raise wages in non-tradables.
5. **Short-run irrelevance** — capital flows, sentiment, and central bank intervention dominate short-run exchange rates. PPP explains long-run trends, not day-to-day movements.

::: {.highlightbox}
**Agricultural implication:** Agricultural commodities — rice, wheat, cotton, edible oils, spices — ARE tradable goods. PPP-based analysis is therefore more applicable to food prices than to services prices. When the IMF/World Bank assess food affordability across countries, they use PPP-adjusted incomes. India's apparent INR undervaluation by PPP helps explain why Indian agricultural exports are consistently price-competitive globally.
:::

## Interest Rate Parity: Uncovered IRP

**Setup:** An Indian investor has ₹1 crore to invest for 1 year. Two options:

1. **INR bond** earning $i_{India} = 7\%$ → get ₹1.07 crore in 1 year
2. **USD bond** earning $i_{USA} = 5\%$, but must convert ₹ to $ now and $ back to ₹ later

**Uncovered Interest Rate Parity (UIP):** In equilibrium, expected returns must be equal:

$$i_{India} = i_{USA} + \frac{E^e - E}{E}$$

where $E^e$ = expected future exchange rate (uncertain — hence "uncovered").

**Rearranged:**

$$\frac{E^e - E}{E} = i_{India} - i_{USA} = 7\% - 5\% = 2\%$$

→ Market expects INR to **depreciate 2% per year** to equalise returns.

::: {.methodbox}
**Interpretation:** India's higher interest rates do NOT simply attract unlimited capital inflows — because investors price in the expected currency depreciation. The interest rate advantage is exactly offset (in equilibrium) by the expected loss from holding a depreciating currency. This is why India cannot run 8% rates indefinitely without currency pressure.
:::

## UIP Diagram

```{r}
#| label: fig-irp
#| fig-cap: "Uncovered Interest Rate Parity: Expected Return Equalization\nSource: Author's illustration."
#| fig-width: 7
#| fig-height: 4.5

depreciation_pct <- seq(-5, 12, by=0.5)
i_usd <- 5
return_usd <- i_usd + depreciation_pct

df_irp <- data.frame(
  dep = depreciation_pct,
  return_inr = 7,
  return_usd = return_usd
)

ggplot(df_irp, aes(x=dep)) +
  geom_line(aes(y=return_inr, color="Return on INR deposits (7%)"), linewidth=1.5) +
  geom_line(aes(y=return_usd, color="Return on USD deposits (5% + depreciation)"), linewidth=1.5) +
  geom_vline(xintercept=2, linetype="dashed", color="grey50") +
  geom_point(aes(x=2, y=7), size=4, color="black") +
  annotate("text", x=2.4, y=7.3, label="UIP Equilibrium\n(2% expected\ndepreciation)", size=3, hjust=0) +
  annotate("segment", x=6, y=8.5, xend=6, yend=11.0,
           arrow=arrow(length=unit(0.25,"cm")), color="red") +
  annotate("text", x=6.4, y=9.8, label="If dep. > 2%:\nUSD more attractive\n-> capital outflow\n-> INR falls until\nUIP restored",
           size=2.8, color="red", hjust=0) +
  scale_color_manual(values=c("Return on INR deposits (7%)"="#012169",
                               "Return on USD deposits (5% + depreciation)"="#B9975B")) +
  labs(title="Uncovered Interest Rate Parity (UIP)",
       subtitle="Equilibrium: Indian rate advantage offset by expected INR depreciation",
       x="Expected INR Depreciation (% per year)", y="Expected Return (%)",
       color=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```

## Covered vs Uncovered IRP

**Covered Interest Rate Parity (CIP):** Uses the *forward* exchange rate — no uncertainty.

$$i_{India} = i_{USA} + \frac{F - E}{E}$$

where $F$ = forward exchange rate (agreed today, settled at future date — no currency risk).

**Uncovered Interest Rate Parity (UIP):** Uses the *expected future spot* rate — uncertain.

$$i_{India} = i_{USA} + \frac{E^e - E}{E}$$

**Empirical verdict:**

- **CIP** holds almost perfectly in practice — arbitrage desks enforce it millisecond-to-millisecond.
- **UIP** weaker empirically — exchange rate risk premium and investor irrationality cause deviations.

**Numerical CIP check (India, 2024):**

- Spot $E$ = ₹83.5/$; 3-month forward $F$ = ₹84.2/$
- Forward premium = $(84.2-83.5)/83.5 \times 4 \times 100 = 3.35\%$ p.a.
- Interest differential: $i_{India}=6.5\%$, $i_{USA}=5.3\%$ → $\Delta i = 1.2\%$
- Gap (3.35% vs 1.2%) is explained by **country risk premium** on Indian sovereign debt.

::: {.keybox}
**Agri exporter hedging via CIP:** A rice exporter expecting $5M in 6 months sells USD forward at ₹84.2/$. CIP guarantees the forward rate is fairly priced relative to the interest differential — no "free money" from hedging, but the exporter eliminates revenue uncertainty. The forward premium (3.35%) is the cost of certainty.
:::

## RBI Intervention in the Forex Market

**How RBI intervenes:**

- **Preventing excessive depreciation:** RBI *sells* USD from reserves → increases USD supply in market → reduces e (strengthens INR)
- **Preventing excessive appreciation:** RBI *buys* USD → reduces USD supply → increases e (weakens INR)

::: {.columns}
::: {.column width="55%"}
**Sterilisation mechanism:**

When RBI buys USD (injects ₹ into economy):
1. Buy USD → ₹ liquidity increases (inflationary)
2. Simultaneously *sell* government securities (G-secs) → absorbs ₹ back
3. Net effect: money supply *neutral*, but forex reserves increase

**India's forex reserve trajectory:**

| Year | Reserves | Import cover |
|------|----------|-------------|
| 1991 | $1.2B | 2 weeks (**crisis**) |
| 2004 | $113B | ~12 months |
| 2014 | $304B | ~8 months |
| 2024 | ~$645B | ~11 months |
:::
::: {.column width="45%"}
::: {.keybox}
**Sterilisation and agriculture:**

When RBI sterilises USD purchases:
- G-sec yields are kept from falling
- Domestic interest rates remain higher
- This keeps credit costs for agri sector elevated

RBI's twin mandate — currency stability + price stability — creates a constant tension in agri credit markets.

**India is NOT classified** as a currency manipulator by the US Treasury — interventions are two-sided and RBI has intervened in both directions over 2022–2024.
:::
:::
:::

## Determinants Summary Table {.smaller}

| Factor | Channel | Effect on e (₹/$) | INR Direction |
|--------|---------|-------------------|---------------|
| India inflation ↑ | PPP/competitiveness | e ↑ | Depreciates |
| India growth ↑ → imports ↑ | Current account | e ↑ | Depreciates |
| FDI/FPI outflows ↑ | Capital account | e ↑ | Depreciates |
| Oil prices ↑ (India imports 85%) | Import demand | e ↑ | Depreciates |
| US Fed raises rates | Carry trade reversal | e ↑ | Depreciates |
| Remittances ↑ | Capital inflows | e ↓ | Appreciates |
| RBI raises repo rate | Capital inflows | e ↓ | Appreciates |
| Export boom (rice, IT) | Current account | e ↓ | Appreciates |

:::{.notes}
**Agricultural angle:** Oil price channel is critical — India imports 85% of crude oil. Every $10/barrel rise in oil adds ~$15B to the import bill, directly pressuring INR. The 2022 Ukraine war triggered this simultaneously with sunflower oil supply disruption — a double agricultural shock transmitted through the forex market.
:::

## Implications for Agricultural Trade {.smaller}

**Depreciation (e ↑: ₹75→₹85) — Winners (Exporters):**

- 🌾 Rice (~$10B/yr): world's No.1 exporter; ~10.7% cheaper for importers
- 🌶️ Spices (~$4B/yr): highly price-elastic globally
- 🐟 Marine products (~$8B/yr): shrimp, fish more competitive

**Appreciation (e ↓: ₹85→₹75) — Winners (Importers):**

- 🫙 Edible oils ($14B+/yr): cheaper cooking oil for consumers
- 🌱 Pulses ($3B+/yr): cheaper lentils, chickpeas

:::{.notes}
**Depreciation losers (input costs rise):**
- 🧪 Fertiliser imports (urea, DAP) cost more
- 🛢️ Diesel for tractors/pumps more expensive
- 🌿 Cotton (~$2B/yr) exporters gain under depreciation

**Appreciation losers (exporters squeezed):**
- 🧪 Fertilisers: urea, DAP, MOP cheaper in ₹ terms
- Rice, spice, marine, cotton exporters lose margin
:::

## Exchange Rate: The Policy Tension

::: {.keybox}
**RBI's dilemma for agricultural trade:**

- Defending INR → helps consumers (lower food inflation) but hurts agri exporters
- Allowing depreciation → helps exporters but risks food inflation spiral

India's response: simultaneously subsidise fertilisers (mitigate input cost channel) **and** impose rice export bans (protect domestic supply) — policies that interact with the exchange rate in complex ways.
:::

## Key Takeaways

::: {.keybox}
**Six core concepts from Lecture 11:**
:::

1. **Exchange rate notation:** $e = ₹/USD$; higher $e$ = weaker INR (depreciation); lower $e$ = stronger INR (appreciation)

2. **Market equilibrium:** $Q_D(e) = Q_S(e)$ determines $e^*$; shifts from trade and capital flows move equilibrium. Solving: $216-2e = -116+2e \Rightarrow e^* = 83$

3. **Real exchange rate** $q = e \cdot P^*/P$: only real depreciation — not just nominal — genuinely improves export competitiveness. REER is the trade-weighted measure RBI monitors.

4. **Absolute PPP:** $e_{PPP} = P/P^*$; India's actual rate >> PPP (Balassa-Samuelson); **Relative PPP:** $\Delta e \approx \pi - \pi^*$ — ~3–4% annual INR depreciation predicted and observed over 2000–2024.

5. **Interest Rate Parity (UIP):** $i_{India} - i_{USA} = (E^e - E)/E$; India's higher rates are offset by expected depreciation — no free lunch from the interest differential.

6. **Agriculture:** depreciation boosts rice/cotton/spice exports; raises edible oil/fertiliser import costs — a constant policy tension that explains India's oscillation between export bans and import duty waivers.

## Next Lecture: Exchange Rate Systems and Open Economy Macroeconomics

::: {.columns}
::: {.column width="60%"}
**Lecture 12 (July 11, 2026):**

*Forex Market — Exchange Rate Systems and Open Economy Macro*

We will cover:

- Fixed vs floating vs managed exchange rate systems
- The Impossible Trinity (Mundell trilemma)
- Mundell-Fleming model: IS-LM-BP
  - Fiscal and monetary policy effectiveness under different regimes
- RBI's managed float in practice
- India's forex reserves: from crisis to abundance
- Forward market and currency hedging
- Exchange rate pass-through to agricultural prices
:::
::: {.column width="40%"}
::: {.methodbox}
**Bridge from today:**

Today we learned *how* the rupee rate is determined by supply/demand, PPP, and interest rate parity.

Next lecture we study:

- What *system* India operates under
- How government **policy** affects the exchange rate
- How open-economy **macroeconomics** links monetary policy, fiscal policy, and the exchange rate

*Reading: Salvatore Ch 14-15; Appleyard Ch 20-21*
:::
:::
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 15-16)
- *International Economics* — Appleyard & Field (Ch. 15-16)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
