---
title: "Lecture 5: Modern Theories — Heckscher-Ohlin & Lewis"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "May 23, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(patchwork)
library(dplyr)
library(scales)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## Recap: The Missing Piece in Ricardo

**Last lecture:** Comparative advantage — trade is driven by *relative* opportunity costs.

**What Ricardo established:** India has lower opportunity cost in Rice → exports Rice; World has lower opportunity cost in Wheat → exports Wheat; both gain from specialisation and trade.

**What Ricardo left unanswered:**

> *Why* do opportunity costs differ across countries in the first place?

::: {.methodbox}
**The Heckscher-Ohlin answer:** Opportunity costs differ because countries differ in their **endowments** of factors of production (labour $L$, capital $K$, land $T$). And goods differ in **factor intensity** — how much of each factor they require.
:::

## The Heckscher-Ohlin Framework {.smaller}

The H-O model is the workhorse of international trade theory. It explains *why* comparative advantages arise from factor abundance.

**Setup:** Two countries (India $I$, USA $U$), two goods (Rice $R$, Machines $M$), two factors (Labour $L$, Capital $K$); identical technologies across countries.

**Factor intensity:** Rice is **labour-intensive** relative to Machines if:
$$\left(\frac{K}{L}\right)_R < \left(\frac{K}{L}\right)_M$$

**Factor abundance (physical definition):** India is **labour-abundant** if:
$$\left(\frac{K}{L}\right)^I < \left(\frac{K}{L}\right)^U$$

**Factor abundance (price definition):**
$$\left(\frac{w}{r}\right)^I < \left(\frac{w}{r}\right)^U \qquad \text{(labour is relatively cheaper in India)}$$

## Assumptions of the H-O Model {.smaller}

| Assumption | Role in the model |
|---|---|
| 2 countries, 2 goods, 2 factors (2×2×2) | Tractable analytical framework |
| **Identical technologies** across countries | Comparative advantage comes *only* from endowments |
| **Different factor endowments** | This is the sole source of trade |
| Constant returns to scale | Concave PPF; no scale effects |
| Perfect competition | Prices equal marginal costs; zero profits |
| Factors mobile *within* countries | Labour/capital shift freely between sectors |
| Factors immobile *across* countries | No international migration or capital flows |
| No transport costs or trade barriers | Goods prices equalise across countries |
| Incomplete specialisation | Both goods produced in both countries |
| Identical, homothetic preferences | Demand differences cannot explain trade |

::: {.methodbox}
**Relaxing assumptions:** Drop identical technologies → Ricardian effects re-enter. Drop immobility → factor flows substitute for goods trade (FPE weakens). Drop 2×2×2 → Heckscher-Ohlin-Vanek factor content theorem.
:::

## H-O Theorem: Formal Statement {.smaller}

**Theorem (Heckscher 1919; Ohlin 1933):**

> A country exports the good that uses its *abundant* factor intensively.

**Applied to India and USA:**

$$\left(\frac{K}{L}\right)^I < \left(\frac{K}{L}\right)^U \quad \text{(India is labour-abundant)}$$

$$\left(\frac{K}{L}\right)_R < \left(\frac{K}{L}\right)_M \quad \text{(Rice is labour-intensive)}$$

$$\therefore \quad \text{India exports Rice; USA exports Machines}$$

**Intuition:** Labour-abundant India faces a low relative wage $\Rightarrow$ labour-intensive Rice is *relatively cheaper* to produce there.

In autarky, India's abundant labour $\Rightarrow$ low $w/r$ $\Rightarrow$ low relative price of Rice:

$$\left(\frac{P_R}{P_M}\right)^I_{\text{autarky}} < \left(\frac{P_R}{P_M}\right)^U_{\text{autarky}}$$

Opening to trade raises $P_R/P_M$ in India — India's Rice sector expands, Machines sector contracts.

## The Edgeworth Box: Factor Allocation

::: {.columns}
::: {.column width="38%"}
The **Edgeworth box** shows all efficient ways to allocate $\bar{L}$ and $\bar{K}$ between two sectors.

- Origin $O_R$: Rice sector
- Origin $O_M$: Machines sector (rotated 180°)
- **Contract curve:** locus of tangencies of Rice and Machine isoquants
  $$\frac{MPL_R}{MPK_R} = \frac{MPL_M}{MPK_M}$$
- All efficient allocations lie on the contract curve
- The contract curve bows toward the labour axis (Rice is labour-intensive)
:::
::: {.column width="62%"}
```{r}
#| label: fig-edgeworth
#| fig-cap: "Edgeworth Box: Efficient Factor Allocation Between Rice and Machines\nSource: Author's illustration."
#| fig-width: 7
#| fig-height: 5
library(ggplot2)

# Box: total Labour = 100, total Capital = 60
# Rice:     Q_R = L_R^0.6 * K_R^0.4  (labour-intensive)
# Machines: Q_M = L_M^0.4 * K_M^0.6  (capital-intensive)
# Contract curve from MRTS_R = MRTS_M:
#   (0.6/0.4)*(K_R/L_R) = (0.4/0.6)*(K_M/L_M)
#   =>  K_R = 240*L_R / (900 - 5*L_R)
L_cc <- seq(1, 99, by = 0.5)
K_cc <- 240 * L_cc / (900 - 5 * L_cc)

# Point E at L_R = 70: visually centred (70% of L-axis, ~51% of K-axis)
L_E <- 70
K_E <- 240 * L_E / (900 - 5 * L_E)   # = 16800/550 ≈ 30.5

# Rice isoquant through E
Q_R_E <- L_E^0.6 * K_E^0.4
iL_R  <- seq(45, 97, by = 0.5)
iK_R  <- (Q_R_E / iL_R^0.6)^(1 / 0.4)
df_R  <- subset(data.frame(L = iL_R, K = iK_R), K >= 0 & K <= 60)

# Machines isoquant through E (drawn in original coords)
# E in machine coords: L_M = 30, K_M ≈ 29.5
Q_M_E <- (100 - L_E)^0.4 * (60 - K_E)^0.6
iLM   <- seq(11, 99, by = 0.5)
iKM   <- (Q_M_E / iLM^0.4)^(1 / 0.6)
df_M  <- subset(data.frame(L = 100 - iLM, K = 60 - iKM),
                K >= 0 & K <= 60 & L >= 0 & L <= 100)

ggplot() +
  annotate("rect", xmin = 0, xmax = 100, ymin = 0, ymax = 60,
           fill = "#f8f8f8", color = "black", linewidth = 1.1) +
  # Isoquants
  geom_line(data = df_R, aes(x = L, y = K),
            color = "#012169", linewidth = 0.9, linetype = "dashed") +
  geom_line(data = df_M, aes(x = L, y = K),
            color = "#B9975B", linewidth = 0.9, linetype = "dashed") +
  # Contract curve
  geom_line(data = data.frame(L = L_cc, K = K_cc), aes(x = L, y = K),
            color = "#012169", linewidth = 1.6) +
  # Efficient allocation E
  geom_point(aes(x = L_E, y = K_E), size = 4, color = "darkgreen") +
  annotate("text", x = 74, y = 33, hjust = 0, size = 3.5,
           label = "E  (efficient alloc.)", color = "darkgreen") +
  # Rice origin
  annotate("text", x = 2, y = 2.5, hjust = 0, size = 4, fontface = "bold",
           label = "O[R]", parse = TRUE, color = "#012169") +
  annotate("text", x = 2, y = 7.5, hjust = 0, size = 3,
           label = "Rice sector", color = "#012169") +
  # Machines origin
  annotate("text", x = 98, y = 57.5, hjust = 1, size = 4, fontface = "bold",
           label = "O[M]", parse = TRUE, color = "#B9975B") +
  annotate("text", x = 98, y = 52.5, hjust = 1, size = 3,
           label = "Machines sector", color = "#B9975B") +
  # Isoquant labels
  annotate("text", x = 82, y = 3, hjust = 0, size = 3,
           label = "Rice isoquant", color = "#012169") +
  annotate("text", x = 5, y = 50, hjust = 0, size = 3,
           label = "Machines\nisoquant", color = "#B9975B") +
  # Contract curve label
  annotate("text", x = 42, y = 26, hjust = 0, size = 3.5, fontface = "bold",
           label = "Contract\ncurve", color = "#012169") +
  labs(
    title    = "Edgeworth box: efficient factor allocations",
    subtitle = expression(paste("At E: ", MPL[R]/MPK[R], " = ", MPL[M]/MPK[M])),
    x        = "Labour allocated to Rice",
    y        = "Capital allocated to Rice"
  ) +
  coord_cartesian(xlim = c(-2, 103), ylim = c(-4, 62)) +
  theme_minimal(base_size = 11)
```
:::
:::

## From Edgeworth Box to PPF

The **Production Possibility Frontier** is derived from the contract curve:

::: {.columns}
::: {.column width="55%"}
**Key insight:**

- Every point on the contract curve corresponds to a point on the PPF
- Moving along the contract curve (reallocating factors) traces out the PPF
- The **slope of the PPF** at any point equals the negative of the relative price $P_R/P_M$

**Factor intensity and PPF shape:**

Because Rice is labour-intensive:
- Moving labour from Machines → Rice: output of Rice rises fast (labour goes to its intensive use)
- PPF is **concave** (bowed outward) — increasing opportunity costs

$$\frac{d^2 R}{d M^2} < 0$$

In the Ricardian model, constant labour requirements → linear PPF. H-O gives a *curved* PPF.
:::
::: {.column width="45%"}
**Illustrative PPFs:**

```{r}
#| fig-width: 5
#| fig-height: 4
library(ggplot2)

M_seq <- seq(0, 100, by=1)
R_HO  <- 200 * sqrt(1 - (M_seq/100)^2)  # concave (H-O)
R_Ric <- 200 - 2*M_seq                    # linear (Ricardo)
R_Ric <- pmax(0, R_Ric)

df_ppf2 <- data.frame(
  M = c(M_seq, M_seq),
  R = c(R_HO, R_Ric),
  model = rep(c("H-O (concave)", "Ricardo (linear)"), each=101)
)

ggplot(df_ppf2, aes(x=M, y=R, color=model, linetype=model)) +
  geom_line(linewidth=1.3) +
  scale_color_manual(values=c("H-O (concave)"="#012169",
                               "Ricardo (linear)"="#B9975B")) +
  labs(title="PPF Shape: H-O vs Ricardo",
       x="Machines", y="Rice", color=NULL, linetype=NULL) +
  theme_minimal(base_size=10) +
  theme(legend.position="bottom")
```
:::
:::

## H-O Welfare: Autarky vs Free Trade {.smaller}

::: {.columns}
::: {.column width="42%"}
**Autarky (E1):** IC1 tangent to the PPF; India produces and consumes at E1.

**Opening to trade:**

- World price $P_R/P_M = 1.5 <$ autarky slope $2$
- Production shifts to P $(M=60,\, R=160)$ — more Rice, fewer Machines
- Trade possibilities line through P, slope $= -1.5$
- India **exports 35 Rice**, **imports 23 Machines**
- Consumption at E2 $(M=83,\, R=125)$ on IC2

**IC2 > IC1 → welfare gain from trade**

The gain comes entirely from **reallocation** — not from more resources, but from trading along comparative advantage.
:::
::: {.column width="58%"}
```{r}
#| label: fig-ho-welfare
#| fig-cap: "H-O Model: Free Trade Raises Welfare Above the Autarky PPF\nSource: Author's illustration."
#| fig-width: 6.5
#| fig-height: 5
library(ggplot2)

# H-O concave PPF: R = 200*sqrt(1-(M/100)^2)
M_seq    <- seq(1, 100, by=0.5)
rice_ppf <- 200 * sqrt(pmax(0, 1 - (M_seq/100)^2))

# Autarky tangency: slope = -2  → M = 100/sqrt(2) ≈ 70.7, R = 200/sqrt(2) ≈ 141.4
E1_M <- 100/sqrt(2);  E1_R <- 200/sqrt(2)
# IC1: U = sqrt(R*M), U1^2 = E1_M * E1_R = 10000
rice_ic1 <- 10000 / M_seq

# Free-trade production: PPF slope = -1.5 → M=60, R=160
P_M <- 60;  P_R_val <- 160
# Trade possibilities line: R = 250 - 1.5*M
M_tr  <- seq(0, 120, by=0.5)
rice_tr <- pmax(0, 250 - 1.5*M_tr)

# Consumption under trade: MRS=R/M=1.5 → R=1.5*M; budget → M=250/3, R=125
E2_M <- 250/3;  E2_R <- 125
# IC2: U2^2 = E2_M * E2_R = (250/3)*125 = 10416.7
rice_ic2 <- 10416.7 / M_seq

df_ppf   <- data.frame(M=M_seq, R=rice_ppf, IC1=rice_ic1, IC2=rice_ic2)
df_trade <- data.frame(M=M_tr,  R=rice_tr)

ggplot() +
  geom_line(data=df_ppf,   aes(x=M, y=R),   color="#012169", linewidth=1.5) +
  geom_line(data=df_trade, aes(x=M, y=R),   color="#B9975B", linewidth=1.3,
            linetype="dashed") +
  geom_line(data=subset(df_ppf, IC1<=245), aes(x=M, y=IC1),
            color="grey50",   linewidth=1, linetype="dotted") +
  geom_line(data=subset(df_ppf, IC2<=245), aes(x=M, y=IC2),
            color="darkgreen", linewidth=1, linetype="dotted") +
  # Points
  geom_point(aes(x=E1_M,   y=E1_R),   color="#012169", size=3.5) +
  geom_point(aes(x=P_M,    y=P_R_val),color="purple",  size=3) +
  geom_point(aes(x=E2_M,   y=E2_R),   color="#B9975B", size=3.5) +
  # Trade triangle
  annotate("segment", x=P_M, y=P_R_val, xend=E2_M, yend=P_R_val,
           color="darkgreen", linewidth=0.8,
           arrow=arrow(length=unit(0.2,"cm"))) +
  annotate("text", x=(P_M+E2_M)/2, y=P_R_val+9,
           label="Import Machines (+23)", size=2.8, hjust=0.5, color="darkgreen") +
  annotate("segment", x=E2_M, y=P_R_val, xend=E2_M, yend=E2_R,
           color="darkgreen", linewidth=0.8,
           arrow=arrow(length=unit(0.2,"cm"))) +
  annotate("text", x=E2_M+3, y=(P_R_val+E2_R)/2,
           label="Export Rice (−35)", size=2.8, hjust=0, color="darkgreen") +
  # Labels
  annotate("text", x=E1_M+2,  y=E1_R-12,   label="E1 (autarky)",    size=3, hjust=0) +
  annotate("text", x=P_M+2,   y=P_R_val+7, label="P (production)",  size=2.8,
           hjust=0, color="purple") +
  annotate("text", x=E2_M+2,  y=E2_R-12,   label="E2 (free trade)", size=3,
           hjust=0, color="#B9975B") +
  annotate("text", x=3, y=210,
           label="IC2 (free trade)", size=2.8, hjust=0, color="darkgreen") +
  annotate("text", x=3, y=175,
           label="IC1 (autarky)",    size=2.8, hjust=0, color="grey50") +
  scale_y_continuous(limits=c(0, 240)) +
  scale_x_continuous(limits=c(0, 115)) +
  labs(title="H-O: free trade moves consumption above the PPF",
       x="Machines", y="Rice") +
  theme_minimal(base_size=11)
```
:::
:::

## Theorem 1: Factor Price Equalisation (FPE)

**Theorem (Samuelson 1948):**

Under H-O assumptions, free trade in goods **completely equalises factor prices** across countries — even without factor mobility:

$$w^I = w^U \qquad \text{(wages equalise)}$$
$$r^I = r^U \qquad \text{(returns to capital equalise)}$$

**Intuition:**

- Trade in goods is an indirect form of trading factors
- India exporting Rice is equivalent to exporting embedded labour
- As India expands Rice production, labour demand rises → Indian wages rise
- As USA imports Rice, labour demand in USA Rice sector falls → US wages fall
- In equilibrium: $w^I = w^U$

::: {.callout-important}
## The FPE insight
Free trade is a **perfect substitute** for factor mobility. This explains why labour migration and goods trade are strategic complements in policy debates.
:::

## FPE: Assumptions and Critique

**FPE requires very strong assumptions:**

| Assumption | Status in reality |
|---|---|
| Identical technologies | Violated (TFP gaps are large) |
| Free trade in all goods | Violated (tariffs, NTBs) |
| No factor intensity reversal | Usually holds |
| Incomplete specialisation | Often violated |
| No transport costs | Violated |

:::{.notes}
**Empirical reality:** Wages in India $\approx$ \$2/hour; USA $\approx$ \$30/hour. FPE clearly does *not* hold in practice.

**Why study FPE then?**

- It reveals the *forces* pushing toward convergence
- Wage convergence between East Asia and OECD is partly consistent with FPE pressure
- The direction of change (not the level) is well-predicted
:::

## Theorem 2: Stolper-Samuelson Theorem

**Theorem (Stolper and Samuelson 1941):**

A rise in the relative price of a good raises the real return to the factor used intensively in that good's production, and lowers the real return to the other factor — **more than proportionally**:

$$\hat{P}_R > 0 \Rightarrow \hat{w} > \hat{P}_R > 0 > \hat{r}$$

where $\hat{x} \equiv d\ln x$ denotes percentage change, and Rice is labour-intensive.

**Formal result:**

$$\hat{w} > \hat{P}_R > \hat{P}_M > \hat{r}$$

This is the **magnification effect**: factor price changes are magnified relative to goods price changes.

**Applied to India opening to trade:**

$$P_R \uparrow \Rightarrow w \uparrow\uparrow, \quad r \downarrow \quad \text{(Indian workers gain; Indian capital owners lose)}$$

$$P_M \uparrow \Rightarrow r \uparrow\uparrow, \quad w \downarrow \quad \text{(US capital owners gain; US workers lose → political economy of protectionism!)}$$

## Stolper-Samuelson: Policy Relevance

**Why does Stolper-Samuelson matter for policy?**

It explains *who lobbies for and against trade liberalisation*:

- Labour in labour-abundant countries → **pro-trade**
- Capital in labour-abundant countries → **anti-trade** (or sector-specific)
- Labour in capital-abundant countries → **anti-trade** (e.g., US manufacturing unions)

**India's agricultural trade policy:** When India raises rice export prices via MSP increases: $P_R \uparrow \Rightarrow w_{\text{agri}} \uparrow \Rightarrow$ rural wages rise — Stolper-Samuelson in action.

::: {.methodbox}
**The compensation principle:** Even when aggregate welfare rises, there are **distributional losers**. A country could use the gains to compensate losers, but in practice: compensation schemes are costly, political constraints prevent redistribution, short-run adjustment costs fall unevenly. This is why **trade adjustment assistance** (TAA) programs exist.
:::

## Theorem 3: Rybczynski Theorem

**Theorem (Rybczynski 1955):**

At constant goods prices, an increase in the endowment of a factor causes:

$$L \uparrow \Rightarrow Q_R \uparrow\uparrow \text{ and } Q_M \downarrow$$

More precisely:

$$\hat{Q}_R > \hat{L} > 0 > \hat{Q}_M$$

- Output of the labour-intensive good rises **more than proportionally** to the labour increase
- Output of the capital-intensive good **falls absolutely**

**Intuition:** Expanding Rice output (to absorb new labour) requires withdrawing capital from Machines, even though capital endowment is unchanged. At constant prices, the only way to re-equate factor ratios is to pull capital into Rice and contract Machines.

**India application:** India's rural-to-urban migration (labour inflow to manufacturing):
$$L_{\text{mfg}} \uparrow \Rightarrow Q_{\text{mfg}} \uparrow\uparrow, \quad Q_{\text{agri}} \downarrow$$
Structural transformation is partly a Rybczynski phenomenon.

## Rybczynski Effect: PPF Diagram

::: {.columns}
::: {.column width="40%"}
An inflow of labour **rotates** the PPF outward, but asymmetrically:

- The Rice (labour-intensive) intercept expands greatly
- The Machines (capital-intensive) intercept contracts

At the **original price ratio**, the economy moves from A to B:

- More Rice produced ($Q_R \uparrow\uparrow$)
- Fewer Machines ($Q_M \downarrow$)

This is the **Rybczynski line** — locus of output combinations at constant prices as labour endowment varies.
:::
::: {.column width="60%"}
```{r}
#| label: fig-rybczynski
#| fig-cap: "Rybczynski Effect: Labour Inflow Shifts PPF\nSource: Author's illustration."
#| fig-width: 7
#| fig-height: 5
library(ggplot2)

M_seq       <- seq(0, 110, by=1)
rice_before <- pmax(0, 200 * sqrt(pmax(0, 1 - (M_seq/100)^2)))
rice_after  <- pmax(0, 260 * sqrt(pmax(0, 1 - (M_seq/88)^2)))

df_ryb <- data.frame(
  M    = c(M_seq, M_seq),
  Rice = c(rice_before, rice_after),
  period = rep(c("Before labour inflow (L=100)",
                 "After labour inflow (L=130)"), each=length(M_seq))
)

ggplot(df_ryb, aes(x=M, y=Rice, color=period, linetype=period)) +
  geom_line(linewidth=1.5) +
  # Price line (slope = constant P_R/P_M)
  geom_abline(slope=-1.8, intercept=162, color="grey60",
              linetype="dotted", linewidth=0.8) +
  # Point A (before)
  geom_point(aes(x=45, y=100), inherit.aes=FALSE,
             color="#012169", size=4) +
  annotate("text", x=48, y=100, label="A (before)",
           hjust=0, size=3.2, color="#012169") +
  # Point B (after) — same price line, more Rice, less M
  geom_point(aes(x=30, y=136), inherit.aes=FALSE,
             color="#B9975B", size=4) +
  annotate("text", x=33, y=136, label="B (after)\nRice↑, Mach↓",
           hjust=0, size=3.2, color="#B9975B") +
  annotate("segment", x=45, y=100, xend=30, yend=136,
           arrow=arrow(length=unit(0.25,"cm")),
           color="darkgreen", linewidth=1) +
  annotate("text", x=22, y=115,
           label="Rybczynski\neffect", color="darkgreen",
           size=3.5, hjust=0.5, fontface="bold") +
  scale_color_manual(values=c(
    "Before labour inflow (L=100)"="#012169",
    "After labour inflow (L=130)" ="#B9975B")) +
  scale_linetype_manual(values=c(
    "Before labour inflow (L=100)"="solid",
    "After labour inflow (L=130)" ="dashed")) +
  labs(title="Labour inflow expands Rice output and contracts Machines",
       subtitle="At constant goods prices — pure Rybczynski movement",
       x="Machines", y="Rice", color=NULL, linetype=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```
:::
:::

## The Leontief Paradox (1953)

**The test:** Wassily Leontief (1953) used US input-output tables to measure the capital and labour content of US exports and imports.

**The prediction (H-O):** The USA, being capital-abundant, should export capital-intensive goods and import labour-intensive goods.

**The finding:** US exports were *more labour-intensive* than US imports — **the opposite of H-O!**

**Formal statement of the paradox:**

$$\left(\frac{K}{L}\right)_{\text{US exports}} < \left(\frac{K}{L}\right)_{\text{US imports}}$$

This contradicts the basic H-O prediction for a capital-abundant country.

## Factor Intensity of India's Trade

```{r}
#| label: fig-leontief
#| fig-cap: "Factor Intensity of India's Exports vs Imports (illustrative)\nSource: Author's illustration based on DGCI&S / World Bank data."
#| fig-width: 8
#| fig-height: 4.5
library(ggplot2)

df_leo <- data.frame(
  category = c("Agricultural\nExports", "Textile\nExports",
                "Pharma\nExports", "Capital Goods\nImports",
                "Electronics\nImports", "Crude Oil\nImports"),
  KL_ratio = c(0.8, 1.2, 2.1, 7.1, 6.3, 8.5),
  type = c("Exports","Exports","Exports","Imports","Imports","Imports")
)
df_leo$category <- factor(df_leo$category,
                           levels=df_leo$category[order(df_leo$KL_ratio)])

ggplot(df_leo, aes(x=category, y=KL_ratio, fill=type)) +
  geom_col(width=0.7) +
  scale_fill_manual(values=c("Exports"="#012169", "Imports"="#B9975B")) +
  coord_flip() +
  labs(title="India: Capital-Labour Ratios in Export vs Import Industries",
       subtitle="India exports labour-intensive goods — consistent with H-O",
       x=NULL, y="Capital-Labour Ratio (relative units)", fill=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```

**India does NOT exhibit a Leontief paradox** — it exports labour-intensive goods, consistent with H-O.

## Explanations for the Leontief Paradox

Several resolutions have been proposed:

1. **Human capital:** US labour is highly skilled — skill-adjusted, US workers embody large amounts of human capital ($\text{Effective } K/L = (K + K_H)/L$)
2. **Factor intensity reversals:** Same good can be labour-intensive in one country and capital-intensive in another (different factor prices → different techniques)
3. **Natural resource abundance:** US imports include oil and raw materials — these are capital-intensive; excluding natural resource industries resolves much of the paradox
4. **Demand bias:** US consumers have stronger preference for capital-intensive goods; demand effects can offset supply-side comparative advantage
5. **Trade barriers:** US tariff structure historically protected labour-intensive industries (agriculture, textiles), distorting trade flows

:::{.notes}
**Modern verdict (Trefler 1995):** Allowing for **technology differences** and **factor productivity** differences across countries reconciles most H-O predictions with data.
:::

## Lewis (1954): Structural Transformation

While H-O focuses on *between-country* differences, the **Lewis model** explains how factor endowments evolve *within* a developing country:

**Two sectors:** Traditional sector (agriculture) with large labour surplus ($MPL_A \approx 0$); Modern sector (industry/services) with higher wages.

**Lewis's key assumption:** $MPL_A < \bar{w}_{\text{subsistence}}$ — a pool of "surplus labour" in agriculture can be absorbed by industry **without raising agricultural wages**.

**Implications for agricultural trade during Lewis transition:**

- Low rural wages → comparative advantage in labour-intensive agricultural exports
- As industrialisation proceeds → rural wages rise → agricultural comparative advantage erodes
- Eventually: Rybczynski-type shift away from agricultural exports

::: {.highlightbox}
**East Asia's trajectory:** Japan (1950s–70s) → South Korea (1970s–90s) → China (2000s–) each followed this path: strong agricultural/labour-intensive exports early, shifting to capital-intensive manufactures.
:::

## Lewis Model: Diagrammatic Analysis

```{r}
#| label: fig-lewis
#| fig-cap: "Lewis Model: Surplus Labour Transfer and Structural Transformation\nSource: Author's illustration after Lewis (1954)."
#| fig-width: 9
#| fig-height: 5
library(ggplot2)
library(patchwork)

# Agriculture: flat MPL (surplus labour) then rising
L_agri <- seq(0, 100, by=1)
MPL_agri <- ifelse(L_agri <= 60, 0.5, 0.5 + 0.08*(L_agri - 60))
subsistence <- rep(1.5, 101)

p_agri <- ggplot(data.frame(L=L_agri, MPL=MPL_agri, sub=subsistence)) +
  geom_line(aes(x=L, y=MPL), color="#012169", linewidth=1.5) +
  geom_line(aes(x=L, y=sub), color="red", linewidth=1, linetype="dashed") +
  geom_vline(xintercept=60, linetype="dotted", color="grey50") +
  annotate("text", x=62, y=1.0,
           label="Turning\npoint", size=3, color="grey40", hjust=0) +
  annotate("text", x=10, y=1.65,
           label="Subsistence wage", color="red", size=3) +
  annotate("rect", xmin=0, xmax=60, ymin=0, ymax=0.5,
           alpha=0.15, fill="#012169") +
  annotate("text", x=30, y=0.25,
           label="Surplus\nlabour", size=3, color="#012169", hjust=0.5) +
  labs(title="Agriculture",
       x="Labour in Agriculture", y="Marginal Product of Labour") +
  theme_minimal(base_size=10) +
  ylim(0, 2.5)

# Industry: upward-sloping demand for labour
L_ind <- seq(0, 80, by=1)
MPL_ind0 <- 4.0 - 0.03*L_ind   # initial
MPL_ind1 <- 5.5 - 0.03*L_ind   # after capital accumulation

p_ind <- ggplot() +
  geom_line(data=data.frame(L=L_ind, MPL=MPL_ind0),
            aes(x=L, y=MPL), color="#B9975B", linewidth=1.5) +
  geom_line(data=data.frame(L=L_ind, MPL=MPL_ind1),
            aes(x=L, y=MPL), color="#B9975B", linewidth=1.5, linetype="dashed") +
  geom_hline(yintercept=1.5, color="red", linetype="dashed", linewidth=1) +
  annotate("segment", x=83, y=2.0, xend=83, yend=3.5,
           arrow=arrow(length=unit(0.2,"cm")),
           color="darkgreen", linewidth=0.8) +
  annotate("text", x=84, y=2.75, label="Capital\naccum.", size=3,
           color="darkgreen", hjust=0) +
  geom_point(aes(x=(4.0-1.5)/0.03, y=1.5), size=3.5, color="#B9975B") +
  geom_point(aes(x=(5.5-1.5)/0.03, y=1.5), size=3.5, color="#B9975B") +
  annotate("text", x=(4.0-1.5)/0.03 + 2, y=1.3,
           label="L1", size=3) +
  annotate("text", x=(5.5-1.5)/0.03 + 2, y=1.3,
           label="L2", size=3) +
  labs(title="Industry",
       x="Labour in Industry", y="Marginal Product of Labour") +
  theme_minimal(base_size=10) +
  ylim(0, 5.5)

p_agri + p_ind
```

## H-O, Lewis, and India's Agricultural Exports {.smaller}

**Connecting the models:**

| Mechanism | H-O | Lewis |
|---|---|---|
| Source of CA | Factor endowments | Wage dualism |
| Key variable | $K/L$ ratio | $MPL_A$ vs $\bar{w}$ |
| Trade prediction | Export labour-intensive goods | Export goods with surplus-labour content |
| Dynamics | Static | Dynamic (transition) |

**Combined implication for India:** India currently sits at an early Lewis stage — large agricultural labour surplus (~40% of workforce); low rural wages → comparative advantage in labour-intensive agri exports.

**India's agricultural export performance (FY2024):** Total \$43.7B; Rice \$10.4B; Marine \$7.6B; Spices \$3.7B — H-O predicts labour-intensive crops dominate (confirmed); Lewis explains low rural wages underpin price competitiveness.

::: {.keybox}
India should invest in agricultural productivity to maintain competitiveness even as wages rise — moving up the value chain rather than defending low-wage advantage.
:::

## The Modern Theory of Trade: Summary {.smaller}

**Three pillars of the H-O framework:**

1. **H-O Theorem:** countries export goods intensive in their abundant factor
2. **Factor Price Equalisation:** $w^I = w^U$, $r^I = r^U$ (free trade substitutes for factor mobility)
3. **Stolper-Samuelson:** $\hat{P}_R > 0 \Rightarrow \hat{w} > \hat{P}_R > 0 > \hat{r}$ (trade has stark distributional effects)
4. **Rybczynski:** $\hat{L} > 0 \Rightarrow \hat{Q}_R > \hat{L} > 0 > \hat{Q}_M$ (factor accumulation biases production)

**Empirical evidence:** Factor content of trade is mixed (Trefler 1995); India's labour-intensive export pattern is confirmed; wage convergence is partial (direction correct); Stolper-Samuelson distributional effects are strong in the long run; Rybczynski structural change is well-supported by East Asian data.

**What H-O cannot explain:** Intra-industry trade; trade between similar countries; scale economies → **New Trade Theory** (Lecture 6)

## Looking Ahead: New Trade Theory

**Why do we need a theory beyond H-O?**

- $\approx 60\%$ of world trade is **intra-industry** (exchanging similar goods)
- Large share of trade is **between similar, high-income countries**
- **Scale economies** matter — firms, not just countries, have comparative advantages

**Krugman's (1979) insight:**

Even with identical factor endowments, countries gain from trade because:

$$\text{Specialisation} \Rightarrow \text{Scale economies} \Rightarrow \text{Lower costs} \Rightarrow \text{More variety}$$

Each country specialises in a *different* variety of the same good — gains arise from **love of variety**, not factor endowments.

**Next lecture:** Economies of scale, monopolistic competition, and the new economic geography — why do trade and production cluster? And what does this mean for India's agricultural processing sector?

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

**Further Reading**

- *International Economics* — Salvatore (Ch. 4-5)
- *International Economics* — Appleyard & Field (Ch. 4-5)
- RBI/DGCI&S/APEDA databases for latest data

**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
