---
title: "Lecture 13: WTO — Origin, Structure and Functions"
shorttitle: "L13 | WTO"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "July 18, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
    code-fold: true
    code-summary: "Show R code"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

## Recap: Lecture 12 {.smaller}

**Lecture 12** examined the **foreign exchange market** — how currencies are priced, how the exchange rate affects export competitiveness, and how India manages the rupee.

::: {.columns}
::: {.column width="55%"}
**Key ideas from Lecture 12:**

- Exchange rate as the price of international trade
- Real exchange rate determines competitiveness
- RBI intervention in the forex market
- Currency depreciation ↑ export attractiveness
:::
::: {.column width="45%"}
::: {.highlightbox}
**The Bridge to Today**

Trade actually happens through the forex system — but **what rules govern the trade itself?**

Who decides that a country cannot ban imports arbitrarily? Who enforces fair competition?

→ Enter the **World Trade Organization**
:::
:::
:::

---

# From GATT to WTO

---

## GATT to WTO: A Brief History {.smaller}

::: {.columns}
::: {.column width="50%"}
**General Agreement on Tariffs and Trade (1947)**

- Signed Oct 1947; entered into force Jan 1948
- **23 original contracting parties** (provisional — never ratified by US Congress)
- Goal: reduce tariffs and quantitative restrictions
- Eight negotiating rounds over 47 years:

| Round | Year | Key Focus |
|-------|------|-----------|
| Geneva | 1947 | Tariff cuts |
| Dillon | 1960–61 | Tariff cuts |
| Kennedy | 1964–67 | Tariff + anti-dumping |
| Tokyo | 1973–79 | NTBs |
| **Uruguay** | **1986–94** | **Agriculture, services, WTO** |
:::
::: {.column width="50%"}
**Why GATT Failed Agriculture**

::: {.methodbox}
**Three critical weaknesses:**

1. **No enforcement** — "diplomatic" dispute resolution, losing party could block adoption of panel report
2. **Agriculture excluded** — US/EU "waivers" exempted farm support from discipline
3. **No services, no IP** — growing share of world trade uncovered
:::

::: {.keybox}
WTO replaced GATT on **January 1, 1995** as a permanent international organisation with legal personality, headquarters in Geneva.
:::
:::
:::

---

## The Uruguay Round (1986–1994) {.smaller}

::: {.highlightbox}
**The most ambitious multilateral trade negotiation in history** — 8 years, 123 countries, launched in Punta del Este, Uruguay
:::

::: {.columns}
::: {.column width="50%"}
**What was new:**

- 🌾 **Agriculture** — AoA for the first time
- 🏢 **Services** — General Agreement on Trade in Services (GATS)
- 💡 **Intellectual Property** — TRIPS Agreement
- 🔨 **Binding dispute settlement** — no more blocking

**Signed at:** Marrakesh, Morocco — **April 15, 1994** (Marrakesh Agreement)

**India's role:** Active participant; founding WTO member
:::
::: {.column width="50%"}
**Uruguay Round Achievements**

- Tariff cuts on industrial goods: ~40%
- Agriculture: brought under multilateral discipline for first time
- Services: GATS framework established
- Intellectual property: TRIPS agreement
- Dispute settlement: binding DSU mechanism
- WTO: permanent institution created
- Single undertaking: all-or-nothing package

::: {.methodbox}
**Single Undertaking Principle:** "Nothing is agreed until everything is agreed" — countries had to accept the entire package, preventing cherry-picking.
:::
:::
:::

---

## Global Trade vs GDP Growth Index {.center}

```{r}
#| label: fig-trade-gdp
#| fig-cap: "Global Trade vs GDP Growth Index (2000-2023)\nSource: WTO, World Trade Statistical Review 2024; World Bank, WDI."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(tidyr)

df <- data.frame(
  year  = c(2000, 2005, 2008, 2010, 2015, 2019, 2020, 2021, 2023),
  Trade = c(55,   75,   85,   87,   100,  108,  95,   115,  118),
  GDP   = c(70,   85,   90,   92,   100,  110,  100,  108,  115)
)

df_long <- pivot_longer(df, cols = c(Trade, GDP),
                        names_to = "Series", values_to = "Index")

ggplot(df_long, aes(x = year, y = Index, color = Series, group = Series)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  annotate("text", x = 2020, y = 88,
           label = "COVID-19\ndrop", color = "#C84B31",
           size = 3.2, fontface = "bold") +
  annotate("segment", x = 2020, xend = 2020, y = 92, yend = 97,
           color = "#C84B31",
           arrow = arrow(length = unit(0.18, "cm"))) +
  scale_color_manual(
    values = c("Trade" = "#012169", "GDP" = "#B9975B"),
    labels = c("Trade Volume Index", "GDP Volume Index")
  ) +
  scale_x_continuous(breaks = c(2000, 2005, 2008, 2010, 2015, 2019, 2020, 2021, 2023)) +
  scale_y_continuous(limits = c(50, 125)) +
  labs(
    title    = "Global Trade vs GDP Growth Index (2000-2023)",
    subtitle = "Index: 2015 = 100",
    x = "Year", y = "Index (2015 = 100)", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## WTO at a Glance {.smaller}

::: {.columns}
::: {.column width="50%"}
**Key Facts (2024)**

| Item | Detail |
|------|--------|
| Headquarters | Geneva, Switzerland |
| Members | **166** (as of 2024) |
| Founded | January 1, 1995 |
| India | **Founding member** |
| Budget | ~CHF 197 million/year |
| Staff | ~700 |
| Official languages | English, French, Spanish |

**Two-thirds of members are developing countries** — WTO is not a "rich country club"
:::
::: {.column width="50%"}
::: {.keybox}
**Director-General:**

**Ngozi Okonjo-Iweala**
(Since February 1, 2021)

- First **woman** DG in WTO history
- First **African** DG in WTO history
- Former Finance Minister of Nigeria
- Former World Bank Managing Director

Re-appointed for second term (2025–2029)
:::

::: {.highlightbox}
India has been a founding and **strategically active** member — one of the most influential developing country voices at the WTO
:::
:::
:::

---

## WTO Membership Growth {.center}

```{r}
#| label: fig-wto-membership
#| fig-cap: "WTO Membership Growth (1995-2024)\nSource: WTO Secretariat."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)

df_mem <- data.frame(
  year    = c(1995, 2000, 2005, 2010, 2015, 2020, 2024),
  members = c(76,   140,  148,  153,  162,  164,  166)
)

ggplot(df_mem, aes(x = year, y = members)) +
  geom_line(color = "#012169", linewidth = 1.3) +
  geom_point(color = "#012169", size = 3.5) +
  geom_text(
    data = subset(df_mem, year == 2024),
    aes(label = "166 members\n(2024)"),
    hjust = -0.1, vjust = 0.5, size = 3.5,
    color = "#C84B31", fontface = "bold"
  ) +
  scale_x_continuous(
    breaks = c(1995, 2000, 2005, 2010, 2015, 2020, 2024),
    limits = c(1994, 2028)
  ) +
  scale_y_continuous(limits = c(60, 185)) +
  labs(
    title = "WTO Membership Growth (1995-2024)",
    x = "Year", y = "Number of Member Countries"
  ) +
  theme_minimal(base_size = 11)
```

---

# WTO Principles

---

## The Five Core Principles of the WTO {.smaller}

::: {.methodbox}
**The WTO rulebook rests on five foundational pillars — all member commitments flow from these**
:::

::: {.columns}
::: {.column width="50%"}
**1. Most Favoured Nation (MFN)**
Non-discrimination among all members: a concession given to one must be given to all.
*Art. I GATT, Art. II GATS*

**2. National Treatment**
Imported goods must be treated no less favourably than domestic goods **once inside the border**.
*Art. III GATT*

**3. Bound Tariffs**
Members commit to ceiling tariff rates (bound rates) in their schedules — cannot raise above without compensation.
:::
::: {.column width="50%"}
**4. Reciprocity**
Concessions are exchanged mutually — no free riding. Basis of negotiating rounds.

**5. Transparency**
Members must notify laws, regulations, and trade measures to the WTO. No hidden barriers.

::: {.highlightbox}
**Why these matter for India:** India's bound tariff schedule gives it enormous policy space (bound 113% average on agriculture vs. applied ~35%). These principles simultaneously protect India from arbitrary foreign barriers.
:::
:::
:::

---

## MFN in Depth: The India Example {.smaller}

::: {.columns}
::: {.column width="55%"}
**How MFN Works**

If India gives the USA a **0% tariff on apples**, under MFN, India must give **the same 0% to ALL WTO members** — Canada, Australia, Chile, New Zealand, China.

You cannot discriminate between WTO members.

**Permitted Exceptions:**

1. **Free Trade Agreements (Art. XXIV GATT):** Deeper preferences within FTA — WTO legal if the FTA covers "substantially all trade"
2. **GSP (Enabling Clause):** Developed countries can give preferential rates to developing countries unilaterally
3. **Security exceptions (Art. XXI):** National security justifications
:::
::: {.column width="45%"}
::: {.keybox}
**India's FTA Example**

India–UAE CEPA (2022):
- India gives 0% tariff on many UAE goods
- WTO members do NOT automatically get this rate
- WTO-legal because CEPA is an FTA under Art. XXIV

But India must ensure CEPA covers substantially all trade — otherwise challenged at WTO
:::

::: {.methodbox}
**India's FTA network** (2024): UAE, Mauritius, Japan, South Korea, ASEAN, Singapore, Sri Lanka, Thailand (partial), Australia, EFTA
:::
:::
:::

---

## WTO Organisational Structure {.smaller}

```
MINISTERIAL CONFERENCE  (Supreme body — every 2 years)
         │
   GENERAL COUNCIL  (Day-to-day — Geneva ambassadors)
    ┌────┴────────────────────────┐
    │                             │
Trade Policy Review Body    Dispute Settlement Body (DSB)
    │
    ├── Council for Trade in GOODS (GATT)
    │       ├── Committee on Agriculture
    │       ├── SPS Committee
    │       ├── TBT Committee
    │       ├── Anti-Dumping Committee
    │       └── Safeguards Committee
    │
    ├── Council for Trade in SERVICES (GATS)
    │
    └── Council for TRIPS
```

::: {.highlightbox}
**India's Geneva Mission** maintains a permanent delegation of ~30 diplomats/trade negotiators to engage all WTO bodies. India chairs or co-chairs several committees.
:::

---

## Ministerial Conferences: Key Milestones {.smaller}

::: {.columns}
::: {.column width="50%"}
| MC | Location | Year | Key Outcome |
|----|----------|------|-------------|
| MC1 | Singapore | 1996 | "Singapore issues": investment, competition, procurement |
| MC3 | Seattle | 1999 | **Failed** — "Battle of Seattle" |
| MC4 | **Doha** | 2001 | Launched Doha Development Agenda |
| MC5 | Cancún | 2003 | **Failed** — G-20 coalition walked out |
| MC6 | Hong Kong | 2005 | Partial agriculture framework |
:::
::: {.column width="50%"}
| MC | Location | Year | Key Outcome |
|----|----------|------|-------------|
| MC9 | **Bali** | 2013 | **TFA + Peace Clause** |
| MC10 | **Nairobi** | 2015 | Export subsidy elimination |
| MC11 | Buenos Aires | 2017 | No consensus |
| MC12 | **Geneva** | 2022 | **Fisheries subsidies** |
| MC13 | Abu Dhabi | 2024 | E-commerce moratorium extended |

::: {.keybox}
**India hosted MC6 in Hong Kong context** and has been central to the G-33 and G-20 agricultural coalitions at every MC
:::
:::
:::

---

## Key WTO Agreements — Overview {.smaller}

::: {.columns}
::: {.column width="50%"}
| Agreement | What it covers |
|-----------|---------------|
| **AoA** | Agriculture: market access, domestic support, export subsidies |
| **SPS** | Food safety, animal & plant health measures |
| **TBT** | Product standards, labelling, testing procedures |
| **Anti-Dumping** | Duties against below-cost imports |
| **SCM** | Subsidies & countervailing duties |
:::
::: {.column width="50%"}
| Agreement | What it covers |
|-----------|---------------|
| **Safeguards** | Emergency import restrictions |
| **GATS** | Trade in services (banking, telecom, education) |
| **TRIPS** | Intellectual property rights |
| **TFA** | Customs procedures, border facilitation |
| **Fisheries** | Subsidies for fishing fleets |

::: {.highlightbox}
**For agricultural trade**, the most important are: AoA, SPS, TBT, Anti-Dumping, and SCM. Lectures 14–16 cover these in depth.
:::
:::
:::

---

## Dispute Settlement Understanding (DSU) {.smaller}

::: {.methodbox}
**WTO's most important institutional innovation** — binding, rules-based dispute resolution replacing GATT's "diplomatic" process
:::

::: {.columns}
::: {.column width="55%"}
**The DSU Process:**

1. **Consultations** (60 days) — bilateral negotiation
2. **Panel established** — 3 trade law experts, 6–9 months
3. **Panel Report** adopted by DSB
4. **Appellate Body** — 3 of 7 members, 60–90 days
5. **DSB adoption** — quasi-automatic (reverse consensus)
6. **Implementation** — reasonable time period
7. **Retaliation** — if not complied, winning party can suspend concessions
:::
::: {.column width="45%"}
::: {.highlightbox}
**Appellate Body Crisis**

Since 2017, the **USA has blocked appointments** to the AB — currently non-functional (needs 3 members, has 0 active).

**MPIA:** Multi-Party Interim Appeal Arbitration Arrangement — 53 members (including India) agreed to use arbitration as workaround.

USA objects: AB "overreached" its mandate by creating precedents.
:::
:::
:::

---

## India in WTO Disputes {.smaller}

::: {.columns}
::: {.column width="50%"}
**As Complainant (India won):**

::: {.keybox}
**DS456 — US-India Solar Cells (2016)**
USA's domestic content requirements for solar panels under the National Solar Mission violated WTO national treatment (Art. III). India won — US had to remove requirements.
:::

**Other active cases India initiated:** challenges to US steel tariffs, EU carbon measures
:::
::: {.column width="50%"}
**As Respondent (India lost/challenging):**

- **DS265** — EC-India Sugar Subsidies: India's sugar export subsidies found WTO-illegal; India appealing
- **DS430** — India-Agricultural Products: US challenged India's avian influenza ban on US poultry — India lost on SPS grounds
- **DS582** — India-IT Tariffs: EU/Japan challenged India's IT product tariffs — India lost
- **DS541** — India-MEIS: US challenged India's export incentive scheme — India lost

::: {.highlightbox}
India has participated in **50+ disputes** as complainant, respondent, or third party — reflecting deep WTO engagement
:::
:::
:::

---

## India's WTO Dispute Cases by Partner {.center}

```{r}
#| label: fig-india-wto-disputes
#| fig-cap: "India's WTO Dispute Settlement Cases by Partner\nSource: WTO Dispute Settlement Gateway (as of 2024)."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)

df_ds <- data.frame(
  country = rep(c("USA", "EU", "Australia", "China", "Brazil", "Pakistan", "Turkey"), 2),
  role    = rep(c("Complainant", "Respondent"), each = 7),
  cases   = c(3, 2, 1, 2, 1, 1, 0,
              6, 4, 2, 0, 1, 0, 1)
)

df_ds$country <- factor(df_ds$country,
  levels = c("Turkey", "Pakistan", "Brazil", "China", "Australia", "EU", "USA"))

ggplot(df_ds, aes(x = cases, y = country, fill = role)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("Complainant" = "#012169", "Respondent" = "#B9975B")) +
  scale_x_continuous(breaks = 0:7) +
  labs(
    title = "India's WTO Dispute Settlement Cases by Partner",
    x = "Number of Cases", y = NULL, fill = "India's Role"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## India's WTO Negotiating Coalitions {.smaller}

::: {.columns}
::: {.column width="55%"}
**India leads or co-leads multiple developing country coalitions:**

::: {.methodbox}
**G-33 (Chair: Indonesia, India active)**
46 developing countries. Demands: Special Products (SP) — exempt sensitive crops from tariff cuts; Special Safeguard Mechanism (SSM) — emergency tariff response to import surges.

**G-20 Agricultural Group (India, Brazil, China, South Africa)**
Demands deep cuts in US/EU trade-distorting domestic support; improved market access; elimination of export subsidies. **Not to be confused with G20 heads of state summit.**

**NAMA-11:** Developing country coalition on industrial tariff negotiations — resist deep cuts.
:::
:::
::: {.column width="45%"}
**Why Coalitions Matter**

WTO operates by **consensus** — no voting for most decisions. Coalitions give developing countries blocking power.

::: {.keybox}
**India's G-33 Asks:**
- 1.5 billion farmers in developing countries need protection from import surges
- Special Products: food security crops (rice, wheat, pulses) protected from tariff cuts
- SSM: right to raise tariffs temporarily if imports surge and hurt local farmers
:::

India has successfully used these coalitions to block unfavourable outcomes at Cancún (2003) and other MCs.
:::
:::

---

## The Doha Development Round {.smaller}

::: {.highlightbox}
**Launched:** November 2001, Doha, Qatar — framed as a "development round" post-9/11 solidarity
:::

::: {.columns}
::: {.column width="55%"}
**The Doha Agenda:**

- **Agriculture:** deep cuts in US/EU domestic support; improved market access; eliminate export subsidies
- **Industrial goods (NAMA):** significant tariff cuts
- **Services (GATS):** further liberalisation
- **TRIPS & Public Health:** access to medicines for poor countries
- **Development:** special and differential treatment (S&DT) strengthened

**Why it stalled — the political economy:**
- USA: Farm Bill politics, Congress won't cut subsidies
- EU: defensive on sugar, dairy, beef
- India/China: demand strong SSM for agriculture; resist industrial tariff cuts (NAMA)
:::
::: {.column width="45%"}
::: {.methodbox}
**July 2008 Mini-Ministerial — The Collapse**

India's then-Commerce Minister **Kamal Nath** walked out over SSM provisions. The US wanted an import surge threshold of 140% of base; India demanded 110%. Talks collapsed.

Symbolically important: India prioritised **farmer protection over deal**.
:::

::: {.keybox}
Doha Round is effectively **dead as a single undertaking**. Some elements (TFA, fisheries subsidies, export subsidy elimination) have been harvested. The agriculture agenda — the core — remains unresolved after 23+ years.
:::
:::
:::

---

## Trade Facilitation Agreement (TFA, 2017) {.smaller}

::: {.columns}
::: {.column width="55%"}
::: {.keybox}
**First multilateral agreement fully concluded at WTO** — in force February 22, 2017
:::

**Key provisions:**
- **Advance rulings** on tariff classification and origin
- **Single window** for customs clearance
- **Expedited release** of perishable goods
- **Risk-based inspection** (not 100% physical checks)
- **Transit facilitation** for landlocked countries
- **Appeal procedures** for customs decisions

**Benefits for India's agri exports:**
Perishables (mangoes, grapes, vegetables) need fast clearance. TFA obliges importing countries to have expedited procedures.
:::
::: {.column width="45%"}
**India's TFA Implementation:**

::: {.methodbox}
- **ICEGATE** single window for customs
- 24/7 operations at major ports
- Risk-based inspection system (SWIFT)
- Authorised Economic Operator (AEO) programme
:::

**Time Release Study (India):**

| Channel | Before | After TFA |
|---------|--------|-----------|
| Air cargo | 120 hrs | **49 hrs** |
| Sea cargo | 180 hrs | **68 hrs** |

::: {.highlightbox}
World Bank estimates: full TFA implementation could reduce trade costs by **14.3%** for developing countries
:::
:::
:::

---

## WTO Fisheries Subsidies Agreement (MC12, 2022) {.smaller}

::: {.columns}
::: {.column width="55%"}
::: {.methodbox}
**First new multilateral WTO agreement on fisheries subsidies**

**Prohibits:**
- Subsidies for Illegal, Unreported, and Unregulated (IUU) fishing
- Subsidies for fishing overfished stocks
- Subsidies for fishing in unregulated high seas

**Context:** ~USD 22 billion/year in global fisheries subsidies; fishing stocks critically depleted — 34% overfished globally
:::

**Developing country flexibilities:**
- Transition period of 2 years for some provisions
- Inshore fishing (within 12 nautical miles) largely exempted
- LDCs: longer transitions
:::
::: {.column width="45%"}
**India's Position and Interests**

::: {.keybox}
India's concerns:
- Protect **small-scale traditional fishermen** (Kerala, Tamil Nadu, Andhra Pradesh coastal communities)
- India's fishing subsidies (fuel subsidy, boat building) support livelihood of ~14 million fishers
- Demand: developed country distant-water fleets (EU, China, Japan) subsidised for decades — they caused the overfishing problem, not Indian small fishermen
:::

India negotiated carve-outs for artisanal fishing within national EEZ (200 nautical miles). Still under negotiation for full implementation.
:::
:::

---

## WTO and India's Food Security Tension {.smaller}

::: {.columns}
::: {.column width="50%"}
**The Policy Conflict**

India's food security system:
1. **MSP** (Minimum Support Price) set above world prices for wheat and rice
2. **FCI** procures at MSP from farmers
3. **PDS** distributes at subsidised rates to poor households

WTO critics argue: purchasing above world prices = **Amber Box support** (trade-distorting domestic support) that must be counted against India's AMS commitment

::: {.methodbox}
India's defence: The **Fixed External Reference Price (FERP)** used to calculate AMS uses 1986–88 world prices — which were artificially depressed by US/EU subsidies at the time. The baseline is unfair.
:::
:::
::: {.column width="50%"}
::: {.keybox}
**The Numbers**

India's rice MSP (FY2024): **₹2,183/quintal**

FERP (1986–88): **USD 116/tonne = ~USD 11.6/quintal**

Current MSP ≈ **USD 26.3/quintal**

AMS = (26.3 − 11.6) × 42 million tonnes procured ≈ **~USD 6.2 billion**

De minimis limit (10% of rice production) ≈ **~USD 2.8 billion**

→ India likely exceeds its de minimis threshold
→ **Peace Clause protection needed**
:::
:::
:::

---

## The Peace Clause (Bali, 2013) {.smaller}

::: {.columns}
::: {.column width="55%"}
**Agreement on Public Stockholding for Food Security**

Bali MC9 (December 2013) — negotiated primarily by **India** under Commerce Minister Anand Sharma

**Protection offered:** WTO members will not formally challenge developing country public stockholding programs even if they technically exceed de minimis AMS limits, **provided:**

1. Program existed before December 2013
2. Regular, detailed notifications to the WTO Committee on Agriculture
3. Stocks not released onto world markets at below-market prices
4. Programme targeted at traditional staple foods and the poor

India notifies rice, wheat, and coarse cereal procurement every year under the Peace Clause.
:::
::: {.column width="45%"}
::: {.methodbox}
**The Catch**

- Agreed as **interim measure** pending permanent solution
- Nairobi MC10 (2015): committed to finding permanent solution by MC11 (2017) — missed
- Permanent solution still **not agreed** as of 2024 despite every MC discussing it
- India's position: permanent solution must include new FERP methodology
:::

::: {.highlightbox}
The Peace Clause is India's single most important WTO negotiating achievement in recent years. Without it, India's PDS + FCI procurement system would be legally vulnerable to challenge.
:::
:::
:::

---

## TRIPS and Indian Agriculture {.smaller}

::: {.columns}
::: {.column width="55%"}
**TRIPS Agreement (1995):** Patents, copyrights, trademarks, geographical indications, and plant variety protection.

**Key tension for agriculture:** TRIPS Art. 27.3(b) requires protection of plant varieties either through patents or a sui generis system.

**India's response — PPVFRA (2001):**
Protection of Plant Varieties and Farmers' Rights Act — India's sui generis system

::: {.keybox}
**Key PPVFRA provisions:**
- **Breeders' rights:** 15–18 years protection for new varieties
- **Farmers' right:** to save, use, sow, re-sow, exchange, and sell farm-saved seed (CANNOT sell under breeder's brand)
- **Researchers' exemption:** use protected variety for research/breeding
- **National Gene Fund:** benefit-sharing from commercialisation
:::
:::
::: {.column width="45%"}
**Monsanto Bt Cotton Case**

::: {.methodbox}
Monsanto held trait patent on Bt cotton gene in India. Charged royalties (trait value) to all seed companies using the Bt technology.

India's seed companies: royalties too high, technology should be accessible.

2021: Supreme Court ruled plant varieties and essentially biological processes are **not patentable** under Indian Patents Act Section 3(j).

Monsanto's patent on Bt gene effectively invalidated in India.
:::

::: {.highlightbox}
India resists **TRIPS-plus** obligations pushed in FTA negotiations (longer patent terms, data exclusivity for agrochem companies) that would restrict farmer seed access
:::
:::
:::

---

## India's WTO Schedules and Policy Space {.smaller}

::: {.columns}
::: {.column width="50%"}
**India's Bound Tariff Schedule — Agriculture**

| Commodity | Bound | Applied (2024) | "Water" |
|-----------|-------|----------------|---------|
| Rice | 80% | 25% | 55% |
| Wheat | 100% | 40% | 60% |
| Maize | 60% | 50% | 10% |
| Sugar | 150% | 100% | 50% |
| Palm oil | 300% | 100% | 200% |
| Milk powder | 60% | 30% | 30% |

**Average bound tariff on agriculture: 113.5%**
**Average applied tariff: ~35%**
:::
::: {.column width="50%"}
::: {.keybox}
**The "Water in the Tariff" = Policy Space**

The gap between bound and applied tariffs means India can **raise applied tariffs without WTO violation**. This is India's key defensive instrument.

Example: When palm oil imports surged hurting domestic oilseed farmers, India raised the applied tariff from 30% to 100% — still below the 300% bound rate. No WTO violation.
:::

::: {.methodbox}
**AMS De Minimis:** 10% of production value (developing country rate — vs. 5% for developed). This is why India gets more space than the US/EU.
:::
:::
:::

---

## Recent WTO Issues Affecting India {.smaller}

::: {.columns}
::: {.column width="50%"}
**E-Commerce Moratorium**

Since 1998: WTO members agreed no customs duties on **electronic transmissions** (digital goods, streaming, software downloads).

Extended at every MC including MC13 (Abu Dhabi 2024).

::: {.methodbox}
**India's position:** End the moratorium. India loses potential tariff revenue on digital goods (music, films, software, data). India and South Africa are the most vocal opponents.

India argues: rich country digital companies benefit most; developing countries lose fiscal revenue.
:::
:::
::: {.column width="50%"}
**Joint Statement Initiatives (JSIs)**

Plurilateral negotiations outside the main WTO consensus framework:
- E-commerce JSI (90 members)
- Investment facilitation JSI
- Domestic regulation (services)

::: {.highlightbox}
**India did NOT join** the e-commerce JSI or investment facilitation JSI.

India's concern: JSI outcomes should not become part of WTO multilateral rules without consensus of all 166 members.

Reflects India's **defensive posture** on digital trade — India wants to preserve policy space to regulate data and digital economy.
:::
:::
:::

---

## Key Takeaways: WTO {.smaller}

::: {.keybox}
**Six Things to Remember from Lecture 13**

1. **WTO replaced GATT on Jan 1, 1995** — the Uruguay Round (1986–94) was the turning point; 166 members, India a founding member

2. **Five core principles:** MFN, National Treatment, Bound Tariffs, Reciprocity, Transparency — these are the rules of the game

3. **DSU is WTO's crown jewel** — binding dispute settlement; India has won and lost important cases; Appellate Body crisis unresolved

4. **Doha Round is effectively dead** — agriculture at the centre of the stalemate; US/EU domestic politics vs. India/developing country demands

5. **Peace Clause (2013) is India's lifeline** — protects MSP-based procurement from WTO challenge; permanent solution still pending

6. **India uses WTO strategically:** high bound tariffs for policy space, G-33/G-20 coalitions for negotiating power, DSU for defending interests
:::

---

## Next Lecture: Agreement on Agriculture {.smaller}

**Lecture 14 (July 28, 2026):** **Agreement on Agriculture (AoA)**

The WTO agreement that **directly governs India's agricultural trade policy**

We will cover:

- **Three Pillars** in depth: Market Access, Domestic Support, Export Competition
- Tariffication and Tariff Rate Quotas; the Three Boxes: Green, Blue, Amber
- India's AMS challenge — the numbers; the Peace Clause revisited
- Special and Differential Treatment

:::{.notes}
- India's export subsidy constraints
- India provides fertiliser subsidies worth ₹2.25 lakh crore per year. Is this WTO-legal? Which 'box' does it fall in? We will work through the answer systematically using the AoA framework.

**Reading:**
- WTO Agreement on Agriculture (full text — 35 pages)
- CUTS International: *India and WTO Agriculture Negotiations* (2023)
- Economic Survey 2023-24, Chapter on Agriculture
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 12)
- *International Economics* — Appleyard & Field (Ch. 12)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
