---
title: "Balance of Trade & Balance of Payments"
subtitle: "ECO 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "June 20, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "ECO 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(patchwork)
options(scipen = 999)
theme_set(theme_minimal(base_size = 11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

# {.section}

::: {style="text-align:center; margin-top: 2em;"}
**Lecture 9**

[Balance of Trade & Balance of Payments]{.hi style="font-size:1.6em;"}

Following the Money: India's External Sector
:::

---

## Introduction: Why the BoP Matters

::: {.columns}
::: {.column width="55%"}
**Trade policy affects the current account.**
Exchange rate policy affects the financial account.
Both show up in the **Balance of Payments (BoP)**.

**Key questions for today:**

- How does a country "pay" for its imports?
- What happens when a country consistently imports more than it exports?
- How do capital flows balance the external accounts?
- When does a BoP *crisis* occur?

**India's external sector at a glance (FY2024):**

- Merchandise exports: **$437 billion** (goods)
- Merchandise imports: **$677 billion** (goods)
- Trade deficit: **-$240 billion** (goods only)
- Services surplus: **+$163B**
- Remittances: **+$120B** (world's largest recipient)
- Current account: **-$23B (-0.7% of GDP)**
:::
::: {.column width="45%"}
::: {.highlightbox}
**BoP accounting identity:**

$$\underbrace{\text{CA}}_{\text{trade + income}} + \underbrace{\text{KA}}_{\text{transfers}} + \underbrace{\text{FA}}_{\text{capital flows}} + \underbrace{\Delta R}_{\text{reserve change}} = 0$$

The BoP **always balances** — by accounting construction.

A current account deficit *must* be financed by:
1. Capital/financial inflows (borrowing, FDI, FPI), or
2. Running down foreign exchange reserves

If neither is available → **BoP crisis** (India 1991).
:::
:::
:::

---

## Balance of Trade: The Basics

::: {.columns}
::: {.column width="55%"}
**Balance of Trade (BoT)** = Merchandise exports − Merchandise imports

$$\text{BoT} = X_{\text{goods}} - M_{\text{goods}}$$

- Positive BoT = **trade surplus** (exports > imports)
- Negative BoT = **trade deficit** (imports > exports)

**India's BoT trend (USD billion):**

| FY | Exports | Imports | Trade Balance |
|---|---|---|---|
| FY20 | 314 | 478 | **-157** |
| FY21 | 291 | 389 | **-98** |
| FY22 | 422 | 614 | **-190** |
| FY23 | 451 | 714 | **-264** |
| FY24 | 437 | 677 | **-238** |

India **always** runs a merchandise trade deficit — driven by oil, gold, and electronics imports.
:::
::: {.column width="45%"}
::: {.methodbox}
**Why India's trade deficit is structural:**

**Import composition (FY24):**

| Category | $B | % |
|---|---|---|
| Petroleum crude | 233 | 34% |
| Gold & silver | 49 | 7% |
| Electronic goods | 82 | 12% |
| Machinery | 55 | 8% |
| Other | 258 | 38% |

India has very limited domestic crude oil production (~0.8 mb/d vs consumption ~5 mb/d) → structurally dependent on oil imports.

Gold imports: 700–800 tonnes/year → cultural and investment demand.
:::
:::
:::

---

## National Income Identity Approach to the Current Account

::: {.columns}
::: {.column width="58%"}
**National Income and Product Accounts (NIPA):**

$$Y = C + I + G + (X - M)$$

Rearranging:

$$X - M = Y - (C + I + G) = Y - A$$

where $A = C + I + G$ is **absorption** (total domestic spending).

**Implication:**

$$\boxed{\text{Current Account} \approx Y - A}$$

- $\text{CA} > 0 \Leftrightarrow Y > A$: country produces more than it spends → *net lender* to world
- $\text{CA} < 0 \Leftrightarrow Y < A$: country spends more than it produces → *net borrower* from world

**Saving-investment identity:**

$$Y - A = Y - C - I - G = (Y - C - G) - I = S_{\text{private}} + S_{\text{govt}} - I$$

$$\boxed{\text{CA} = (S - I) + (T - G)}$$

CA deficit = private dis-saving OR investment boom OR government deficit (or combination).
:::
::: {.column width="42%"}
::: {.keybox}
**India's CA decomposition (FY2024):**

$$\text{CA} = (S - I) + (T - G)$$

- Private saving rate: ~31% of GDP
- Investment rate: ~32% of GDP → private $S - I \approx -1\%$ GDP
- Fiscal deficit: ~5.1% of GDP → $T - G \approx -5.1\%$
- CA = $-1\% + (-5.1\%) \approx -6\%$?

But India's CA = only $-0.7\%$ GDP!

**Why?** Remittances ($120B) and services surplus ($163B) add back ~9-10% of GDP in non-merchandise credits.

**Lesson:** CA = S-I identity holds, but "saving" must include remittance income and service exports.
:::
:::
:::

---

## Current and Capital Account Structure

::: {.columns}
::: {.column width="55%"}
**Account 1: Current Account (CA)**

- Merchandise trade (goods): exports (+), imports (-)
- Services trade: software, tourism, BPO
- Primary income: investment income (interest, dividends, profits) — net
- Secondary income: remittances, grants, transfers

**Account 2: Capital Account (KA)**

- Capital transfers (debt forgiveness, migrant assets)
- Acquisition/disposal of non-produced, non-financial assets
- (Small for most countries including India)
:::
::: {.column width="45%"}
::: {.methodbox}
**Accounting rule:**

Every international transaction has **two entries** of equal and opposite sign (double-entry bookkeeping):

**Example: India imports $10M of machines:**

- Current Account (imports): $-10M$
- Financial Account (credit from foreign bank): $+10M$

**Example: Indian software firm receives $5M:**

- Current Account (services export): $+5M$
- Financial Account (forex held abroad initially): $-5M$
:::
:::
:::

## Financial Account and BoP Identity

::: {.columns}
::: {.column width="55%"}
**Account 3: Financial Account (FA)**

- Foreign Direct Investment (FDI): equity + reinvested earnings
- Portfolio Investment (FPI/FII): equities, bonds
- Other investment: loans, trade credit, bank deposits
- Official reserves (RBI intervention)

**BoP identity:**

$$\text{CA} + \text{KA} + \text{FA} + \Delta R = 0$$

The BoP **always balances** — by accounting construction. A current account deficit *must* be financed by:

1. Capital/financial inflows (borrowing, FDI, FPI), or
2. Running down foreign exchange reserves

If neither is available → **BoP crisis** (India 1991).
:::
::: {.column width="45%"}
::: {.keybox}
**India's BoP summary (FY2024):**

| Component | USD B |
|---|---|
| Current Account | −23 |
| Financial Account | +20 |
| Δ Reserves | +3 |
| **Total** | **≈ 0** |

India's current account deficit of ~$23B is financed primarily through FDI ($22B) and FPI ($41B) inflows — making financial account management critical to external stability.
:::
:::
:::

---

## India's BoP Components: FY2020–FY2024

```{r}
#| label: fig-bop-india
#| fig-cap: "India's Balance of Payments Components, FY2020–FY2024 (USD billion)\nSource: RBI, Balance of Payments Statistics."
#| fig-width: 9
#| fig-height: 5

bop_data <- data.frame(
  year           = rep(c("FY20","FY21","FY22","FY23","FY24"), each = 6),
  component      = rep(c("Merchandise\nTrade Balance", "Services\nSurplus",
                         "Remittances", "Investment\nIncome (net)",
                         "FDI (net)", "FPI (net)"), times = 5),
  value          = c(
    # FY20
    -157,  85,  83, -26,  43, -14,
    # FY21
     -98,  89,  83, -29,  44,  36,
    # FY22
    -190, 107,  89, -33,  39, -16,
    # FY23
    -264, 143, 112, -39,  28, -16,
    # FY24
    -238, 163, 120, -68,  22,  41
  )
)

bop_data <- bop_data |>
  mutate(
    account = ifelse(component %in% c("Merchandise\nTrade Balance",
                                       "Services\nSurplus",
                                       "Remittances",
                                       "Investment\nIncome (net)"),
                     "Current Account", "Financial Account"),
    component = factor(component,
                       levels = c("Merchandise\nTrade Balance",
                                  "Investment\nIncome (net)",
                                  "FPI (net)",
                                  "FDI (net)",
                                  "Services\nSurplus",
                                  "Remittances"))
  )

ggplot(bop_data, aes(x = year, y = value, fill = component)) +
  geom_col(position = "dodge", width = 0.85) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "grey30") +
  scale_fill_manual(values = c(
    "Merchandise\nTrade Balance" = "#8B0000",
    "Investment\nIncome (net)"  = "#cc4444",
    "FPI (net)"                  = "#B9975B",
    "FDI (net)"                  = "#7B9F35",
    "Services\nSurplus"          = "#012169",
    "Remittances"                = "#5B8DB8"
  )) +
  labs(
    title    = "India's Balance of Payments Components, FY2020–FY2024",
    subtitle = "Merchandise deficit offset by services surplus, remittances, and (usually) financial inflows",
    x = NULL, y = "USD Billion", fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(size = 8),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2))
```

---

## India's Current Account: FY2024 Decomposition

```{r}
#| label: fig-ca-waterfall
#| fig-cap: "India's Current Account, FY2024: Component Contributions (USD Billion)\nSource: RBI, Balance of Payments Statistics."
#| fig-width: 7.5
#| fig-height: 4.5

# CA = merch balance + services + remittances + investment income
# = -238 + 163 + 120 - 68 = -23 (approx)
ca_items <- data.frame(
  item      = factor(
    c("Merch.\nExports", "Merch.\nImports", "Services\nSurplus",
      "Remittances", "Inv. Income\n(net)", "CA Balance"),
    levels = c("Merch.\nExports", "Merch.\nImports", "Services\nSurplus",
               "Remittances", "Inv. Income\n(net)", "CA Balance")
  ),
  value     = c( 437, -677,  163,  120, -68, -23),
  fill_col  = c("#012169", "#8B0000", "#012169", "#5B8DB8", "#8B0000", "grey30"),
  label_pos = c( 437, -677,  163,  120, -68, -23)
)

ggplot(ca_items, aes(x = item, y = value, fill = fill_col)) +
  geom_col(width = 0.65) +
  scale_fill_identity() +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  geom_text(
    aes(
      label = paste0(ifelse(value >= 0, "+", ""), value, "B"),
      vjust = ifelse(value >= 0, -0.4, 1.3)
    ),
    size = 3.5, fontface = "bold"
  ) +
  scale_y_continuous(
    limits = c(-720, 520),
    labels = label_dollar(prefix = "$", suffix = "B")
  ) +
  labs(
    title    = "India's Current Account Components, FY2024 (USD Billion)",
    subtitle = "Gross exports shown separately from imports for clarity\nCA deficit = -$23B (-0.7% of GDP) — manageable",
    x = NULL, y = "USD Billion"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

---

## Capital and Financial Account: India FY2024

::: {.columns}
::: {.column width="55%"}
**Financial Account flows (FY2024, USD billion):**

| Component | Amount | Notes |
|---|---|---|
| FDI (net inflows) | +22 | Manufacturing, services |
| FPI (net) | +41 | Equity + bonds |
| External commercial borrowings | +5 | Corporate loans |
| NRI deposits | +8 | FCNR, NRE accounts |
| Banking capital | +12 | |
| **Total FA** | **+88** | |

**Reserve change:**
$$\Delta R = -(CA + KA + FA) = -(-23 + 0 + 88) = -65$$

Reserves **increased** by $65B in FY2024 (net RBI purchases of forex).

India's forex reserves: $646B (March 2024) — ~11 months of imports.
:::
::: {.column width="45%"}
::: {.methodbox}
**FDI vs FPI — key distinctions:**

| Feature | FDI | FPI |
|---|---|---|
| Threshold | ≥10% equity | <10% equity |
| Stability | Long-term | Short-term ("hot money") |
| Volatility | Low | High |
| Technology transfer | Yes | No |
| RBI regulation | FEMA routes | SEBI (FPI licence) |

**India's FDI by sector (FY2024):**
Services (27%), Computer software (18%), Construction (7%), Auto (5%), Pharma (5%).

**Agriculture FDI:** Only ~1.5% of total — food processing and cold chain attract more than primary agriculture.
:::
:::
:::

---

## Official Reserves and RBI Intervention

::: {.columns}
::: {.column width="55%"}
**Official Foreign Exchange Reserves** = foreign assets held by central bank (RBI):

- Foreign currency assets (mainly USD, EUR, GBP, JPY)
- Gold
- SDRs (IMF Special Drawing Rights)
- Reserve tranche position at IMF

**India's reserves trajectory:**

| Year | Reserves (USD B) | Import cover |
|---|---|---|
| FY1991 | 1.2 | 2 weeks |
| FY2000 | 38 | 8 months |
| FY2008 | 309 | 14 months |
| FY2020 | 477 | 12 months |
| FY2024 | 646 | 11 months |

**Adequacy benchmark:** 3 months of imports (IMF rule of thumb). India holds >11 months.
:::
::: {.column width="45%"}
::: {.highlightbox}
**RBI's intervention role:**

RBI buys/sells USD to manage the exchange rate:

- **Buy USD** (sell INR) → prevents excessive appreciation, builds reserves
- **Sell USD** (buy INR) → prevents excessive depreciation, depletes reserves

**Managed float:**
India operates a "managed float" — neither fully fixed nor fully free.

RBI intervenes to:
1. Smooth volatility (not set a target level)
2. Prevent disorderly market conditions
3. Maintain competitive export pricing (keep INR from appreciating too fast)

**FY2024:** RBI was a net buyer of USD — absorbing strong capital inflows.
:::
:::
:::

---

## BoP Adjustment Mechanisms

::: {.columns}
::: {.column width="55%"}
When a BoP deficit exists, three adjustment mechanisms operate:

**1. Price mechanism (exchange rate adjustment):**

Currency depreciates → exports cheaper, imports expensive → CA improves (if M-L condition holds).

**2. Income/expenditure mechanism:**

CA deficit → domestic income falls (through multiplier) → absorption $A$ falls → CA improves.

$$\Delta M = m \cdot \Delta Y \quad \text{(marginal propensity to import)}$$

**3. Monetary mechanism:**

CA deficit → reserve outflow → money supply falls → domestic prices fall → exports competitive again (Hume's price-specie-flow mechanism).

Modern version: reserve loss tightens monetary conditions → interest rates rise → capital inflows.
:::
::: {.column width="45%"}
::: {.keybox}
**Policy instruments:**

| Problem | Instrument | Channel |
|---|---|---|
| CA deficit | Devaluation | Price competitiveness |
| CA deficit | Expenditure reduction | Income mechanism |
| FA deficit | Interest rate rise | Capital inflows |
| Reserve crisis | IMF credit | Financing |

**Expenditure-switching vs Expenditure-reducing:**

- *Switching*: devaluation shifts demand from foreign to domestic goods
- *Reducing*: fiscal/monetary austerity reduces total absorption

Both often needed simultaneously — as India learned in 1991.
:::
:::
:::

---

## Marshall-Lerner Condition: When Does Depreciation Help?

::: {.columns}
::: {.column width="58%"}
**Question:** Does a currency depreciation always improve the trade balance?

**Answer:** Only if demand is elastic enough.

**Derivation:**

Let exchange rate $e$ = domestic currency per unit of foreign. Depreciation means $\Delta e > 0$.

Trade balance in domestic currency:

$$TB = P_X \cdot Q_X - e \cdot P_M^* \cdot Q_M$$

where $P_M^*$ = foreign price of imports, $Q_X, Q_M$ = quantities.

After depreciation, $e \uparrow$:
- Export price (foreign) falls → $Q_X \uparrow$ by $|e_X| \cdot \frac{\Delta e}{e}$
- Import price (domestic) rises → $Q_M \downarrow$ by $|e_M| \cdot \frac{\Delta e}{e}$

**Marshall-Lerner (M-L) condition:**

$$\boxed{|e_X| + |e_M| > 1}$$

If satisfied, depreciation *improves* the trade balance. If $|e_X| + |e_M| < 1$, depreciation *worsens* TB (short-run typical).
:::
::: {.column width="42%"}
::: {.methodbox}
**Intuition:**

A 10% depreciation:

- Makes exports 10% cheaper abroad → quantity effect (positive)
- Makes imports 10% expensive at home → quantity effect (positive for TB)
- But also raises value of existing imports → value effect (negative for TB)

**M-L says:** quantity effects must outweigh value effect.

**Typical elasticity estimates (long-run):**

| $|e_X|$ | $|e_M|$ | Sum | M-L satisfied? |
|---|---|---|---|
| India | 0.8–1.2 | 0.7–1.0 | ~1.5–2.2 | **Yes (long-run)** |
| Short-run | 0.3–0.5 | 0.2–0.4 | ~0.5–0.9 | **No → J-curve** |

Short-run inelasticity: contracts pre-set, adjustment takes time.
:::
:::
:::

---

## The J-Curve Effect

```{r}
#| label: fig-jcurve
#| fig-cap: "J-Curve: Trade Balance Initially Worsens After Currency Depreciation\nSource: Author's illustration."
#| fig-width: 8
#| fig-height: 4.5

set.seed(42)
t <- seq(-5, 14, by = 0.2)

# J-curve shape: initial deterioration then improvement
# TB(t<0) = 0 (baseline)
# After depreciation at t=0:
#   short run (0-3 months): TB falls (pre-set contracts, inelastic demand)
#   medium (3-12 months):   TB recovers as quantities adjust
#   long run (12+ months):  TB improves above baseline if M-L holds

tb_j <- ifelse(
  t < 0, 0,
  ifelse(
    t < 3,
    -4 * (t/3),                                    # deterioration phase
    ifelse(
      t < 8,
      -4 + 5 * ((t - 3)/5),                        # recovery phase
      1 + 0.3 * (t - 8)                             # long-run improvement
    )
  )
)

# Add some noise for realism
tb_j <- tb_j + c(rep(0, sum(t < 0)), rnorm(sum(t >= 0), 0, 0.15))

df <- data.frame(time = t, tb = tb_j)

ggplot(df, aes(x = time, y = tb)) +
  geom_line(color = "#012169", linewidth = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1.2) +
  # Shaded deterioration region
  geom_ribbon(
    data = df |> filter(time >= 0 & time <= 4),
    aes(ymin = pmin(tb, 0), ymax = 0),
    fill = "#cc0000", alpha = 0.2
  ) +
  # Shaded improvement region
  geom_ribbon(
    data = df |> filter(time > 4),
    aes(ymin = 0, ymax = pmax(tb, 0)),
    fill = "#012169", alpha = 0.15
  ) +
  # Annotations
  annotate("text", x = 0.3, y = 3.8,
           label = "Depreciation\noccurs (t=0)", color = "red", size = 3.5, hjust = 0) +
  annotate("text", x = -3.5, y = 0.4,
           label = "Pre-depreciation\nbaseline", size = 3, color = "grey40") +
  annotate("text", x = 1.5, y = -2.8,
           label = "Short-run\ndeterioration\n(contracts pre-set,\nM-L not satisfied)",
           size = 3, color = "#cc0000", hjust = 0.5) +
  annotate("text", x = 11, y = 2.2,
           label = "Long-run\nimprovement\n(M-L holds,\nquantities adjust)",
           size = 3, color = "#012169", hjust = 0.5) +
  annotate("segment",
           x = 1.5, y = -2.3, xend = 1.5, yend = -1,
           arrow = arrow(length = unit(0.25, "cm")), color = "#cc0000") +
  annotate("segment",
           x = 11, y = 1.8, xend = 11, yend = 0.6,
           arrow = arrow(length = unit(0.25, "cm")), color = "#012169") +
  # Phase labels
  annotate("text", x = 1.5, y = -4.3,
           label = "Phase I:\n0-3 months", size = 2.8, color = "#cc0000") +
  annotate("text", x = 5.5, y = -4.3,
           label = "Phase II:\n3-12 months\n(recovery)", size = 2.8, color = "#B9975B") +
  annotate("text", x = 11,  y = -4.3,
           label = "Phase III:\n12+ months\n(improvement)", size = 2.8, color = "#012169") +
  scale_x_continuous(breaks = seq(-4, 14, 2),
                     labels = paste0(seq(-4, 14, 2), "m")) +
  scale_y_continuous(limits = c(-5.5, 5)) +
  labs(
    title    = "The J-Curve Effect",
    subtitle = "Trade balance worsens before improving after currency depreciation\nM-L condition (|eX|+|eM|>1) is satisfied only in the long run",
    x = "Time (months after depreciation)", y = "Trade Balance (change from baseline)"
  ) +
  theme_minimal(base_size = 11)
```

---

## J-Curve: Mechanisms

::: {.columns}
::: {.column width="55%"}
**Why does TB worsen initially?**

**Phase I (0–3 months): Contracts and inelasticity**

- Import/export contracts already written in foreign currency
- Prices change, but quantities cannot adjust instantly
- Rising import prices → import *bill* rises (value effect > quantity effect)
- M-L condition NOT satisfied in short run: $|e_X^{SR}| + |e_M^{SR}| < 1$

**Phase II (3–12 months): Gradual adjustment**

- Firms begin adjusting: export more, seek import substitutes
- Trade volumes start responding to price signals

**Phase III (12+ months): Full adjustment**

- Contracts renegotiated
- Production capacity adjusted
- Consumers switch to domestic substitutes
- M-L holds: $|e_X^{LR}| + |e_M^{LR}| > 1$
:::
::: {.column width="45%"}
::: {.methodbox}
**India's empirical evidence:**

INR depreciation episodes:
- **1991:** 25% devaluation → J-curve lasted ~2 years (severe structural rigidities)
- **2013 taper tantrum:** INR fell 20% → TB improved within 12 months (better-integrated trade)
- **2022 (post-Russia-Ukraine):** INR weakened 10% → limited TB improvement (oil price dominated)

**Agricultural trade J-curve:**

Agricultural exports show faster adjustment than industrial goods — lower contract lock-in for spot commodity trades.

But Indian farmers often price in INR and don't automatically capture depreciation gains without better market linkages.
:::
:::
:::

---

## Absorption Approach: Policy Implications

::: {.columns}
::: {.column width="55%"}
**Absorption approach (Sidney Alexander, 1952):**

$$CA = Y - A$$

To improve the CA (reduce deficit), need:

$$\Delta CA = \Delta Y - \Delta A > 0$$

Either:
1. **Raise output** $Y$: supply-side reforms, productivity growth
2. **Reduce absorption** $A$: cut $C$, $I$, or $G$ (expenditure reduction)
3. **Both** (typical IMF conditionality)

**Devaluation alone is insufficient:**

If $A$ rises as fast as $Y$ (say, because higher export revenues are spent on imports), CA does not improve.

**Sterilisation problem:**
If devaluation raises domestic price level (inflation), real $A$ may not fall → M-L gains are inflated away.
:::
::: {.column width="45%"}
::: {.keybox}
**India 1991 adjustment:**

$$\Delta CA = \Delta Y - \Delta A$$

IMF conditionality required:
1. Fiscal deficit cut: $\Delta G < 0$ → reduces $A$
2. 25% devaluation → switches $A$ toward domestic goods
3. Trade liberalisation → improves efficiency of $Y$

**Result:**
- CA went from -3.1% GDP (FY91) to -1.0% (FY93)
- Growth fell temporarily (-0.1% in FY91-92)
- Inflation spiked (13.7% in FY92)

Classic adjustment tradeoff: **external balance restored at cost of internal balance** — at least temporarily.
:::
:::
:::

---

## India's 1991 BoP Crisis

::: {.columns}
::: {.column width="55%"}
**Chronology of the crisis:**

| Date | Event |
|---|---|
| 1988–90 | Gulf remittances fall (Gulf War I); oil prices spike |
| 1990 | Fiscal deficit ~8.5% GDP; political instability |
| Oct 1990 | Moody's downgrades India |
| Jan 1991 | Non-resident Indian deposits withdrawn en masse |
| Mar 1991 | Forex reserves: $5.8B (barely 3 weeks imports) |
| **June 1991** | **Reserves: $1.2B — 2 WEEKS of imports** |
| July 1991 | India pledges **67 tonnes of gold** to Bank of England + Bank of Japan for $600M loan |
| Aug 1991 | IMF Structural Adjustment Package: $2.2B |
| 1991–92 | 25% devaluation; trade and investment liberalisation |
:::
::: {.column width="45%"}
::: {.highlightbox}
**Root causes:**

1. **Fiscal profligacy:** 8.5% fiscal deficit financed by short-term external borrowing
2. **Oil shock (1990):** Iraq invasion of Kuwait → oil price doubled
3. **Remittance shock:** NRI deposits withdrawn on political uncertainty
4. **Structural rigidities:** ISI (import substitution) model had exhausted itself

**IMF conditionality:**

- Fiscal deficit reduction
- Trade liberalisation: import licensing abolished
- Industrial delicensing (1991 Industrial Policy)
- FDI liberalisation: automatic route up to 51%

**Legacy:** The 1991 crisis was the *proximate cause* of India's liberalisation — turning a crisis into a structural reform.
:::
:::
:::

---

## India FY2024 BoP Assessment

::: {.columns}
::: {.column width="55%"}
**Current Account:**

- CA deficit: $23.2B (-0.7% of GDP) — **comfortable**
- Merchandise trade deficit: $238B (oil, gold, electronics)
- Services surplus: $163B (IT/BPO dominance)
- Remittances: $120B — world's largest recipient (World Bank data)

**Financial Account:**

- FDI net inflows: $22B (declined from $83B peak in FY22)
- FPI: +$41B (strong equity market inflows)
- ECB + banking capital: +$25B

**Reserve adequacy:**

- Forex reserves: $646B (March 2024)
- Import cover: ~11 months (**well above IMF 3-month threshold**)
- Short-term debt/reserves: 19% (**well below 100% danger threshold**)
:::
::: {.column width="45%"}
::: {.methodbox}
**Risks and concerns:**

1. **Oil price vulnerability:** $10/barrel rise adds ~$15B to import bill
2. **Gold imports:** Cultural demand makes gold import compression difficult
3. **FPI volatility:** $41B inflow in FY24; could reverse quickly (as in 2013 taper tantrum)
4. **FDI decline:** Net FDI fell from $55B (FY22) to $22B (FY24) — concerning for long-run financing
5. **China factor:** RCEP non-membership means India lacks preferential access; competition from China in exports

**Sustainable CA deficit?**

$$\text{CA/GDP} < \text{long-run growth rate} \times \frac{\text{Net foreign debt}}{\text{GDP}}$$

India's CA/GDP (-0.7%) << GDP growth (7%) — external position is **sustainable** and **comfortable**.
:::
:::
:::

---

## BoP and Agricultural Policy Linkages

::: {.columns}
::: {.column width="55%"}
**How agricultural policy affects BoP:**

1. **Food import substitution:** MSP → higher domestic production → reduces food import bill
   - India became self-sufficient in pulses (FY2023) — saved ~$3B/year
   - Oilseed: still imports ~$22B of edible oils despite MSP

2. **Export competitiveness:** High MSP + input subsidies → high domestic costs → reduces agricultural export competitiveness

3. **Rice exports:** Restrictions in 2023 saved BoP in short run (reduced export earnings but improved food security), hurt long-run reputation

4. **Remittance multiplier:** 5-6M agricultural workers in Gulf + Western countries send ~$40-50B/year from agri-related labour → largest single CA credit item after IT services
:::
::: {.column width="45%"}
::: {.highlightbox}
**India's agricultural trade balance (FY2024):**

| Category | Exports | Imports | Net |
|---|---|---|---|
| Rice | $9.9B | neg. | +$9.9B |
| Spices | $3.7B | $0.4B | +$3.3B |
| Sugar | $2.8B | neg. | +$2.8B |
| Edible oils | $0.8B | $21.7B | **-$20.9B** |
| Pulses | $0.6B | $2.8B | **-$2.2B** |
| **Total Agri** | **~$50B** | **~$30B** | **+$20B** |

Agriculture is a **net positive contributor** to India's BoP — one of the few sectors where India has a structural surplus.

Edible oil imports (dominated by palm oil from Indonesia/Malaysia) are the single largest structural weakness.
:::
:::
:::

---

## India's Services Exports: The IT-BPO Engine

::: {.columns}
::: {.column width="55%"}
**Why India runs a services surplus despite a goods deficit:**

India's IT/ITES sector is the single largest contributor to the services surplus:

| Service category | FY24 exports (USD B) | % share |
|---|---|---|
| Software services | 100 | 61% |
| Business services (BPO) | 28 | 17% |
| Travel (tourism) | 22 | 14% |
| Financial services | 9 | 5% |
| Other | 4 | 3% |
| **Total** | **163** | 100% |

**Top software exporters:** TCS ($26B), Infosys ($18B), HCL ($13B), Wipro ($11B).

India's services exports grew at **12.7% CAGR** over FY2015–FY2024 — one of the fastest-growing export sectors globally.
:::
::: {.column width="45%"}
::: {.highlightbox}
**Why services exports matter for BoP:**

Services exports are **highly stable** (long-term contracts, stickier than goods trade) and **high value-added** (software $100/hour vs rice $0.30/kg).

**Contrast with goods exports:**
- Goods exports volatile (commodity price cycles, logistics disruptions)
- Services exports smooth — even during COVID-19, India's IT exports grew 15%

**Policy implication:**
India's BoP resilience rests on the IT sector. Any disruption (H-1B visa restrictions, AI automation of BPO) would immediately show up in the current account.

**WTO Mode 3 & 4 negotiations:**
India pushes for freer movement of skilled professionals (Mode 4) — a key interest in services trade negotiations.
:::
:::
:::

---

## Remittances: India's Largest Single CA Credit

::: {.columns}
::: {.column width="55%"}
India received **$120 billion** in remittances in FY2024 — the world's largest recipient (World Bank, 2024).

**Top source countries:**

| Country | Share | INR remittance |
|---|---|---|
| USA | 23% | $27.6B |
| UAE | 18% | $21.6B |
| UK | 7% | $8.4B |
| Saudi Arabia | 8% | $9.6B |
| Other Gulf | 12% | $14.4B |
| Others | 32% | $38.4B |

**State distribution:**
- Kerala: 21% (Gulf-dependent)
- Maharashtra: 16%
- Tamil Nadu: 10%
- UP, Bihar: growing Gulf migration

**Remittances are countercyclical:** rise when sending-country economy booms AND when INR depreciates (migrants send more home when rupee is weak).
:::
::: {.column width="45%"}
::: {.methodbox}
**Why remittances matter for agriculture:**

Large fraction of Gulf remittances goes to **agricultural states** (Kerala, UP, Bihar, Rajasthan).

**Channels:**

1. **Consumption smoothing:** farm households use remittances during crop failure
2. **Investment:** farm equipment, irrigation, land purchases
3. **Risk-taking:** remittance income allows shift to higher-value crops
4. **Labour market:** outmigration raises rural wages → pushes up agricultural labour costs

**BoP significance:**

Remittances = $120B vs FDI = $22B (FY24).

India is less dependent on volatile FDI/FPI for BoP balance than most large EMs — remittances provide stable "private foreign aid."
:::
:::
:::

---

## India's Forex Reserves: RBI Intervention Chart

```{r}
#| label: fig-forex-reserves
#| fig-cap: "India's Forex Reserves and INR/USD Exchange Rate, FY2015–FY2024\nSource: RBI, Database on Indian Economy (DBIE)."
#| fig-width: 8.5
#| fig-height: 4.5

reserves_data <- data.frame(
  year      = 2015:2024,
  reserves  = c(341, 360, 370, 424, 412, 477, 590, 633, 578, 646),
  inr_usd   = c(62.1, 65.5, 67.1, 64.5, 69.9, 70.4, 74.2, 75.0, 82.7, 83.3)
)

# Dual-axis plot using secondary axis
coeff <- 10   # scale factor for INR (reserves ~100x INR so scale accordingly

p_res <- ggplot(reserves_data, aes(x = year)) +
  # Reserves bars
  geom_col(aes(y = reserves), fill = "#012169", alpha = 0.75, width = 0.6) +
  # INR/USD line (scaled)
  geom_line(aes(y = inr_usd * coeff), color = "#cc0000", linewidth = 1.5) +
  geom_point(aes(y = inr_usd * coeff), color = "#cc0000", size = 2.5) +
  # Mark 2022 taper tantrum equivalent
  annotate("text", x = 2023, y = 920,
           label = "Post-Ukraine\nrupee weakness\n(₹83/USD)", size = 3, color = "#cc0000") +
  # Mark 2020 reserve build
  annotate("text", x = 2021, y = 650,
           label = "$590B\n(COVID\nreserve build)", size = 3, color = "#012169") +
  scale_x_continuous(breaks = 2015:2024,
                     labels = paste0("FY", substr(2015:2024, 3, 4))) +
  scale_y_continuous(
    name     = "Forex Reserves (USD billion)",
    limits   = c(0, 1000),
    sec.axis = sec_axis(~ . / coeff,
                        name   = "INR per USD",
                        breaks = seq(60, 90, 5))
  ) +
  labs(
    title    = "India's Forex Reserves (bars) and INR/USD Rate (line), FY2015–FY2024",
    subtitle = "RBI builds reserves during capital inflow surges; deploys during stress periods",
    x = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.title.y.right = element_text(color = "#cc0000"),
    axis.text.y.right  = element_text(color = "#cc0000"),
    axis.title.y.left  = element_text(color = "#012169")
  )

p_res
```

---

## Key Takeaways: Lecture 9

::: {.keybox}
**1. National income identity:** $\text{CA} = Y - A = (S - I) + (T - G)$. A CA deficit means the country borrows from abroad — financed by FA inflows or reserve drawdown.

**2. BoP accounting:** $\text{CA} + \text{KA} + \text{FA} + \Delta R = 0$. The BoP always balances — surpluses and deficits are in *individual accounts*, not the overall BoP.

**3. India's external position (FY2024):** CA = -0.7% GDP (comfortable); remittances = $120B (world's largest); services surplus = $163B; forex reserves = $646B (11 months imports). External position is **strong**.

**4. Marshall-Lerner condition:** $|e_X| + |e_M| > 1$ must hold for depreciation to improve TB. In the short run, M-L is typically NOT satisfied → J-curve: TB worsens before improving.

**5. India 1991 crisis:** Reserves fell to $1.2B (2 weeks imports); India pledged 67 tonnes gold to the BoE. IMF bailout triggered liberalisation — delicensing, trade reform, FDI opening. Crisis → structural reform.

**6. Agricultural BoP:** India runs a trade surplus in agriculture (~$20B), but edible oil imports (-$21B) are the single largest structural weakness. MSP policy affects both food security and export competitiveness.
:::

---

## Next Lecture Preview

::: {.methodbox}
**Lecture 10 — Exchange Rates and Agricultural Trade**
*June 27, 2026*

- Fixed vs flexible exchange rate systems
- Purchasing Power Parity (PPP): absolute and relative
- Real exchange rate: $q = eP^*/P$ and agricultural competitiveness
- How rupee depreciation affects Indian rice and wheat exports
- Dutch disease: can commodity booms harm agricultural exporters?
- India's managed float: RBI intervention and the REER
:::

::: {.highlightbox}
**Preparation:** Review Salvatore Ch. 14–15 (Exchange Rates). Note that India operates a **managed float** — this means understanding both PPP theory AND RBI's intervention behaviour. We will use regression analysis of real exchange rate and export volume data.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 13-14)
- *International Economics* — Appleyard & Field (Ch. 13-14)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
