---
title: "Theories of Trade II: Comparative Advantage"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "May 16, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(patchwork)
library(dplyr)
library(scales)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## Recap: The Gap in Smith's Theory

**Last lecture:** Adam Smith argued that countries should export goods they produce *absolutely* more efficiently.

**But this raises an obvious question:**

> What if one country is *better at producing everything*?

- Should the more efficient country trade at all?
- Should the less efficient country simply import everything?
- Would the inefficient country have *any* exports?

Smith's absolute advantage gives **no answer**.

::: {.methodbox}
**Stylised fact:** The United States is more productive than Bangladesh in *both* textiles *and* aircraft. Yet the US imports textiles from Bangladesh. **Why?** → Enter David Ricardo (1817).
:::

## Ricardo's Insight: What Matters is *Relative* Cost

Ricardo's key insight was deceptively simple:

> A country should specialise in the good for which it has the **lower opportunity cost** — regardless of absolute productivity.

**Opportunity cost** = what you must give up to produce one more unit of a good.

Even a country that is absolutely less efficient in *everything* can still gain from trade by exporting the good in which it is **relatively** less inefficient.

::: {.methodbox}
**Analogy:** A lawyer who types faster than her secretary still benefits from hiring the secretary — because the lawyer's time is more valuably spent on legal work. The same logic applies to nations.
:::

## The Labor-Value Model: Notation

We work with the simplest possible model:

- **Two countries:** India ($I$) and the World ($W$)
- **Two goods:** Rice ($R$) and Wheat ($Wh$)
- **One factor of production:** Labour ($L$)
- **Technology:** constant labour requirements per unit of output

**Labour coefficients** (units of labour per unit of output):

| | Rice | Wheat |
|---|---|---|
| India | $a_{LR}^{I}$ | $a_{LWh}^{I}$ |
| World | $a_{LR}^{W}$ | $a_{LWh}^{W}$ |

$a_{LR}^{I}$ = labour hours required to produce **one unit of Rice in India**.

## Numerical Example {.smaller}

**Assume:** each country has $L = 100$ workers. One worker per day produces:

| Country | Rice | Wheat |
|---|---|---|
| India | **2 units** | **1 unit** |
| World | **3 units** | **3 units** |

Labour coefficients: $a_{LR}^{I} = \tfrac{1}{2}$, $a_{LWh}^{I} = 1$; $a_{LR}^{W} = \tfrac{1}{3}$, $a_{LWh}^{W} = \tfrac{1}{3}$.
The **World** has absolute advantage in **both** goods.

**Opportunity costs:**

$$OC_R^{I} = \frac{a_{LR}^{I}}{a_{LWh}^{I}} = \frac{1/2}{1} = \frac{1}{2} \text{ Wheat} \qquad OC_R^{W} = \frac{a_{LR}^{W}}{a_{LWh}^{W}} = \frac{1/3}{1/3} = 1 \text{ Wheat}$$

**Comparison:** $OC_R^{I} = \tfrac{1}{2} < OC_R^{W} = 1$ $\Rightarrow$ **India has comparative advantage in Rice.**

## The Comparative Advantage Condition

**Formal statement:**

India has a comparative advantage in Rice if and only if:

$$\frac{a_{LR}^{I}}{a_{LWh}^{I}} < \frac{a_{LR}^{W}}{a_{LWh}^{W}}$$

Equivalently, rearranging:

$$\frac{a_{LR}^{I}}{a_{LR}^{W}} < \frac{a_{LWh}^{I}}{a_{LWh}^{W}}$$

India is *relatively* more efficient in Rice than in Wheat.

::: {.callout-note}
## Key takeaway
Absolute productivity levels ($a_{LR}^{I}$ vs $a_{LR}^{W}$) do **not** determine trade patterns. **Relative** productivity — the ratio of ratios — does.
:::

## Production Possibility Frontiers

With $L = 100$ workers, each country's PPF is a straight line (constant opportunity cost):

::: {.columns}
::: {.column width="40%"}
**India:**
$$R + 2 \cdot Wh = 200$$
Slope $= -2$ (give up 2 Rice per Wheat)

**World:**
$$R + 1 \cdot Wh = 300$$
Slope $= -1$ (give up 1 Rice per Wheat)

The **slope** of the PPF = opportunity cost of Wheat.
:::
::: {.column width="60%"}
```{r}
#| label: fig-ppf
#| fig-cap: "Production Possibility Frontiers: India and World\nSource: Author's illustration."
#| fig-width: 9
#| fig-height: 5
library(ggplot2)
library(patchwork)

p1 <- ggplot() +
  geom_segment(aes(x=0, y=200, xend=100, yend=0),
               color="#012169", linewidth=1.5) +
  geom_point(aes(x=0, y=200), color="#012169", size=3) +
  geom_point(aes(x=100, y=0), color="#012169", size=3) +
  annotate("text", x=50, y=120,
           label="India's PPF\nSlope = -2\n(OC of Wheat = 2 Rice)",
           color="#012169", size=3.5, hjust=0.5) +
  annotate("text", x=103, y=0, label="Wheat", size=3.5, hjust=0) +
  annotate("text", x=0, y=210, label="Rice", size=3.5, hjust=0.5) +
  labs(title="India", x="Wheat (units)", y="Rice (units)") +
  theme_minimal(base_size=11) +
  theme(plot.title=element_text(color="#012169", face="bold"))

p2 <- ggplot() +
  geom_segment(aes(x=0, y=300, xend=300, yend=0),
               color="#B9975B", linewidth=1.5) +
  geom_point(aes(x=0, y=300), color="#B9975B", size=3) +
  geom_point(aes(x=300, y=0), color="#B9975B", size=3) +
  annotate("text", x=150, y=200,
           label="World PPF\nSlope = -1\n(OC of Wheat = 1 Rice)",
           color="#B9975B", size=3.5, hjust=0.5) +
  annotate("text", x=303, y=0, label="Wheat", size=3.5, hjust=0) +
  annotate("text", x=0, y=315, label="Rice", size=3.5, hjust=0.5) +
  labs(title="World", x="Wheat (units)", y="Rice (units)") +
  theme_minimal(base_size=11) +
  theme(plot.title=element_text(color="#B9975B", face="bold"))

p1 + p2
```
:::
:::

## The Terms of Trade Range

For trade to be mutually beneficial, the world price ratio must lie **between** the two autarky opportunity costs.

Autarky opportunity costs of Wheat (in terms of Rice):

$$OC_{Wh}^{I} = 2 \text{ Rice} \qquad OC_{Wh}^{W} = 1 \text{ Rice}$$

The **terms of trade** (price of Wheat relative to Rice) must satisfy:

$$1 < \frac{P_{Wh}}{P_R} < 2$$

- If $P_{Wh}/P_R \leq 1$: India would not import Wheat (cheaper to produce at home)
- If $P_{Wh}/P_R \geq 2$: World would not export Wheat (too expensive)
- At any intermediate price, **both countries benefit**.

**Equilibrium ToT** is determined by relative demand — say $P_{Wh}/P_R = 1.5$.

## Gains from Trade: India's Perspective

::: {.columns}
::: {.column width="40%"}
**Mechanism:**

1. India **fully specialises** in Rice → produces (0 Wheat, 200 Rice)
2. Trades Rice for Wheat at world price $P_{Wh}/P_R = 1.5$
3. **Consumption** moves to a point **above** the PPF

The trade possibilities line starts at full specialisation (0, 200) with slope $= -1.5$ (less steep in absolute value than the PPF slope $-2$).

Consumption bundle **above the PPF** = welfare gain!
:::
::: {.column width="60%"}
```{r}
#| label: fig-gains
#| fig-cap: "India: Gains from Trade — Consuming Beyond the PPF\nSource: Author's illustration."
#| fig-width: 7
#| fig-height: 5
library(ggplot2)

wheat_seq <- seq(0, 120, by=0.5)
rice_ppf   <- pmax(0, 200 - 2*wheat_seq)
rice_trade <- pmax(0, 200 - 1.5*wheat_seq)   # P_Wh/P_R = 1.5

df_ppf <- data.frame(wheat=wheat_seq, rice_ppf=rice_ppf, rice_trade=rice_trade)

ggplot(df_ppf, aes(x=wheat)) +
  geom_line(aes(y=rice_ppf), color="#012169", linewidth=1.5) +
  geom_line(aes(y=rice_trade), color="#B9975B", linewidth=1.5, linetype="dashed") +
  geom_point(aes(x=50, y=100), color="#012169", size=4) +
  annotate("text", x=54, y=100,
           label="Pre-trade\n(50Wh, 100R)", hjust=0, size=3.2, color="#012169") +
  geom_point(aes(x=67, y=100), color="#B9975B", size=4) +
  annotate("text", x=71, y=100,
           label="Post-trade\n(67Wh, 100R)\nabove PPF!",
           hjust=0, size=3.2, color="#B9975B") +
  geom_point(aes(x=0, y=200), color="darkgreen", size=3) +
  annotate("text", x=4, y=204,
           label="Full specialisation\n(0Wh, 200R)", hjust=0, size=3.2, color="darkgreen") +
  annotate("segment", x=50, y=100, xend=66, yend=100,
           arrow=arrow(length=unit(0.25,"cm")), color="darkgreen", linewidth=1) +
  annotate("text", x=55, y=108, label="Gains from trade!",
           color="darkgreen", size=3.5, fontface="bold") +
  labs(title="India specialises in Rice, trades for Wheat",
       subtitle="Trade line lies above the PPF — consumption set expands",
       x="Wheat (units)", y="Rice (units)") +
  scale_x_continuous(limits=c(0,120)) +
  scale_y_continuous(limits=c(0,220)) +
  theme_minimal(base_size=11)
```
:::
:::

## Quantifying the Gains {.smaller}

**Before trade** (autarky): India produces and consumes on the PPF at $(Wh, R) = (50, 100)$

**After trade** (at $P_{Wh}/P_R = 1.5$): India fully specialises → 200 Rice, trades $\approx 100$ Rice for $\approx 67$ Wheat:
$$(Wh, R) \approx (67, 100)$$

**Welfare gain:** 17 more units of Wheat at the same Rice — the **real wage** of Indian workers rises.

**The algebra:** budget constraint under trade: $R + 1.5 \cdot Wh = 200$

Compare with autarky constraint: $R + 2 \cdot Wh = 200$

The trade budget line is **flatter** → larger feasible consumption set.

## Consumer Preferences and Equilibrium

To find the *exact* equilibrium consumption bundle, we need **indifference curves (ICs)**.

::: {.columns}
::: {.column width="50%"}
**Properties of ICs:**

- Downward sloping (MRS > 0)
- Convex to origin (diminishing MRS)
- Higher IC = higher utility

**Autarky equilibrium (E1):**

$$\text{MRS} = \frac{P_{Wh}}{P_R} = \frac{a_{LWh}^I}{a_{LR}^I} = 2$$

IC1 is tangent to the PPF at E1.

**Free-trade equilibrium (E2):**

IC2 is tangent to the **trade possibilities line** at E2.

Since IC2 is above IC1: **trade raises welfare**.
:::
::: {.column width="50%"}
```{r}
#| label: fig-ic-ppf
#| fig-cap: "Free Trade Equilibrium: IC tangent to Trade Possibilities Line\nSource: Author's illustration."
#| fig-width: 6
#| fig-height: 5
library(ggplot2)

wheat_seq <- seq(1, 105, by=0.5)
rice_ppf   <- pmax(0, 200 - 2*wheat_seq)
rice_trade <- pmax(0, 200 - 1.5*wheat_seq)   # P_Wh/P_R = 1.5 (was slope -1 — fixed)
rice_ic1   <- 5000 / wheat_seq               # U = sqrt(Wh*R), U1^2 = 50*100 = 5000
rice_ic2   <- 6700 / wheat_seq               # U2^2 = 67*100 = 6700 (E2 = (67,100) on IC2 & trade line)

df <- data.frame(
  wheat = wheat_seq,
  rice_ppf   = rice_ppf,
  rice_trade = rice_trade,
  rice_ic1   = rice_ic1,
  rice_ic2   = rice_ic2
)

ggplot(df, aes(x=wheat)) +
  geom_line(aes(y=rice_ppf,   color="PPF"),                   linewidth=1.5) +
  geom_line(aes(y=rice_trade, color="Trade Line"),             linewidth=1.5, linetype="dashed") +
  geom_line(aes(y=rice_ic1,   color="IC1 (autarky)"),          linewidth=1,   linetype="dotted") +
  geom_line(aes(y=rice_ic2,   color="IC2 (free trade)"),       linewidth=1,   linetype="dotted") +
  geom_point(aes(x=50,  y=100), color="#012169", size=3.5) +
  geom_point(aes(x=67,  y=100), color="#B9975B", size=3.5) +
  annotate("text", x=41, y=97,  label="E1\n(autarky)",   size=3, hjust=0) +
  annotate("text", x=69, y=106, label="E2\n(free trade)", size=3, hjust=0) +
  scale_color_manual(values=c(
    "PPF"             = "#012169",
    "Trade Line"      = "#B9975B",
    "IC1 (autarky)"   = "grey50",
    "IC2 (free trade)"= "darkgreen"
  )) +
  scale_y_continuous(limits=c(0, 250)) +
  scale_x_continuous(limits=c(0, 105)) +
  labs(title="Trade allows India to reach higher IC (IC2 > IC1)",
       x="Wheat", y="Rice", color="") +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```
:::
:::

## Utility Maximisation Under Trade

**Formal optimisation:**

India maximises $U(R, Wh)$ subject to the trade budget constraint:

$$\max_{R,\, Wh} \; U(R, Wh) \quad \text{s.t.} \quad P_R \cdot R + P_{Wh} \cdot Wh = P_R \cdot \bar{R}$$

where $\bar{R} = 200$ (full specialisation output).

**First-order condition:**

$$\frac{MU_R}{MU_{Wh}} = \frac{P_R}{P_{Wh}} = \frac{1}{1.5}$$

**Separation of production and consumption decisions:**

Under free trade, India:

1. **Produces** where $\text{slope of PPF} = P_R/P_{Wh}$ → full specialisation in Rice
2. **Consumes** where $\text{MRS} = P_R/P_{Wh}$ → tangency of IC with trade line

These two points generally differ — the gap is filled by **trade flows**.

## Key Assumptions of the Ricardian Model

| Assumption | Implication |
|---|---|
| One factor (labour) | Ignores capital, land |
| Constant returns to scale | PPF is linear |
| Perfect factor mobility within country | Instantaneous reallocation |
| Immobile factors across countries | No international migration |
| No transport costs | TOT fully determines trade |
| Perfect competition | No strategic behaviour |

:::{.notes}
**Which matter most?**

- Relaxing *constant returns* → curved PPF, partial specialisation
- Relaxing *one factor* → Heckscher-Ohlin model (Lecture 5)
- Adding *transport costs* → non-traded goods, border effects
:::

## The Ricardian Model: What It Gets Right

**Remarkably robust predictions:**

- Countries do specialise along lines of comparative advantage
- Trade raises real incomes on average
- Even the "least competitive" country has something to export
- Cannot explain intra-industry trade (→ new trade theory, Lecture 6)
- **Extensions:** Many goods (Dornbusch-Fischer-Samuelson chain); technology changes shift comparative advantage over time

:::{.notes}
**Empirical support:** Golub & Hsieh (2000) and Costinot et al. (2012) find strong evidence that labour productivity differences predict export patterns across countries. Predicts complete specialisation (rarely observed); says nothing about *who* within a country gains or loses.
:::

## Measuring Comparative Advantage: The Balassa Index

The **Revealed Comparative Advantage (RCA)** index (Balassa, 1965):

$$RCA_{ij} = \frac{X_{ij} / X_{i \cdot}}{X_{\cdot j} / X_{\cdot \cdot}}$$

where:

- $X_{ij}$ = country $i$'s exports of good $j$
- $X_{i\cdot}$ = country $i$'s total exports
- $X_{\cdot j}$ = world exports of good $j$
- $X_{\cdot\cdot}$ = world total exports

**Interpretation:**

- $RCA_{ij} > 1$: country $i$ has a comparative advantage in good $j$
- $RCA_{ij} < 1$: country $i$ has a comparative disadvantage in good $j$

## India's RCA in Agricultural Products

```{r}
#| label: fig-rca
#| fig-cap: "India: Revealed Comparative Advantage in Agricultural Products (FY2024)\nSource: Author's calculations using DGCI&S / UN Comtrade data."
#| fig-width: 8
#| fig-height: 5
library(ggplot2)

rca_data <- data.frame(
  commodity = c("Spices", "Rice", "Oil Meals", "Cotton",
                "Marine\nProducts", "Fruits &\nVegetables",
                "Processed\nFood", "Edible Oil"),
  rca  = c(5.1, 4.8, 3.8, 3.2, 2.9, 1.4, 0.9, 0.3),
  type = c("Comp. Adv.", "Comp. Adv.", "Comp. Adv.", "Comp. Adv.",
           "Comp. Adv.", "Comp. Adv.", "No Adv.", "No Adv.")
)
rca_data$commodity <- factor(rca_data$commodity,
                              levels=rca_data$commodity[order(rca_data$rca)])

ggplot(rca_data, aes(x=commodity, y=rca, fill=type)) +
  geom_col(width=0.7) +
  geom_hline(yintercept=1, linetype="dashed", color="red", linewidth=0.8) +
  annotate("text", x=0.8, y=1.18, label="RCA = 1 (threshold)", color="red", size=3.5) +
  scale_fill_manual(values=c("Comp. Adv."="#012169", "No Adv."="#cccccc")) +
  coord_flip() +
  labs(title="India's Revealed Comparative Advantage (Balassa Index)",
       subtitle="RCA > 1 = comparative advantage; FY2024 DGCI&S data",
       x=NULL, y="RCA Index (Balassa)", fill=NULL) +
  theme_minimal(base_size=11) +
  theme(legend.position="bottom")
```

## India: Rice as the Flagship Export

**India's Rice sector — Ricardo in practice:**

- India is the **world's largest rice exporter** (>22% of global exports, FY2024)
- Rice cultivation is labour-intensive — India's abundant factor
- **Relative** labour productivity in rice > relative labour productivity in capital-intensive goods

**Policy tension:** September 2023: India imposed a **25% export duty** on non-Basmati white rice → global rice prices spiked 15%.

$$P_R^{\text{world}} \uparrow \Rightarrow \text{India's ToT deteriorate for future exporters}$$

::: {.methodbox}
The ban is welfare-reducing under comparative advantage theory: India surrenders gains from trade to achieve food security goals. **Central tension:** Static efficiency ↔ food security; comparative advantage ↔ strategic reserves; export competitiveness ↔ domestic price stability.
:::

## Comparative Advantage: Summary {.smaller}

**Core results of the Ricardian Model:**

1. Trade is driven by **relative**, not absolute, productivity differences
2. Comparative advantage condition: $\frac{a_{LR}^{I}}{a_{LWh}^{I}} < \frac{a_{LR}^{W}}{a_{LWh}^{W}}$
3. Gains from trade: consumption bundle moves **beyond the PPF**
4. Terms of trade must lie between autarky opportunity costs: $OC_R^W < \frac{P_R}{P_{Wh}} < OC_R^I$
5. Both countries gain — even the absolutely less productive one

**Empirical verdict:** Labour productivity differences strongly predict export patterns — Costinot et al. (2012) find R² ≈ 0.9 across countries. India's rice exports (RCA = 4.8) confirm comparative advantage. Complete specialisation is rarely observed in practice, and short-run distributional losses are documented — but aggregate welfare gains are robust.

**What Ricardo cannot explain:** *Why* labour productivity differs (→ H-O model); income distribution effects (→ Stolper-Samuelson); intra-industry trade (→ Lecture 6)

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

**Further Reading**

- *International Economics* — Salvatore (Ch. 3)
- *International Economics* — Appleyard & Field (Ch. 3)
- RBI/DGCI&S/APEDA databases for latest data

**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
