---
title: "Lecture 17: Agricultural Export Promotion Organizations"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "August 15, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

# {.section}

::: {style="text-align:center; margin-top: 2em;"}
**Lecture 17**

[Agricultural Export Promotion Organizations]{.hi style="font-size:1.8em;"}

The Institutions Behind India's $43.7 Billion Agricultural Export Ecosystem
:::

---

## The Institutional Architecture of India's Agri Exports

::: {.columns}
::: {.column width="58%"}
India's agricultural export performance does not emerge spontaneously from market forces. It is shaped by a **dense ecosystem of statutory bodies, commodity boards, and promotional agencies** working in parallel.

**Three tiers of the ecosystem:**

**Tier 1 — Apex Mandate Bodies:**  
APEDA and MPEDA cover the broadest range of products

**Tier 2 — Commodity-Specific Boards:**  
Tea, Coffee, Rubber, Spices, Tobacco Boards target specific value chains

**Tier 3 — Quality and Finance Infrastructure:**  
EIC (certification), EXIM Bank (trade finance), ECGC (risk cover)
:::
::: {.column width="42%"}
::: {.highlightbox}
**India's agri export organizations at a glance:**

| Body | Est. | Ministry |
|---|---|---|
| APEDA | 1985 | Commerce |
| MPEDA | 1972 | Commerce |
| Tea Board | 1953 | Commerce |
| Coffee Board | 1942 | Commerce |
| Spices Board | 1987 | Commerce |
| Rubber Board | 1947 | Commerce |
| Tobacco Board | 1976 | Commerce |
| EIC | 1964 | Commerce |
| EXIM Bank | 1982 | Finance |
| ECGC | 1957 | Finance |
:::
:::
:::

---

## APEDA: Mandate and Structure

::: {.columns}
::: {.column width="55%"}
**APEDA** — Agricultural and Processed Food Products Export Development Authority

- Established under the **APEDA Act, 1985**
- Under the **Ministry of Commerce and Industry**
- Headquarters: **New Delhi**; 13 regional offices across India

**Scheduled products:**

APEDA covers **14 scheduled product groups** including fresh produce, processed foods, meat, dairy, cereals, and marine products (see APEDA Act Schedule).

::: {.keybox}
APEDA's scheduled products cover **~60% of India's agricultural export basket by value.** Rice (basmati) alone accounts for $5.8B of APEDA's promoted exports.
:::
:::
::: {.column width="45%"}
::: {.methodbox}
**APEDA's governing structure:**

- **Chairperson:** IAS officer (Secretary-level appointment)
- **Governing Board:** 25 members — government, industry, research, farmers
- **Annual budget:** ~₹650 crore (FY2025)

**Funding mechanism:**

APEDA collects a **cess** on APEDA-scheduled exports (₹2 per quintal for most commodities) which funds Market Development Assistance (MDA) grants for trade fair participation, pack houses, and quality testing.
:::
:::
:::

:::{.notes}
Full list of APEDA's 14 scheduled product groups:

Fresh fruits and vegetables · Processed fruits and vegetables · Meat and poultry products · Dairy products · Confectionery and bakery products · Cereals and cereal products · Groundnut, oilseeds, oil meals · Floriculture and seeds · Plantation crops · Organic products · Alcoholic beverages · Poultry and poultry products · Animal casings · Miscellaneous processed items
:::

---

## APEDA Scheduled Product Exports: FY2024

```{r}
#| label: fig-apeda-products
#| fig-cap: "Processed food is the single largest APEDA export category; marine products rank second. Rice (basmati + non-basmati) together constitute ~28% of the basket.\nSource: APEDA Annual Report 2023-24."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(dplyr)

df_apeda <- data.frame(
  Product = c("Non-Basmati Rice","Basmati Rice","Wheat","Other Cereals",
              "Processed Food","Marine Products","Spices",
              "Buffalo Meat","Fresh Fruits/Veg","Other"),
  Value_FY2024 = c(6.4, 5.8, 0.7, 0.5, 11.2, 7.6, 3.7, 3.9, 3.1, 0.8)
) |>
  dplyr::arrange(Value_FY2024) |>
  dplyr::mutate(Product = factor(Product, levels = Product),
                Rank    = dplyr::row_number())

ggplot(df_apeda, aes(x = Value_FY2024, y = Product, fill = Rank)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", Value_FY2024, "B")), hjust = -0.12, size = 3.2, colour = "grey25") +
  scale_fill_gradient(low = "#a8c4e5", high = "#012169") +
  scale_x_continuous(limits = c(0, 14), labels = function(x) paste0("$", x, "B")) +
  labs(title    = "APEDA Scheduled Product Exports: FY2024 (USD Billion)",
       subtitle = "Total APEDA exports = ~USD 43.7 Billion",
       x = "Export Value (USD Billion)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

---

## APEDA's Key Functions in Practice

::: {.columns}
::: {.column width="50%"}
**1. Financial Assistance (MDA Grants):**

- **Infrastructure:** pack houses, pre-cooling facilities, CA stores, vapour heat treatment (VHT) plants — subsidy up to 50% of cost (cap ₹50 lakh–₹3 crore)
- **Quality labs:** NABL accreditation support; pesticide residue testing equipment
- **Market promotion:** 75% reimbursement of stall cost at international fairs

**2. Quality Standards:**

- Issues guidelines for pesticide residue management (grape, mango protocols)
- Coordinates NABL-accredited testing laboratories for certifying shipments
- Manages **Generic India brand** (Incredible India! branding at trade fairs)
:::
::: {.column width="50%"}
**4. Bilateral Market Access Protocols:**

Market access for perishables requires country-to-country pest risk assessment (PRA). APEDA negotiates these protocols:

| Protocol | Country | Product |
|---|---|---|
| VHT Protocol | **Japan** | Mango (Alphonso) |
| Cold Treatment | **South Korea** | Grapes |
| Systems Approach | **USA** | Mangoes |
| Residue Monitoring | **EU** | Grapes, pomegranate |

::: {.highlightbox}
**The Japan mango success story:**  
APEDA negotiated a **vapour heat treatment (VHT) protocol** with Japan's Ministry of Agriculture (MAFF) in 2006. Indian mango exports to Japan grew from near-zero to **$50+ million** in FY2024.
:::
:::
:::

:::{.notes}
**3. Market Intelligence:**

- APEDA's AgriXchange portal: real-time export statistics, market prices, tariff information for 50+ destinations
- Issues market intelligence reports on key commodities and destinations
- Organises buyer-seller meets (BSMs) in India and abroad
- Publishes commodity-specific export guidelines and market reports
:::

---

## Top 10 Destinations for India's Agricultural Exports

```{r}
#| label: fig-agri-destinations
#| fig-cap: "USA leads with 11.2% share driven by marine products and spices; Bangladesh and UAE are the next largest markets for rice and meat exports.\nSource: APEDA Annual Report 2023-24."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(dplyr)

df_dest <- data.frame(
  Country   = c("USA","Bangladesh","UAE","China","Saudi Arabia",
                "Indonesia","Malaysia","Nepal","Vietnam","UK"),
  Share_Pct = c(11.2, 9.8, 8.4, 6.7, 5.3, 4.8, 3.9, 3.7, 3.2, 2.8)
) |>
  dplyr::arrange(Share_Pct) |>
  dplyr::mutate(Country = factor(Country, levels = Country))

ggplot(df_dest, aes(x = Share_Pct, y = Country)) +
  geom_segment(aes(x = 0, xend = Share_Pct, yend = Country),
               colour = "#012169", linewidth = 1.15) +
  geom_point(colour = "#B9975B", size = 5) +
  geom_text(aes(label = paste0(Share_Pct, "%")), hjust = -0.45, size = 3.2, colour = "grey25") +
  scale_x_continuous(limits = c(0, 14), labels = function(x) paste0(x, "%")) +
  labs(title    = "Top 10 Destinations for India's Agricultural Exports (FY2024)",
       subtitle = "Share of total agri exports (%)",
       x = "Export Share (%)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

---

## APEDA's Success Metrics: Selected Achievements

::: {.columns}
::: {.column width="55%"}
**Basmati Rice:**

- FY2010 export: $1.5 billion  
- FY2024 export: **$5.8 billion** (50+ destinations)
- APEDA's GI application for Basmati at WTO (ongoing): protects brand value
- APEDA basmati traceability: QR code on bags linking to farm, miller, testing report

**Organic Exports:**

- FY2024: **$1.1 billion** (50+ countries)
- India: 4th largest organic area globally (4.43 million ha, 4.5 million farmers)
- APEDA manages NPOP (National Programme for Organic Production) — certification, accreditation
- USA (37%) and EU (35%) are top organic destinations

**Fresh Grapes (EU):**

- EU had threatened ban due to pesticide residues (2014 crisis)
- APEDA's Grape Net portal: farmer-level residue monitoring of 1.5 lakh acres
- Residue violations reduced from 180 (2013) to 9 (FY2024)
- Grape exports: **$260 million** FY2024; sustained EU market access
:::
::: {.column width="45%"}
::: {.keybox}
**APEDA trade fair presence:**

| Fair | Location | India Stall |
|---|---|---|
| Gulfood | Dubai, Feb | ~400 exhibitors |
| SIAL | Paris, Oct | ~150 exhibitors |
| Annapoorna | Mumbai | Domestic + buyers |
| FMI Show | Chicago | Marine, organic |
| WorldFood | Moscow | Basmati, spices |

APEDA's "India Pavilion" concept — government-funded stall space offered to SME exporters at subsidised rates. In FY2024, APEDA facilitated participation of **3,200 exporters** across 47 international fairs.
:::
:::
:::

---

## India Agricultural Exports: Long-run Trend

```{r}
#| label: fig-apeda-trend
#| fig-cap: "India's APEDA-promoted exports grew 2.3x from FY2014 to the FY2023 peak of 53.2B USD, before declining to 43.7B in FY2024 after non-basmati rice and onion export bans.\nSource: APEDA Annual Report 2023-24."
#| warning: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(dplyr)

df_trend <- data.frame(
  Year_n    = 1:11,
  Year_lbl  = c("FY2014","FY2015","FY2016","FY2017","FY2018",
               "FY2019","FY2020","FY2021","FY2022","FY2023","FY2024"),
  Exports_B = c(23.2, 23.7, 22.8, 24.1, 26.1, 32.8, 35.1, 41.3, 50.2, 53.2, 43.7)
)

ggplot(df_trend, aes(x = Year_n, y = Exports_B)) +
  geom_area(fill = "#012169", alpha = 0.13) +
  geom_line(colour = "#012169", linewidth = 1.3) +
  geom_point(colour = "#012169", size = 3) +
  geom_point(data = dplyr::filter(df_trend, Year_lbl == "FY2023"),
             colour = "#C84B31", size = 6, shape = 1, stroke = 2) +
  annotate("text", x = 10, y = 56,
           label = "FY2023 peak: $53.2B",
           colour = "#C84B31", size = 3.2, hjust = 0.5, fontface = "bold") +
  annotate("segment", x = 11, xend = 11, y = 46, yend = 44.8,
           colour = "#B9975B", linewidth = 0.8,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed")) +
  annotate("text", x = 10.4, y = 48.5,
           label = "FY2024: $43.7B\nRice & onion export bans",
           colour = "#B9975B", size = 2.9, hjust = 0.5) +
  scale_x_continuous(breaks = 1:11, labels = df_trend$Year_lbl) +
  scale_y_continuous(limits = c(18, 60), labels = function(x) paste0("$", x, "B")) +
  labs(title    = "India Agricultural Exports: Long-run Trend (USD Billion)",
       subtitle = "APEDA-promoted product exports, FY2014–FY2024",
       x = NULL, y = "USD Billion") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 35, hjust = 1))
```

---

## MPEDA: India's Seafood Export Authority

::: {.columns}
::: {.column width="55%"}
**MPEDA** — Marine Products Export Development Authority

- Established under **MPEDA Act, 1972**
- Ministry of Commerce and Industry
- Headquarters: **Kochi, Kerala**; offices in all coastal states

**India's seafood exports (FY2024):**

- Total: **$7.6 billion** (3.5 million MT)
- India: **4th largest** seafood exporter globally (after China, Norway, Vietnam)
- Shrimp: 66% of export value
- Top markets: USA (24%), EU (22%), China (14%), Japan (8%), Southeast Asia (7%)

**Product mix:**

Frozen shrimp · Dried fish · Frozen fish (ribbonfish, pomfret) · Cephalopods (squid, cuttlefish, octopus) · Fresh/chilled fish · Seaweed · Surimi

::: {.keybox}
**Andhra Pradesh** contributes **70% of India's shrimp exports** — the largest concentration of aquaculture in the world outside China.
:::
:::
::: {.column width="45%"}
::: {.methodbox}
**MPEDA's registration system:**

All marine product exporters must register with MPEDA. As of FY2024:

- Processing plants: **1,200+** (EU-approved: 350+)
- Hatcheries: **2,800**
- Aquaculture farmers linked: **~1.2 million**

**Traceability and CoC (Chain of Custody):**

MPEDA's CoC system tracks shrimp from farm pond to export container:

1. Farm registration (GPS-tagged)
2. Feed purchase records (antibiotic-free)
3. Harvest sampling (NRCP residue test)
4. Processing plant records
5. Container ID linked to farm batch

This traceability is required for EU and USA market access — buyers can scan a QR code and trace any pack of Indian shrimp to the source farm.
:::
:::
:::

---

## MPEDA: Quality Systems and International Compliance

::: {.columns}
::: {.column width="55%"}
**National Residue Control Plan (NRCP):**

The EU requires all third-country seafood exporters to operate an NRCP — a systematic national programme testing for:

- Antibiotics (nitrofurans, chloramphenicol, tetracyclines)
- Heavy metals (mercury, lead, cadmium)
- Pesticide residues

MPEDA coordinates NRCP testing across states:
- 18,000+ samples per year
- Tested at CIFT, EIC, and NABL labs
- Annual audit by EU's DG-SANTE (Health and Food Safety)

**Result:** EU has kept Indian seafood on its approved list continuously since 2006 (minor suspension in 2002 was reversed after MPEDA reforms).
:::
::: {.column width="45%"}
::: {.highlightbox}
**USA FDA import alerts — India's challenge:**

The USA FDA has issued import alerts on Indian shrimp over antibiotic residues multiple times.

**2022 import alert:** 46 Indian processing plants placed under increased surveillance; exports to USA dipped 8% in FY2023.

**MPEDA's response:**
- Mandatory antibiotic-free certification for all USA-bound shrimp
- MPEDA training for 50,000 Andhra farmers on prohibited antibiotics
- 3rd party testing at NABL labs (EIC, SGS) before shipment

**Outcome:** USA-bound exports recovered to $1.8B in FY2024.

::: {.keybox}
SPS compliance is not a one-time event — it requires continuous investment in testing, farmer training, and government–industry coordination.
:::
:::
:::
:::

---

## The Five Commodity Boards: Overview

| Board | Est. | HQ | Key Products | FY2024 Export | Top Markets |
|---|---|---|---|---|---|
| **Tea Board** | 1953 | Kolkata | CTC tea, orthodox, green | ~$750M | Russia, UAE, UK |
| **Coffee Board** | 1942 | Bengaluru | Arabica, Robusta | ~$1.2B | Italy, Germany, USA |
| **Rubber Board** | 1947 | Kottayam | NR, latex, crepe | ~$250M | Germany, USA, Italy |
| **Spices Board** | 1987 | Kochi | Chilli, cumin, cardamom, pepper | ~$3.7B | China, USA, Vietnam |
| **Tobacco Board** | 1976 | Guntur | FCV tobacco, cheroots | ~$800M | Belgium, Germany, Russia |

::: {.methodbox}
All five boards perform similar functions: **research, quality standards, market development, financial assistance, and export promotion.** Each has a statutory role in the respective value chain — from regulating auctions (Tea, Tobacco) to issuing GI certificates (Spices Board for Alleppey Green Cardamom).
:::

---

## Tea Board of India

::: {.columns}
::: {.column width="55%"}
**India's tea sector:**

- **2nd largest producer globally** (after China): 1.35 billion kg/year (FY2024)
- Domestic consumption: ~1.1 billion kg (India is its own largest consumer)
- Exports: **200–250 million kg (~$750–800 million)**

**Tea Board functions:**

- Regulates tea auctions: **Kolkata, Guwahati, Coimbatore, Cochin** auction centres
- Tea Research Association (Tocklai, Assam): oldest tropical crop research station in the world (est. 1911)
- Issues **Tea Certificate** for all shipments
- Darjeeling GI: **world's first GI-tagged tea** (EU recognition 1995, India 2004). 87 registered gardens. ~7 million kg/year.

**Top export markets:** Russia (20%), UAE (18%), UK (12%), USA (7%), Germany (5%)
:::
::: {.column width="45%"}
::: {.highlightbox}
**Challenges facing Indian tea:**

1. **Competition:** Kenya (low-cost CTC), China (premium green/white), Sri Lanka (strong Ceylon brand)
2. **Price stagnation:** Assam CTC auction prices have grown only 4% annually since 2015 despite input cost inflation
3. **Climate risk:** North Bengal tea gardens face erratic monsoon; 2023 drought cut Darjeeling first-flush yield by 30%
4. **Small-tea growers:** 35% of tea production from 200,000+ small growers (< 10 ha) — difficult to comply with Rainforest Alliance/UTZ standards required by European retail chains

**Tea Board response:** Grameen Krishi Mitra (GKM) extension workers linking small growers to certification. Target: 50% certified by 2028.
:::
:::
:::

---

## Coffee Board and Spices Board

::: {.columns}
::: {.column width="50%"}
**Coffee Board of India (est. 1942, Bengaluru):**

- India: **3rd largest Arabica producer in Asia**
- Production: ~350,000 MT/year
- Exports: **$1.2 billion (FY2024)** — record high; 15% growth over FY2023
- [Arabica]{.hi}: Coorg, Chikmagalur (Bababudangiri GI); shade-grown, organic premium
- [Robusta]{.hi}: Wayanad, Hassan; used in espresso blends globally
- Top markets: Italy (30%), Germany (18%), Belgium (12%), USA (9%)
- CCRI (Central Coffee Research Institute, Chikmagalur): disease-resistant variety development
- Specialty coffee promotion: India now at World Barista Championship; "Indian single origin" brand growing

::: {.keybox}
Coffee Board's **Café Coffee Day** predecessor was a Board initiative. Domestic coffee consumption now growing 7% annually — youngest coffee drinking demographic globally.
:::
:::
::: {.column width="50%"}
**Spices Board of India (est. 1987, Kochi):**

- India: **world's largest spice exporter by value**
- 109 varieties; 8 million MT production; **4.5 million MT exports**
- **$3.7 billion (FY2024)** — 5-year CAGR of 12%

**Top export items (FY2024):**

| Spice | Value |
|---|---|
| Chilli | $1.1B |
| Spice Oils & Oleoresins | $800M |
| Cumin | $720M |
| Turmeric | $200M |
| Cardamom | $150M |
| Pepper | $140M |

::: {.highlightbox}
**GI products:** Alleppey Green Cardamom, Malabar Black Pepper, Guntur Sannam Chilli, Coorg Orange Cardamom. Premium over non-GI: **20–40%** in EU market.

**SPS challenge:** EU raised MRL concerns on Indian chilli (Sudan Red dye, aflatoxin). Spices Board's GAP training programme covers 500,000 chilli farmers in Andhra Pradesh.
:::
:::
:::

---

## Export Inspection Council (EIC)

::: {.columns}
::: {.column width="55%"}
**EIC** — statutory body under the **Export (Quality Control and Inspection) Act, 1963**

- Ministry of Commerce and Industry; HQ New Delhi
- Purpose: independent certification that Indian exports meet quality standards acceptable to importing countries

**Structure:**

- **5 regional EIOs** (Export Inspection Offices): Mumbai, Chennai, Kochi, Kolkata, Delhi
- **40+ sub-offices** in export-intensive centres (Vizag, Surat, Amritsar, Ludhiana)
- Accredited by **EU's DG-SANTE, USFDA, Japan MHLW, Korean MFDS**

**Mandatory certification products:**

Fish and fishery products · Meat and meat products · Poultry products · Dairy products · Egg products · Honey · Organic products · Castor oil · Processed food (for EU, Japan, USA)
:::
::: {.column width="45%"}
::: {.methodbox}
**What EIC does per shipment:**

1. Document verification (invoice, packing list, test reports)
2. Physical inspection: random sampling from lot
3. Lab analysis: microbiological, chemical, pesticide, heavy metals
4. Issuance of **Health Certificate** (format accepted by destination country)
5. Post-shipment surveillance: tracks rejection reports from destination

**EIC performance (FY2024):**

- Certificates issued: **1.8 lakh**
- Value covered: **₹1.1 lakh crore**
- Rejection rate (border rejections at EU/USA): **0.4%** (one of the lowest among major exporters)
- Turnaround time: **2–3 working days** for most certificates

::: {.keybox}
An **FSSAI-EIC joint protocol** (2022) has harmonised testing requirements so exporters need not test separately for domestic FSSAI and export EIC compliance.
:::
:::
:::
:::

---

## NAFED: Price Stabilisation and Export Implications

::: {.columns}
::: {.column width="55%"}
**NAFED** — National Agricultural Cooperative Marketing Federation of India

- Established 1958; cooperative society under Multi-State Cooperative Societies Act
- Functions: **Price Support Scheme (PSS) procurement**, buffer stock management, retail distribution

**PSS and its export connection:**

When prices of oilseeds (mustard, groundnut, soybean) and pulses fall below MSP:
- Government activates PSS; NAFED procures from farmers at MSP
- NAFED accumulates buffer stocks of oilseeds/pulses
- Stocks are sold through NAFED retail, e-commerce, and open market sales when prices rise
- Procurement **reduces export availability** (absorbs domestic surplus before it can be exported); but **stabilises farm incomes**

::: {.keybox}
In kharif 2023: NAFED procured **38 lakh MT of tur dal** at MSP (₹6,600/quintal) from Karnataka, Maharashtra, Madhya Pradesh. This stock was later released to control prices — but reduced the quantity available for the export market in FY2024.
:::
:::
::: {.column width="45%"}
::: {.highlightbox}
**The NAFED export dilemma:**

NAFED is **not primarily an export organization** — it is a price stabilisation agency. But its procurement and buffer management decisions directly affect India's agri export supply.

**Tension:** When NAFED holds large tur dal stocks, export prices are firm because domestic supply is constrained. When NAFED releases stocks, domestic prices fall → farmer incomes hurt, but consumers benefit.

**FPO linkage:**

NAFED's eMandis portal now links FPOs to government procurement at MSP and to private buyers. 1,200 FPOs linked as of FY2024, aggregating procurement from 2.5 million farmers.

**AAPL (NAFED's retail arm):** sells agri commodities B2C through 1,200+ stores, Amazon/Flipkart — brand building for Indian agri.
:::
:::
:::

---

## Farmer Producer Organizations as Export Units

::: {.columns}
::: {.column width="55%"}
**The small-farm problem:**

India's average farm size: **1.08 hectares** (Agriculture Census 2015–16).  
Average EU/USA buyer minimum order: **20 MT per lot.**

No individual smallholder can meet buyer requirements for:
- Consistent volume
- Uniform quality and grading
- Traceable production records
- Third-party certification costs (organic, GlobalGAP)

**FPOs as aggregators:**

An FPO with 500 members × 1 ha each = 500 ha aggregate. Can produce 300–500 MT of a single crop — meeting buyer minimums.
:::
::: {.column width="45%"}
::: {.methodbox}
**Successful FPO export examples:**

**Mahagrapes (Maharashtra):**  
Cooperative of 46 grape-growing cooperatives, ~25,000 farmers.  
Exports EU-compliant grapes to Netherlands, Germany, UK.  
FY2024 exports: ~8,000 MT (~$40M).  
Uses APEDA-funded pack houses and cold chain in Nashik.

**Sahyadri Farms (Nashik):**  
12,000 farmer members; India's largest single FPO.  
GlobalGAP certified; supplies Tesco UK, Carrefour.  
Own cold chain, pack house, CA storage.

**NDDB dairy cooperatives:**  
Amul (GCMMF): 3.6 million farmer members; dairy exports to 50+ countries ~$500M.

::: {.keybox}
**APEDA's FPO Connect:** 1,200+ FPOs registered, linked to APEDA's buyer database and MDA grant system. FPOs can now access MDA grants directly — not just through large exporters.
:::
:::
:::
:::

---

## Government Schemes Supporting Agri Export Organizations

::: {.columns}
::: {.column width="50%"}
**PM FME (Formalisation of Micro Food Enterprises) Scheme:**
- ₹10,000 crore over 5 years (2020–25)
- Credit-linked subsidy (35%) for micro-enterprises upgrading to FSSAI standards
- ODOP (One District One Product) focus products receive priority
- 2 lakh+ enterprises formalized by FY2024 → many now eligible for export

**NABARD's Agri Infrastructure Fund (AIF):**
- ₹1 lakh crore; 3% interest subvention
- Pack houses, cold stores, CA chambers, primary processing centres
- Linked to FPO and cooperative projects — directly supports export infrastructure

**SFAC (Small Farmers Agribusiness Consortium):**
- FPO promotion, equity grant support
- Connects FPOs to APEDA/MPEDA registered exporters
- Pilot: 100 agri export FPOs linked directly to 3 commodity board markets
:::
::: {.column width="50%"}
**APEDA-NHB Horticulture Export Collaboration:**

NHB (National Horticulture Board) funds pack houses; APEDA funds market promotion. Joint program in:
- Banana (Jalgaon, Maharashtra → Middle East)
- Guava (Allahabad, UP → UK diaspora market)
- Litchi (Muzaffarpur, Bihar → EU, West Asia)

::: {.highlightbox}
**Districts as Export Hubs (DEH) — FTP 2023–28:**

Commerce Ministry has identified **100+ districts** with specific "champion products."

Examples:
- Kalaburagi (Karnataka): Tur dal
- Nashik (Maharashtra): Grapes, onions
- Tirupur (Tamil Nadu): Knitwear + agri
- Coorg (Karnataka): Coffee, pepper
- Guntur (Andhra Pradesh): Chilli, turmeric

District Collectors coordinate between farmers, FPOs, exporters, and commodity boards. Target: every DEH district to double export value in 5 years.
:::
:::
:::

---

## India's 2030 Agricultural Export Ambition {.smaller}

**APEDA's target: $60 billion by 2030** (from $43.7B in FY2024; CAGR ~5.4%)

**National agri-food export target (FTP 2023–28): $100 billion by 2030**

**Strategy pillars:**
1. Diversify beyond rice dominance
2. Open new markets: Africa, Southeast Asia, South Korea
3. Value addition: processed, organic, functional foods
4. Invest in cold chain: PM GatiShakti nodes
5. Leverage GI brand premiums

::: {.keybox}
**Key Takeaways:**

1. **APEDA** is the backbone of India's agri export promotion — covers 60% of the export basket
2. **MPEDA** manages India's $7.6B seafood export with rigorous quality systems essential for EU/USA access
3. **Commodity Boards** are sector-specific engines — Spices Board's $3.7B is remarkable for a relatively small value chain
4. **EIC** provides the third-party quality assurance without which EU/USA market access is impossible
5. **FPOs** are solving the small-farm aggregation problem — Mahagrapes, Sahyadri, Amul show it is achievable at scale
:::

:::{.notes}
**Commodity Board targets:** Spices Board: $5B (from $3.7B); Coffee Board: $2B (from $1.2B); Tea Board: $1.5B (from $750M). MPEDA's target: $12B in seafood.

6. **District-level implementation** through DEH is the newest institutional innovation — connecting national targets to local value chains.

FTA market access: India-UAE CEPA live; India-UK FTA being negotiated.
:::

---

## Next Lecture

::: {.methodbox}
**Lecture 18 — India's Agricultural Trade and Foreign Trade Policy**  
*August 22, 2026 — FINAL LECTURE*

- Course synthesis: from comparative advantage to India's $43.7 billion agri export reality
- India's trade performance: FY2024 big picture and 10-year trends
- Foreign Trade Policy 2023–28: vision, themes, and agricultural provisions
- Millets, GIs, and organic exports as strategic niches
- The farmer welfare vs export competitiveness dilemma
:::

:::{.notes}
- Export composition, import profile, and the value-addition opportunity
- Challenges India must overcome and opportunities ahead
- Course synthesis: what 18 weeks has taught us
- India's Agricultural Export Vision 2030

**Final lecture preparation:** Review all lecture notes from Lectures 1–17. Bring questions — we will have a 15-minute open Q&A on examination topics at the end of Lecture 18.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- Lecture notes and APEDA/WTO official documents
- APEDA Annual Report 2023-24
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
