---
title: "Features of International Trade"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "May 2, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    logo: ""
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(scales)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

# Part I: Recap and Overview {background-color="#E8EDF5"}

## Recap: Lecture 1

::: {.methodbox}
**Three things to remember from last week**

1. **International trade** = cross-border exchange of goods, services, and capital; driven by differences in resource endowments, technology, and opportunity costs

2. **India's position:** 18th-largest merchandise exporter; agricultural exports ~\$43.7B (FY2024); structural trade surplus in agriculture of ~\$15.5B

3. **Agricultural trade is special** — food security concerns, perishability, seasonality, and powerful political economy shape how governments intervene
:::

**Today's agenda**

- What distinguishes international trade from *domestic* trade?
- How open is India to global trade — and how has this changed over time?
- Advantages and risks of participating in global markets

## India's Growing Trade Openness

```{r}
#| label: fig-trade-gdp
#| fig-cap: "India: Trade Openness (Exports + Imports as % of GDP)\nSource: World Bank, World Development Indicators (WDI)."
#| fig-width: 7
#| fig-height: 4
library(ggplot2)

openness <- data.frame(
  year = c(1990, 1995, 2000, 2005, 2010, 2015, 2020, 2024),
  trade_gdp = c(14, 22, 27, 35, 48, 42, 38, 44)
)

ggplot(openness, aes(x=year, y=trade_gdp)) +
  geom_line(color="#012169", linewidth=1.5) +
  geom_point(color="#012169", size=3) +
  geom_area(fill="#012169", alpha=0.1) +
  geom_vline(xintercept=1991, linetype="dashed", color="red", linewidth=0.8) +
  annotate("text", x=1993, y=16, label="1991\nLiberalisation", color="red", size=3) +
  labs(title="India's Trade Openness: (Exports + Imports) / GDP",
       subtitle="India has become significantly more open since 1991 liberalisation",
       x=NULL, y="Trade / GDP (%)") +
  scale_x_continuous(breaks=c(1990,1995,2000,2005,2010,2015,2020,2024)) +
  theme_minimal(base_size=11)
```

::: {.keybox}
Trade/GDP rose from **14% (1990)** to **44% (2024)** — India is significantly more integrated into the global economy post-liberalisation.
:::

# Part II: Six Key Features {background-color="#E8EDF5"}

## Features That Make International Trade Distinct {.smaller}

International trade is not simply domestic trade "at a larger scale." Six features create unique challenges:

**(a) Geographical separation** — Long distances → higher transport costs, insurance, cold-chain requirements, longer lead times

**(b) Different currencies** — Each transaction requires currency conversion; exchange rate fluctuations create financial risk

**(c) Factor immobility** — Labour and capital cannot move freely across borders — immigration controls, capital account restrictions

**(d) Multiple political jurisdictions** — Each country sets its own trade policy — tariffs, quotas, export bans; no single governing authority

**(e) Different legal systems & culture** — Food labelling laws, packaging standards, contract enforcement vary widely

:::{.notes}
**(f) Greater risk and uncertainty** — Combination of all above → higher commercial, political, and exchange-rate risk than domestic commerce. Indian spice exporters must comply with EU pesticide MRL standards.
:::

## Feature 1: Factor Immobility

**Why factors don't move freely**

- **Labour:** Visa restrictions, language barriers, immigration quotas
  - India sends ~8 million workers to Gulf countries — under strict bilateral agreements, not free movement
- **Capital:** FDI restrictions, capital account controls, regulatory hurdles
  - India maintains partial capital account convertibility (RBI-regulated)
- **Land:** Completely immobile by definition

**Why this matters for trade theory**

- Classical models (Ricardo, H-O) assume *factor immobility* across borders
- This is what makes trade in goods a *substitute* for factor flows
- If labour moved freely, there would be less need to trade goods

::: {.highlightbox}
**India case: Gulf remittances** — Indian workers (largely from Kerala, UP, Bihar) work in UAE, Saudi Arabia, Kuwait — sending back **~\$120 billion in remittances (FY2024)**, making India the world's largest remittance recipient.
:::

## Feature 2: Currency and Exchange Risk

**The exchange rate problem**

- Indian exporters price in **USD** (world standard for commodities)
- Their costs are in **Indian Rupees (INR)**
- If INR appreciates (e.g., ₹80 → ₹75 per USD), the same USD revenue converts to *fewer rupees* → profit squeezed
- If INR depreciates (₹80 → ₹85), exporters gain; importers lose

| Year | INR/USD | | Year | INR/USD |
|------|---------|--|------|---------|
| 2000 | 45 | | 2020 | 75 |
| 2010 | 46 | | 2024 | 83–84 |
| 2015 | 65 | | | |

::: {.methodbox}
**Agricultural export example:** A Punjab wheat exporter quotes **\$250/tonne** FOB Mumbai. At ₹80/USD → realises ₹20,000/tonne; at ₹75/USD → only ₹18,750/tonne. A **5% INR appreciation = 6.25% reduction** in rupee revenue — potentially wiping out margins. **Hedging tools:** Forward contracts, futures on NSE/BSE, ECGC insurance.
:::

# Part III: Internal vs International Trade {background-color="#E8EDF5"}

## Internal vs International Trade: Key Differences

::: {.smaller}
| Dimension | Internal (Domestic) Trade | International Trade |
|-----------|--------------------------|---------------------|
| **Factor mobility** | Labour and capital move freely | Highly restricted — visas, FDI rules |
| **Currency** | Single national currency | Multiple currencies; exchange rate risk |
| **Political jurisdiction** | One government, one legal system | Multiple governments, international law |
| **Trade barriers** | None within India (post-GST) | Tariffs, quotas, SPS, NTBs |
| **Transport costs** | Relatively low | Higher — ocean freight, insurance, customs |
| **Documentation** | Invoice, transport document | B/L, L/C, Certificate of Origin, SPS certificate |
| **Market size** | Limited to one country | Global — 195 countries |
| **Dispute resolution** | Domestic courts | WTO DSM, bilateral/regional bodies |
:::

::: {.keybox}
**Key insight:** India's internal trade was itself *fragmented* before GST (2017) — state-level taxes created internal barriers. International trade faces all these barriers *and more*.
:::

## The Gravity Model of Trade

**Inspired by Newton's law of gravitation:**

$$T_{ij} = G \cdot \frac{Y_i \cdot Y_j}{D_{ij}^2}$$

Where $T_{ij}$ = trade between $i$ and $j$; $Y_i, Y_j$ = GDP; $D_{ij}$ = distance.

**Interpretation:** Trade is **proportional to economic size** and **inversely proportional to distance**.

| Partner | GDP | Distance | Trade rank |
|---------|-----|----------|------------|
| USA | Very large | Far | #1 export dest. |
| China | Very large | Medium | #1 import source |
| UAE | Medium | Close | #2 export dest. |
| Bangladesh | Small | Very close | #5 export dest. |

::: {.highlightbox}
UAE ranks highly despite moderate GDP: it is a regional trade hub + large Indian diaspora creates demand for Indian food products.
:::

# Part IV: Advantages and Risks {background-color="#E8EDF5"}

## Advantages of International Trade for India

**For producers and the economy**

1. **Wider market** — Indian basmati rice sold in 150+ countries; domestic market alone could not absorb export volumes
2. **Access to unavailable inputs** — India imports potash fertiliser (not produced domestically in sufficient quantity)
3. **Technology transfer** — FDI brings modern food processing technology
4. **Employment generation** — Marine product sector employs ~1.5 million workers, mostly coastal women
5. **Foreign exchange earnings** — Agricultural exports earn \$43.7B; crucial for financing oil imports

::: {.highlightbox}
**Rice example:** India is the [world's largest rice exporter]{.hi} (~22 million tonnes in 2022). Exports earn [~\$10B forex]{.hi-gold} and support 30 million paddy farmers.
:::

:::{.notes}
Additional advantages of international trade for India:

6. **Economic growth** — Export-led growth: every \$1B of agri exports generates ~\$1.5–2B in backward linkages
7. **Better utilisation of resources** — India's abundant agricultural land and labour efficiently deployed through exports
:::

## Disadvantages and Risks of International Trade

**Economic risks**

1. **Import dependence** — India's edible oil dependence (60–65% imported) exposes inflation to global price shocks
2. **Dumping risk** — China dumped cheap steel and tomato paste in India; domestic industries harmed
3. **BoP strain** — India's merchandise deficit ~\$241B (FY2024); requires large capital inflows to finance
4. **Domestic industry harm** — Cheap palm oil lowered demand for domestic groundnut oil

::: {.highlightbox}
**Policy lesson:** The goal is not to maximise trade at all costs, but to *manage* trade so its gains exceed its disruption costs.
:::

:::{.notes}
Political and structural risks:
5. **Economic dependency** — Over-specialisation creates vulnerability (e.g., countries heavily dependent on single commodity exports)
6. **Trade disputes** — India has faced WTO challenges on sugar subsidies, poultry import bans
7. **Structural unemployment** — Import competition can displace workers; adjustment is slow
:::

# Part V: India Case Studies {background-color="#E8EDF5"}

## Case Study: India's Edible Oil Import Dependence

**The structural problem**

- India consumes ~24–25 million tonnes of edible oil per year
- Domestic production: ~9–10 million tonnes
- Import requirement: **~14–15 million tonnes** (60–65% of need)
- Cost: **~\$12–14 billion per year** (FY2024)

| Oil type | Source | Share |
|----------|--------|-------|
| Palm oil | Indonesia, Malaysia | ~65% |
| Soybean oil | Argentina, Brazil | ~15% |
| Sunflower oil | Ukraine, Russia | ~15% |

::: {.methodbox}
**Lesson:** Import dependence in a strategic food commodity creates price transmission vulnerability. PM KISAN Oilseeds Mission (2023) aims to double domestic production by 2030.
:::

:::{.notes}
**The 2021–22 price shock:**
- Russia-Ukraine war → sunflower oil disruption
- Palm oil export ban by Indonesia (April–May 2022)
- Global edible oil prices rose 50–80% in 6 months
- India's retail edible oil inflation peaked at ~20% mid-2022
- Government response: cut import duties, banned futures trading
:::

## Case Study: India's Rice Export Ban (2023)

**What happened:**

- India banned non-basmati white rice exports in August 2023
- Rationale: control domestic food inflation, protect consumers
- India is the world's largest rice exporter — ~22 million tonnes/year

**Who gained:** Urban consumers (domestic prices stabilised); RBI's CPI inflation target

**Who lost:** Paddy farmers (domestic price fell); rice millers and exporters (contracts cancelled); Bangladesh and West Africa (supply shortage; global prices rose +20%)

::: {.highlightbox}
The ban caused a **global rice price spike of ~20%** — demonstrating India's market power as the world's dominant rice exporter. India partially reversed the ban for specific grades in early 2024 under WTO member pressure.

**This is a classic domestic policy vs. trade commitment conflict** — a recurring theme in agricultural trade policy.
:::

## Government Policy in Agricultural Trade

**Price support and export competitiveness**

- **MSP:** Raises domestic price above world price → may make exports uncompetitive without subsidy
- **Export bans:** Rice ban (non-basmati) August 2023; wheat ban (May 2022) — protect domestic prices, hurt exporters
- **Import duties on pulses and oilseeds:** Reduced to zero during 2021–22 price shock; raised again when harvest recovered
- **TRQs:** Limited quantities at low duty; higher duty beyond the quota

::: {.highlightbox}
**Core tension:** India's domestic farm support policies often conflict with WTO commitments on export subsidies and import duties — the AoA bindings create real constraints (Lecture 12).
:::

:::{.notes}
- **Export subsidies:** Previously used for sugar (challenged at WTO); now replaced by transport/marketing assistance
- **PLI scheme for food processing:** Incentivise value-added food exports
:::

## Trade Finance and Documentation

**Key trade documents**

- **Bill of Lading (B/L):** Shipping company's receipt; title document for goods
- **Letter of Credit (L/C):** Bank's guarantee of payment; protects both parties
- **Certificate of Origin:** Proves provenance; needed for preferential duty under FTAs
- **Phytosanitary Certificate:** Certifies plants/plant products are pest-free
- **SPS Compliance Certificate:** Confirms pesticide residues, aflatoxin levels meet standards

::: {.methodbox}
**Why small agri-exporters struggle:** Indian rice exporters to EU must meet **Maximum Residue Levels (MRLs)** for 300+ pesticides. Failure → entire consignment rejected at port. L/C requires strong banking relationships. **ECGC** provides credit insurance — covers non-payment risk for Indian exporters.
:::

## Summary and Key Takeaways

::: {.methodbox}
**What we covered today**

- International trade has **6 key features** distinguishing it from domestic trade: geographic separation, currency risk, factor immobility, multiple jurisdictions, cultural/legal differences, and greater uncertainty
- **India's trade openness** rose from 14% of GDP (1990) to 44% (2024) — driven by 1991 liberalisation
- **Factor immobility** is the fundamental reason countries trade goods rather than simply moving workers
- **Currency risk** directly affects agricultural export profitability — a 5% INR appreciation can wipe out export margins
- **Edible oil dependence** is India's most significant agricultural import vulnerability: \$12–14B/year
:::

:::{.notes}
- **Gravity model** predicts India trades most with large (USA, China) and close (UAE, Bangladesh) economies
:::

## Discussion Question

::: {.highlightbox}
**"India has a comparative advantage in rice. Bangladesh also exports rice. How can both export the same commodity?"**
:::

**Think through:**

- Is India's advantage in *all* varieties of rice?
- Can a country export *and* import the same good? (Intra-industry trade)
- How do quality differentiation, variety, and destination markets matter?
- What does the rice export ban tell us about the limits of comparative advantage as a policy guide?

::: {.methodbox}
**Answer preview:** Intra-industry trade, quality segmentation (basmati vs parboiled), and market access geography all allow two countries to simultaneously export similar products. We formalise this in Lecture 6.
:::

## Next Lecture: Lecture 3

**Theories of International Trade I — Mercantilism and Absolute Advantage**
*(May 12, 2026)*

We will cover:

- **Mercantilism** (1500–1750): wealth = gold, maximise exports, minimise imports — and why it is fundamentally flawed
- **Colonial trade:** India's experience under British mercantilism — the "drain of wealth" thesis (Dadabhai Naoroji)
- **Adam Smith (1776):** Absolute advantage — trade based on genuine productivity differences
- **Numerical example:** India vs Bangladesh in rice and jute
- **Limitations** of absolute advantage — setting up for Ricardo

:::{.notes}
**Suggested reading**

- Krugman, Obstfeld & Melitz, *International Economics*, Ch. 2
- Adam Smith, *Wealth of Nations*, Book IV, Ch. 1 (excerpt)

**To think about:**

If every country has absolute advantage in *something*, does that mean every country should trade? What about a country that has absolute advantage in *nothing*?
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

**Further Reading**

- *International Economics* — Salvatore (Ch. 1)
- *International Economics* — Appleyard & Field (Ch. 1)
- RBI/DGCI&S/APEDA databases for latest data

**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
