---
title: "Lecture 3: Theories of International Trade I"
subtitle: "Econ 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "May 9, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    logo: ""
    footer: "Econ 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(patchwork)
options(scipen=999)
theme_set(theme_minimal(base_size=11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

## The Central Question

**Why do countries trade?**

This question has occupied economists for over 500 years. Three broad answers have emerged:

::: {.fragment}
1. **Mercantilism (1500–1750):** Trade is a zero-sum competition for power; exports = good, imports = bad; accumulate gold
:::

::: {.fragment}
2. **Absolute Advantage — Adam Smith (1776):** Trade is win-win; each country specialises in what it produces *most efficiently*
:::

::: {.fragment}
3. **Comparative Advantage — David Ricardo (1817):** Even if one country is better at *everything*, both countries gain from trade based on *relative* costs
:::

::: {.methodbox}
**Why theory matters for policy:** India bans rice exports → mercantilist instinct. India signs FTAs → comparative advantage logic. These theories are the intellectual foundation of every trade policy debate happening right now.
:::

## Trade theory timeline (I): early foundations

- **Mercantilism (1500–1750):** wealth = gold; maximise trade surplus
- **Smith (1776):** absolute advantage (higher productivity)
- **Ricardo (1817):** comparative advantage (lower opportunity cost)
- **H–O (1933):** trade reflects factor endowments
- **Big shift:** from “win–lose trade” → **gains from specialization**

---

## Trade theory timeline (II): modern extensions

- **Leontief (1953):** data challenge to simple H–O predictions
- **Krugman (1970s–80s):** scale economies + imperfect competition
- **Porter (1990s):** national competitiveness and institutions
- **Melitz (2000s):** firm heterogeneity; only productive firms export
- **Course roadmap:** by Lecture 6 we cover the arc; today starts with mercantilism + Smith

## Mercantilism: the setting

- Dominant European doctrine: **1500–1750**
- Bullion money: gold/silver treated as “wealth”
- Nation-states + colonial expansion + chartered companies

---

## Mercantilism: core beliefs

- Wealth ≈ **gold/silver stock**
- Goal: **export > import** (surplus)
- Trade seen as **zero-sum**
- State actively manages trade (tariffs, subsidies, monopolies)

:::{.notes}
Modern echo: “increase the trade surplus” rhetoric often reflects mercantilist instincts.
:::

## Mercantilist policies (export side)

- Export subsidies / rebates to push exports
- Chartered trading companies (EIC, VOC) with monopoly rights
- Shipping restrictions (Navigation Acts) to control trade routes

---

## Mercantilist policies (import side + currency)

- Import duties / quotas to restrict imports
- State monopolies to control strategic goods
- Currency undervaluation to make exports cheaper

---

## Modern equivalents (examples)

- Export rebates: **RoDTEP** (India)
- Import duties: India’s tariff schedule
- Trading houses: **MMTC/STC** (public trading)
- Undervaluation episode: China (2000s)

::: {.methodbox}
Mercantilism isn’t always “irrational”; the error is the **zero-sum worldview**.
:::

## Colonial trade: British mercantilism in India

- India as **raw material source** + **captive market**
- Cotton chain: India → Manchester mills → textiles back to India
- Trade rules tilted toward Britain (favouring British textiles)

---

## The “drain” idea (why this matters)

- Naoroji: colonial trade + administration created a systematic transfer outward
- Channels: home charges, interest payments, profits remitted
- Big estimates exist (e.g., Patnaik 2018 ≈ $45T PV) — **methodology debated**

---

## Agricultural dimension

- Cash-crop push (indigo/opium/jute) could displace food crops
- Higher food vulnerability + famine risk (Bengal 1943)
- Key point: **institutions + trade policy shape what gets produced**

## The Drain of Wealth: Naoroji's Evidence

**Dadabhai Naoroji's Drain Theory**

- Naoroji (1901): *Poverty and Un-British Rule in India*
- Calculated that Britain extracted **£200–300 million per year** from India as "drain"
- Home charges: India paid for British administration costs, interest on British-held debt, railway profits
- The mechanism: India exported goods but received paper IOUs rather than gold — a systematic wealth transfer disguised as trade

::: {.quotebox}
"England has impoverished India by her own prosperity." — Dadabhai Naoroji, 1901
:::

::: {.methodbox}
**Modern reassessment:** Economists Utsa Patnaik (2018) and Sashi Tharoor (*Inglorious Empire*, 2017) updated Naoroji's estimates. The drain theory remains contested on methodology but its directional conclusion — that colonial trade systematically transferred wealth from India to Britain — is broadly accepted.
:::

## Critique of Mercantilism

**David Hume's price-specie-flow mechanism (1752)**

Mercantilism has a self-defeating logic:

1. Country A runs a trade surplus → gold flows in → money supply rises
2. More money → prices rise (quantity theory of money: MV = PQ)
3. Higher prices → Country A's exports become more expensive
4. Exports fall, imports rise → trade surplus disappears automatically

**Adam Smith's critique (1776):** "Wealth of a nation is not measured by the gold in its vaults but by the goods and services it can produce and consume"; import restrictions reduce consumption possibilities

::: {.methodbox}
**The fundamental error:** Mercantilists confused the **means** (gold) with the **end** (well-being). A country with more gold but fewer goods is *poorer*, not richer.
:::

:::{.notes}
**Modern parallel:** Japan in the 1980s ran huge trade surpluses with the USA. Was Japan richer for it? Its citizens had high savings but *lower consumption* than Americans with similar incomes — a classic mercantilist outcome.
:::

## Mercantilist Arithmetic: Why the Logic Fails {.smaller}

**The mercantilist objective function**

A nation seeks to accumulate gold $G$ by maximising its trade surplus:

$$\text{Objective: } \max\; G = X - M$$

$$\Rightarrow \text{policy prescriptions: } X \uparrow \text{ (export subsidies)}, \quad M \downarrow \text{ (tariffs, prohibitions, bans)}$$

**The zero-sum fallacy:** If *all* $N$ nations simultaneously pursue a surplus:

$$\sum_{i=1}^{N}(X_i - M_i) = 0 \quad \forall \text{ nations simultaneously}$$

**Hume's price-specie-flow refutation (1752):** Even a single nation's surplus is self-correcting:

$$\text{Gold inflow} \;\Rightarrow\; M_s \uparrow \;\Rightarrow\; P \uparrow \;\Rightarrow\; X \downarrow,\; M \uparrow \;\Rightarrow\; \text{surplus eliminated}$$

::: {.methodbox}
A mercantilist surplus is self-defeating: the very gold inflow it generates raises domestic prices, eroding the export competitiveness that created the surplus.
:::

## Smith's Critique: Real Wealth vs Gold

**Smith's fundamental restatement**

Mercantilists measured national wealth by the gold stock. Smith showed the correct measure is real consumption:

$$W_{\text{real}} = \frac{\text{goods and services consumed}}{P} \;\neq\; \text{gold stock}$$

**The trade restriction paradox — applied to India**

If India bans rice exports to keep supply at home:
$$W_{\text{domestic,rice}} \uparrow \quad \text{(rice cheaper domestically)}$$
$$W_{\text{global}} = 0 \quad \text{(no new rice created — redistributed, not added)}$$

If instead India exports rice and specialises:
$$W_{\text{global}} \uparrow \quad \text{(specialisation gains; world output expands)}$$
$$W_{\text{India}} \uparrow \quad \text{(farmers earn more; import cheaper goods)}$$

::: {.highlightbox}
> *"It is the maxim of every prudent master of a family, never to attempt to make at home what it will cost him more to make than to buy."* — Wealth of Nations, Book IV, Ch. 2

National self-sufficiency achieved through restriction destroys, not creates, real wealth.
:::

## Mercantilism vs Smith: Policy Implications

::: {.smaller}
| Dimension | Mercantilism | Adam Smith |
|---|---|---|
| **Goal** | National power / gold accumulation | National prosperity / consumption |
| **View of trade** | Zero-sum: one nation's gain = another's loss | Positive-sum: specialisation grows world output |
| **Role of imports** | Evil — gold drains out | Beneficial — access to cheaper goods |
| **Role of exports** | Good — gold flows in | A means to pay for imports |
| **State role** | Regulate all trade; grant monopolies | Laissez-faire; let markets direct resources |
| **India example (2023)** | Rice/wheat export ban — hoard grain, protect prices | RCEP/FTA participation — specialise, exchange |
:::

::: {.keybox}
**The modern policy echo:** Every time an Indian policymaker imposes an export ban to protect domestic consumers, they invoke mercantilist logic. Every time an economist argues for free trade to raise farmer incomes, they invoke Smith. Both impulses co-exist in India's agricultural trade policy — understanding the theory clarifies which logic applies when.
:::

## Neo-Mercantilism: Export-Led Development Strategies

**Modern mercantilist practices**

- **Export-led growth strategy** (China, South Korea, Germany): Use undervalued currency and export subsidies to drive export surplus; accumulate foreign exchange reserves
- **Currency undervaluation:** China's renminbi was widely believed to be 20–40% undervalued vs USD in 2000s
- **"Beggar-thy-neighbour" policies:** Competitive devaluations — my currency devaluation boosts my exports at *your* expense
- **Industrial policy for export champions:** China's \$5 trillion in state subsidies to EVs, solar panels (WTO disputes pending)
- **India's FTP 2023–28 target** of \$2 trillion exports has a mercantilist *flavour* — emphasises export growth

:::{.notes}
**The "China is mercantilist" debate:** China kept RMB undervalued by ~20–40%, providing massive subsidies to export industries → \$3T+ in foreign exchange reserves. Critics: unfair competition. China's response: infant industry protection, just like US and UK in 19th century. **Who is right?** Both, somewhat.
:::

## Neo-Mercantilism: Critique and Assessment

::: {.keybox}
**Critical Analysis of Neo-Mercantilist Strategies**

While export-led growth produced remarkable results in East Asia (South Korea, Taiwan, China), neo-mercantilism has fundamental limitations:

- **Not universally replicable:** If all developing countries simultaneously pursue export surpluses, not all can succeed — the global sum of surpluses must be zero
- **Trading partner retaliation:** Persistent surpluses invite countervailing duties, anti-dumping actions, and trade wars (e.g., US-China trade war, 2018–present)
- **Domestic welfare cost:** Currency undervaluation raises the cost of imports for domestic consumers; workers in export industries may face poor conditions
- **Bretton Woods system collapse (1971):** The original system broke down partly because of mercantilist pressures from surplus countries

**For India:** The challenge is to use strategic export promotion (FTP, APEDA, SEZs) without triggering WTO violations or partner retaliation.
:::

## Adam Smith's Insight

**The context: Wealth of Nations (1776)**

- Published the same year as the American Declaration of Independence
- A direct intellectual attack on mercantilism and the colonial trade system
- Argued for **free markets** and **free trade** as the sources of national wealth

> *"If a foreign country can supply us with a commodity cheaper than we ourselves can make it, better buy it of them with some part of the produce of our own industry, employed in a way in which we have some advantage."*
> — Adam Smith, *Wealth of Nations*, Book IV, Ch. 2

**Smith's core insight:** Trade is **not zero-sum** — both countries gain. Each country should produce what it is **best at** — *absolute advantage*. The standard of living rises when countries specialise and exchange.

::: {.methodbox}
**Absolute advantage:** Country A has an absolute advantage in producing good X if it can produce **more units of X per unit of labour** than Country B.
:::

## Assumptions of the Absolute Advantage Model {.smaller}

| Assumption | Implication |
|---|---|
| Two countries (A, B) | Tractable bilateral comparison |
| Two goods (X, Y) | One good per country's advantage |
| One factor — labour | Productivity differences drive trade |
| Labour mobile *within* country | Workers shift between sectors freely |
| Labour immobile *across* countries | No international migration |
| Constant returns to scale | Linear PPF; constant opportunity costs |
| Perfect competition | Prices equal marginal (labour) costs |
| Full employment | All labour is always in use |
| No transport costs | World price fully determines trade |

::: {.methodbox}
**Key limitation:** If one country is better at *both* goods, the model predicts *no* trade. Ricardo's comparative advantage (Lecture 4) resolves this.
:::

## Absolute Advantage: Formal Definition {.smaller}

**Setting up the model**

Assumptions: two countries (A, B), two goods (X, Y), one factor (labour), labour mobile within but immobile between countries, constant returns to scale, full employment.

Country A has an **absolute advantage** in Good X if:

$$a_{LX}^A < a_{LX}^B$$

where $a_{LX}$ = labour required to produce one unit of Good X.

*Units of output per worker-day:*

| | Good X | Good Y |
|--|--------|--------|
| Country A | $\frac{1}{a_{LX}^A}$ | $\frac{1}{a_{LY}^A}$ |
| Country B | $\frac{1}{a_{LX}^B}$ | $\frac{1}{a_{LY}^B}$ |

Country A has absolute advantage in X if the value in row A, column X **exceeds** the value in row B, column X.

::: {.highlightbox}
**Smith's prescription:** A has absolute advantage in X → A should *export X*. B has absolute advantage in Y → B should *export Y*. Both specialise → both gain.
:::

## Absolute Advantage: India vs USA (Salvatore Notation) {.smaller}

**Production possibilities (units per worker-day)**

Following Salvatore (2013, Ch. 2) — two nations, two commodities:

| | India | USA |
|---|---|---|
| Rice (units/worker-day) | **4** | 1 |
| Wheat (units/worker-day) | 1 | **3** |

- **Rice:** India produces 4 vs USA's 1 → [India has absolute advantage in Rice]{.hi}
- **Wheat:** USA produces 3 vs India's 1 → [USA has absolute advantage in Wheat]{.hi-gold}

:::{.notes}
**Opportunity costs (= PPF slopes):**

$$\text{India: OC of 1 Wheat} = 4 \text{ Rice} \quad (\text{steep PPF — Rice-abundant})$$
$$\text{USA: OC of 1 Wheat} = \tfrac{1}{3} \text{ Rice} \quad (\text{flat PPF — Wheat-efficient})$$

Each country has **40 worker-days** of labour. PPF equations (linear = constant opportunity cost):
$$\text{India: } R = 160 - 4W \qquad \text{USA: } R = 40 - \tfrac{1}{3}W$$
:::

```{r}
#| label: fig-linear-ppf
#| fig-cap: "Linear PPFs: India and USA — Absolute Advantage (constant opportunity costs)\nSource: Author's illustration."
#| fig-width: 9
#| fig-height: 4.3

# India: L = 40 worker-days. Rice max = 40×4 = 160; Wheat max = 40×1 = 40
# PPF: R = 160 - 4W  (OC of 1 Wheat = 4 Rice → steep slope)
# USA: L = 40 worker-days. Rice max = 40×1 = 40; Wheat max = 40×3 = 120
# PPF: R = 40 - (1/3)W  (OC of 1 Wheat = 1/3 Rice → flat slope)

wheat_india <- seq(0, 40, by = 0.5)
rice_india  <- 160 - 4 * wheat_india

wheat_usa <- seq(0, 120, by = 0.5)
rice_usa  <- 40 - (1/3) * wheat_usa

p1 <- ggplot() +
  geom_line(data = data.frame(w = wheat_india, r = rice_india),
            aes(x = w, y = r), color = "#012169", linewidth = 2) +
  geom_point(aes(x = 20, y = 80), color = "#B9975B", size = 4) +
  annotate("text", x = 22, y = 82,
           label = "Autarky\n(20W, 80R)", size = 3.2, hjust = 0, color = "#B9975B") +
  geom_point(aes(x = 0, y = 160), color = "darkgreen", size = 3) +
  annotate("text", x = 2, y = 155,
           label = "Full specialisation\n(0W, 160R)", size = 3, hjust = 0, color = "darkgreen") +
  annotate("segment", x = 5, y = 140, xend = 10, yend = 120,
           arrow = arrow(length = unit(0.2, "cm")),
           color = "#012169", linewidth = 0.8) +
  annotate("text", x = 11, y = 135,
           label = "OC of Wheat = 4 Rice\n(steep = Rice-abundant)",
           size = 3, hjust = 0, color = "#012169") +
  labs(title    = "India",
       subtitle = "Absolute advantage in Rice (4 units/worker)",
       x = "Wheat (units)", y = "Rice (units)") +
  scale_x_continuous(limits = c(0, 45)) +
  scale_y_continuous(limits = c(0, 175)) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(color = "#012169", face = "bold", size = 13))

p2 <- ggplot() +
  geom_line(data = data.frame(w = wheat_usa, r = rice_usa),
            aes(x = w, y = r), color = "#B9975B", linewidth = 2) +
  geom_point(aes(x = 60, y = 20), color = "#012169", size = 4) +
  annotate("text", x = 62, y = 21,
           label = "Autarky\n(60W, 20R)", size = 3.2, hjust = 0, color = "#012169") +
  geom_point(aes(x = 120, y = 0), color = "darkgreen", size = 3) +
  annotate("text", x = 118, y = 3,
           label = "Full\nspecialisation", size = 3, hjust = 1, color = "darkgreen") +
  annotate("text", x = 30, y = 30,
           label = "OC of Wheat = 1/3 Rice\n(flat = Wheat-efficient)",
           size = 3, hjust = 0, color = "#B9975B") +
  labs(title    = "USA",
       subtitle = "Absolute advantage in Wheat (3 units/worker)",
       x = "Wheat (units)", y = "Rice (units)") +
  scale_x_continuous(limits = c(0, 135)) +
  scale_y_continuous(limits = c(0, 45)) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(color = "#B9975B", face = "bold", size = 13))

p1 + p2 + plot_annotation(
  title    = "Linear PPFs Under Absolute Advantage (Constant Opportunity Costs)",
  subtitle = "Slope of PPF = opportunity cost. India: steep → Rice-abundant. USA: flat → Wheat-efficient. Each exports its absolute advantage good."
)
```

## Gains from Trade: India and USA Under Absolute Advantage {.smaller}

**Without trade — autarky (labour split 50/50)**

| | India | USA | World |
|---|---|---|---|
| Rice (units) | 80 | 20 | **100** |
| Wheat (units) | 20 | 60 | **80** |

**With trade — full specialisation**

| | India | USA | World |
|---|---|---|---|
| Rice (units) | **160** | 0 | **160** |
| Wheat (units) | 0 | **120** | **120** |

World gains: $\Delta \text{Rice} = +60 \text{ units}$ (+60%); $\Delta \text{Wheat} = +40 \text{ units}$ (+50%)

::: {.keybox}
At terms of trade **2 Rice = 1 Wheat** (India exports Rice, imports Wheat): both countries consume **outside** their individual PPFs — trade expands the consumption frontier. Absolute advantage gives each nation a clear signal about where to specialise.
:::

## Numerical Example: India and Bangladesh {.smaller}

**The production possibilities**

| Country | Rice (quintals/worker-day) | Jute (metres/worker-day) |
|---------|----------------|---------------|
| **India** | **4** | **2** |
| **Bangladesh** | **1** | **4** |

- **Rice:** India has absolute advantage. **Jute:** Bangladesh has absolute advantage.

**Before trade (autarky — 50/50 split, 100 workers each):**

| Country | Rice | Jute |
|---------|------|------|
| India | 200 | 100 |
| Bangladesh | 50 | 200 |
| **World total** | **250** | **300** |

**After full specialisation:**

| Country | Rice | Jute |
|---------|------|------|
| India | 400 | 0 |
| Bangladesh | 0 | 400 |
| **World total** | **400 (+60%)** | **400 (+33%)** |

::: {.keybox}
**Both countries consume more** at terms of trade 1:1 — trade is genuinely win-win.
:::

## India's Absolute Advantage in Agriculture {.smaller}

**Where India has clear absolute advantage**

| Commodity | India's global rank | Key advantage |
|-----------|-------------------|---------------|
| Milk production | #1 | 230M+ cattle; Operation Flood legacy |
| Pulse production | #1 | Dryland farming expertise; diverse varieties |
| Spice production | #1 | Tropical climate; accumulated expertise |
| Rice production | #2 (after China) | Diverse agro-climates; Green Revolution |

**Why India has these advantages:** Diverse agro-climatic zones (12 of 15 major global types); abundant agricultural labour (50% of workforce); long cultivation history → accumulated knowledge

::: {.highlightbox}
**Spices: India's oldest absolute advantage.** India produces ~75% of world's spices; exports ~\$3.7B in spices (FY2024). The same advantage that drove European colonial expansion to India's coasts persists today — climate, soils, and accumulated expertise that competitors cannot easily replicate.
:::

## India's Agricultural Productivity: Absolute Advantage Indicators

```{r}
#| label: fig-specialisation-gains
#| fig-cap: "India's Yield Productivity Relative to World Average — Absolute Advantage Indicators\nSource: FAO, FAOSTAT (2023)."
#| fig-width: 8
#| fig-height: 4.5

labour_prod <- data.frame(
  commodity   = c("Basmati Rice", "Non-Basmati\nRice", "Cotton", "Spices\n(Pepper)",
                  "Wheat", "Soybean", "Maize"),
  india_yield = c(3.8, 4.1, 2.1, 2.8, 3.5, 1.2, 3.0),
  world_yield = c(2.9, 4.4, 2.0, 1.5, 3.5, 2.8, 5.8),
  india_adv   = c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE)
)

labour_prod <- labour_prod |>
  arrange(india_yield / world_yield) |>
  mutate(
    commodity = factor(commodity, levels = commodity),
    ratio     = round(india_yield / world_yield, 2)
  )

ggplot(labour_prod, aes(x = commodity, y = ratio, fill = india_adv)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  geom_text(aes(label = paste0(ratio, "x")), hjust = -0.1, size = 3.2) +
  annotate("text", x = 1, y = 1.1,
           label = "World average = 1.0", color = "red", size = 3.5) +
  scale_fill_manual(
    values = c("TRUE" = "#012169", "FALSE" = "#cccccc"),
    labels = c("TRUE" = "India's absolute advantage", "FALSE" = "Below world average")
  ) +
  coord_flip() +
  labs(
    title    = "India's Yield Productivity Relative to World Average",
    subtitle = "Above 1.0x = India produces more per hectare than world average (absolute advantage indicator)",
    x = NULL, y = "India Yield / World Average Yield", fill = NULL
  ) +
  scale_y_continuous(limits = c(0, 2.2)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

::: {.methodbox}
**Reading the chart:** Spices (Pepper) at 1.87x and Basmati Rice at 1.31x show the strongest absolute advantage — India produces far more per hectare than the world average. Maize (0.52x) and Soybean (0.43x) show India well below world average — not where export effort should be directed. Cotton (1.05x) and Non-Basmati Rice (0.93x) are near parity — other factors (price, quality, logistics) determine trade patterns.
:::

## Limitations of Absolute Advantage Theory {.smaller}

**Theoretical limitations**

1. **Ignores the case where one country is better at everything** — If India has absolute advantage in *both* rice and jute, Smith's theory says Bangladesh should not trade → clearly wrong
2. **Labour-only model** — Ignores capital, land, technology; these matter enormously in agriculture
3. **Constant returns to scale** — Real agriculture has diminishing returns (limited fertile land)
4. **Transport costs ignored** — A commodity with a slight absolute advantage may not be worth trading if freight is high

**Practical limitations**

5. **No room for trade policy** — The model says free trade is always optimal; ignores infant industry, food security arguments
6. **No product quality differences** — India exports basmati rice (premium quality) and non-basmati (commodity); model treats all rice as identical
7. **Static model** — Absolute advantage can change over time (South Korea: no electronics advantage in 1960; now it does)

:::{.notes}
Full employment assumption: In reality, India has seasonal unemployment; trade may employ idle labour rather than divert it.
:::

## The Problem with Absolute Advantage

**The critical case: What if one country is better at everything?**

Suppose USA vs India:

| Country | Wheat (tonnes/worker) | Software (units/worker) |
|---------|----------------------|------------------------|
| **USA** | **4** | **8** |
| **India** | **2** | **2** |

USA has absolute advantage in **both** wheat and software. Smith's theory predicts: No basis for trade — India has no absolute advantage in anything.

**But we *know* India and USA trade extensively!** (India exports software services \$80–100B/year + agricultural products)

::: {.highlightbox}
**What Smith missed — comparative advantage:** Even though USA is better at *both* goods, the *degree* of advantage differs. USA is 2× better at wheat (4 vs 2) but **4× better** at software (8 vs 2). USA has a *stronger* advantage in software → USA should specialise in software. India has a *lesser* disadvantage in wheat → India should specialise in wheat. This is **comparative advantage** — David Ricardo's great insight. We prove this formally in Lecture 4.
:::

## Smith vs Mercantilism: A Summary Comparison

::: {.smaller}
| Dimension | Mercantilism | Adam Smith (Absolute Advantage) |
|-----------|-------------|--------------------------------|
| **Source of national wealth** | Stock of gold and silver | Productive capacity; goods and services |
| **Nature of trade** | Zero-sum (one country's gain = another's loss) | Win-win (both countries gain from specialisation) |
| **Objective of trade policy** | Maximise exports; minimise imports | Maximise specialisation; allow free exchange |
| **Role of government** | Active: subsidise exports, restrict imports, grant monopolies | Minimal: allow markets to direct specialisation |
| **Imports: good or bad?** | Bad — gold flows out | Good — access to cheaper goods; frees resources for comparative advantage sectors |
| **Trade balance** | Surplus is the goal | Irrelevant; what matters is consumption possibilities |
| **Colonies** | Essential — raw material source + captive market | Unnecessary and harmful to the coloniser too (waste of resources) |
| **Legacy** | Colonial trade systems; modern currency wars | Foundation of free trade economics; WTO ideology |
:::

## India Policy Application: Mercantilism vs Smith in Real Decisions

**The rice export ban (2023): a case study**

*The mercantilist argument for the ban:*
- Domestic rice prices rising → risk food inflation rising → protect domestic consumers
- Keep rice in India → ensure food security → classic mercantilist hoarding of "strategic resources"

*The Smithian argument against the ban:*
- India has absolute (and comparative) advantage in rice → should be *exporting*, not restricting
- Ban reduces income of 30 million paddy farmers; damages long-term export relationships

**Who won?** The ban was implemented in August 2023 — mercantilist/food security logic prevailed.

::: {.keybox}
India's trade policy oscillates between **export restrictionism** (rice ban, wheat ban, onion export bans) and **export promotion** (FTP 2023–28, APEDA support, RoDTEP). Both impulses co-exist because the political economy is complex: consumers want cheap food; farmers want high prices; government wants foreign exchange.
:::

## The Case of Indian Spices: Absolute Advantage in Action {.smaller}

**India's dominance in global spice markets**

- **Production:** India produces 75% of world's spices by volume; 50%+ by value
- **Export performance (FY2024):** \$3.7 billion; 14% growth over FY2023

| Spice | Export value | Top markets |
|-------|-------------|-------------|
| Chilli | \$982M | USA, Vietnam, Sri Lanka |
| Cumin | \$700M | USA, Bangladesh, China |
| Cardamom | \$350M | Saudi Arabia, UAE |
| Turmeric | \$300M | Bangladesh, USA |
| Pepper | \$190M | USA, Vietnam |

**Source of absolute advantage:** Specific agro-climatic conditions (Kerala for pepper/cardamom, AP/Rajasthan for chilli/cumin) that cannot be replicated at comparable cost elsewhere

::: {.highlightbox}
The search for direct sea routes to India's spice markets was the primary driver of the "Age of Exploration" — Columbus was looking for a route to India's spices when he accidentally reached the Americas. India's spice export advantage today is the same: climate, soils, and accumulated expertise that competitors cannot easily replicate.
:::

## Summary: Mercantilism and Absolute Advantage

::: {.keybox}
**Key points from Lecture 3**

**Mercantilism (1500–1750):**
- Wealth = gold; goal = trade surplus; trade is zero-sum
- Led to colonial exploitation, including India's "drain of wealth"
- Critiqued by Hume (price-specie-flow) and Smith (wealth is production, not gold)

**Absolute Advantage (Adam Smith, 1776):**
- Each country should specialise in what it produces most efficiently
- Trade is win-win — world production rises with specialisation

**The fatal limitation:** What if one country is better at *everything*? Smith cannot explain trade between USA and India. This is the gap Ricardo fills with **comparative advantage**.
:::

:::{.notes}
- India and Bangladesh example: rice vs jute — both gain from specialisation under absolute advantage.
:::

## Key Takeaways: Lecture 3

::: {.methodbox}
**What we covered today**

- **Mercantilism:** Trade as power accumulation; maximise surplus; zero-sum worldview → systematically flawed, but echoes persist in modern policy
- **Colonial trade in India:** British mercantilism extracted surplus, destroyed Indian textile industry → "drain of wealth" (Naoroji)
- **Hume and Smith's critiques:** Trade surpluses are self-correcting (price-specie-flow); real wealth is productive capacity, not gold
- **Absolute advantage:** Specialise in what you produce with highest *absolute* efficiency; trade is win-win
- **Fundamental limitation:** Absolute advantage cannot explain trade when one country is more efficient in *everything* → sets up Ricardo's comparative advantage
:::

:::{.notes}
- **India's absolute advantages:** Rice, spices, pulses, milk, jute — rooted in agro-climatic endowments and accumulated human capital
:::

## Next Lecture: Lecture 4

**Theories of International Trade II — Comparative Advantage**
*(May 19, 2026)*

We will cover:

- **David Ricardo (1817):** The most profound insight in economics — comparative advantage
- **The opportunity cost logic:** Why the *relative* cost, not absolute cost, determines trade patterns
- **Formal model:** Two countries, two goods, with full numerical derivation
- **Production Possibility Frontier (PPF):** Formal diagrammatic treatment
- **Empirical evidence:** Does the Ricardian model predict actual trade patterns?

:::{.notes}
**Suggested reading**

- Krugman, Obstfeld & Melitz, *International Economics*, Ch. 3
- Ricardo, D. (1817). *Principles of Political Economy and Taxation*, Ch. 7 (excerpt)
- Goldin, I. and Reinert, K. (2012). *Globalization for Development*, Ch. 2

**Key question to prepare:**

If India is only half as productive as the USA in *all* agricultural commodities, should India and USA trade at all? If yes, *what* should each produce?
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

**Further Reading**

- *International Economics* — Salvatore (Ch. 2-3)
- *International Economics* — Appleyard & Field (Ch. 2-3)
- RBI/DGCI&S/APEDA databases for latest data

**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
