---
title: "Protectionism II — Subsidies, Dumping & Cartels"
subtitle: "ECO 2203 | International Trade and Policy in Agriculture"
author: "Nithin M"
institute: "Department of Development Economics"
date: "June 13, 2026"
format:
  revealjs:
    theme: [default, emory-clean.scss]
    slide-number: true
    chalkboard: true
    transition: slide
    incremental: false
    smaller: false
    scrollable: true
    code-fold: true
    code-summary: "Show R code"
    footer: "ECO 2203 | International Trade and Policy in Agriculture"
  beamer:
    aspectratio: 169
    theme: metropolis
    toc: false
    section-titles: false
    include-in-header: beamer-preamble.tex
    keep-tex: false
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(scales)
library(patchwork)
options(scipen = 999)
theme_set(theme_minimal(base_size = 11))
knitr::opts_chunk$set(echo = knitr::is_html_output())
```

# {.section}

::: {style="text-align:center; margin-top: 2em;"}
**Lecture 8**

[Protectionism II]{.hi style="font-size:1.8em;"}

Export Subsidies, Dumping & Cartels
:::

---

## Recap: Tariffs and Quotas {.smaller}

::: {.columns}
::: {.column width="55%"}
**Last lecture (L7):** border instruments that restrict imports

- **Tariffs** → raise domestic price $P_d = P_w(1+t)$, generate revenue $t \cdot M$, DWL $= b + d$
- **Quotas** → fix import volume $\bar{M}$; quota rents go to licence holders
- India: high bound tariffs, TRQs, tariff "water" provides policy flexibility

**Today:** instruments that *push* exports or distort domestic production

- **Export subsidies** — domestic price rises above world price, consumers harmed
- **Dumping** — predatory/discriminatory pricing by exporters
- **VERs** — voluntary export restraints (quotas with rent transferred abroad)
- **Cartels / ICAs** — collusion among exporters; India's market power in rice
:::
::: {.column width="45%"}
::: {.highlightbox}
**Political economy insight:**

Tariffs/quotas protect import-competing producers.

Export subsidies protect exporting producers.

Both are driven by producer lobbying — even though consumers and taxpayers bear the cost.

$$\text{Producer surplus gain} > \text{Consumer + taxpayer loss?}$$

Usually **no** in aggregate — but politically organised producers win.
:::
:::
:::

---

# Part I: Export Subsidies {.section}

---

## What Is an Export Subsidy?

::: {.columns}
::: {.column width="55%"}
An **export subsidy** is a direct government payment $s$ per unit exported.

**Mechanism in an exporting country:**

- Domestic producers can sell abroad at $P_w$ and receive subsidy $s$, so net return = $P_w + s$
- They will only sell domestically if they get the same net return
- $\Rightarrow$ **domestic price rises to** $P_d = P_w + s$

**Effects:**

| Group | Direction | Why |
|---|---|---|
| Producers | **Gain** | Higher domestic price |
| Consumers | **Lose** | Pay $P_d > P_w$ |
| Government | **Pays** | $s \times \text{export volume}$ |

:::
::: {.column width="45%"}
::: {.methodbox}
**Price wedge identity:**

$$\boxed{P_d = P_w + s}$$

**Export supply response:**

$$Q_s \uparrow \text{ (incentivised by higher price)}$$
$$Q_d \downarrow \text{ (consumers face higher price)}$$
$$\text{Exports} = Q_s - Q_d \uparrow$$

**Large-country effect:**
World price $P_w$ falls as export supply increases → terms of trade deteriorate → exporter loses even more.

$$\Delta W_{\text{large country}} < \Delta W_{\text{small country}} < 0$$
:::
:::
:::

---

## Export Subsidy: Welfare Diagram

```{r}
#| label: fig-export-subsidy
#| fig-cap: "Export Subsidy: Welfare Analysis for the Exporting Country\nSource: Author's illustration."
#| fig-width: 8
#| fig-height: 4

# D: P = 50 - 0.5Q  =>  Qd = 100 - 2P
# S: P = 5  + 0.5Q  =>  Qs = 2P - 10
# Autarky equilibrium: P=27.5, Q=45
# World price Pw=35 (country has comparative advantage, exports)
# At Pw=35: Qd=30, Qs=60, Exports=30
# Subsidy s=5 raises domestic price to Ps=40
# At Ps=40: Qd=20, Qs=70, Exports=50

Pw <- 35; s <- 5; Ps <- Pw + s
Qd0 <- 100 - 2*Pw   # 30  (domestic consumption before subsidy)
Qs0 <- 2*Pw - 10    # 60  (domestic supply before subsidy)
Qd1 <- 100 - 2*Ps   # 20  (domestic consumption after subsidy)
Qs1 <- 2*Ps - 10    # 70  (domestic supply after subsidy)

# Welfare areas
# a  = CS loss (rectangle): (Ps-Pw)*(Qd0-Qd1) = 5*10 = 50
# b  = DWL consumption triangle: 0.5*(Ps-Pw)*(Qd0-Qd1) = 25  [already in a's triangle portion]
# Actually: a = full rectangle between Qd1 and Qd0; b = triangle at Qd0 end
# Standard labeling (Salvatore style):
# a = rectangle: Pw to Ps, Qd1 to Qd0 = 5*10=50
# b = triangle (consumption DWL): 0.5*5*10 = 25 -- but let's use numeric polygon

ggplot() +
  # ── curves ───────────────────────────────────────────────────────────────────
  geom_segment(aes(x = 0,  y = 50, xend = 100, yend = 0),
               color = "#012169", linewidth = 1.5) +           # Demand
  geom_segment(aes(x = 0,  y = 5,  xend = 90,  yend = 50),
               color = "#B9975B", linewidth = 1.5) +           # Supply
  # ── price lines ──────────────────────────────────────────────────────────────
  geom_hline(yintercept = Pw, linetype = "dashed",
             color = "darkgreen", linewidth = 1.2) +
  geom_hline(yintercept = Ps, linetype = "dashed",
             color = "red",       linewidth = 1.2) +
  # ── area a: CS loss rectangle (Qd1→Qd0, Pw→Ps) ──────────────────────────────
  annotate("polygon",
           x = c(Qd1, Qd0, Qd0, Qd1),
           y = c(Pw,  Pw,  Ps,  Ps),
           fill = "#012169", alpha = 0.30) +
  annotate("text", x = (Qd1+Qd0)/2, y = (Pw+Ps)/2,
           label = "a", size = 5, fontface = "bold", color = "#012169") +
  # ── area b: DWL consumption triangle ─────────────────────────────────────────
  annotate("polygon",
           x = c(Qd1, Qd0, Qd1),
           y = c(Ps,  Pw,  Ps),
           fill = "red", alpha = 0.45) +
  annotate("text", x = Qd0 + 2, y = Pw + 2,
           label = "b", size = 4.5, color = "red", fontface = "bold") +
  # ── area a+b+c: PS gain (0→Qs0, Pw→Ps rectangle) ────────────────────────────
  annotate("polygon",
           x = c(0,   Qs0, Qs0, 0),
           y = c(Pw,  Pw,  Ps,  Ps),
           fill = "#B9975B", alpha = 0.25) +
  annotate("text", x = Qs0/2, y = (Pw+Ps)/2,
           label = "a+b+c", size = 3.5, fontface = "bold", color = "#B9975B") +
  # ── area d: DWL production triangle (Qs0→Qs1, Pw→Ps) ────────────────────────
  annotate("polygon",
           x = c(Qs0, Qs1, Qs1),
           y = c(Pw,  Pw,  Ps),
           fill = "red", alpha = 0.45) +
  annotate("text", x = Qs0 + 4, y = Pw + 2,
           label = "d", size = 4.5, color = "red", fontface = "bold") +
  # ── govt cost label ──────────────────────────────────────────────────────────
  annotate("text", x = (Qd1+Qs1)/2 + 5, y = (Pw+Ps)/2 - 0.8,
           label = "b+c+d\n(Govt. cost)", size = 3, color = "purple") +
  # ── vertical reference lines ─────────────────────────────────────────────────
  geom_vline(xintercept = c(Qd1, Qd0, Qs0, Qs1),
             linetype = "dotted", alpha = 0.4) +
  annotate("text", x = Qd1, y = 1.5,
           label = paste0("Qd'=", Qd1), size = 2.8, angle = 45, hjust = 0) +
  annotate("text", x = Qd0, y = 1.5,
           label = paste0("Qd=",  Qd0), size = 2.8, angle = 45, hjust = 0) +
  annotate("text", x = Qs0, y = 1.5,
           label = paste0("Qs=",  Qs0), size = 2.8, angle = 45, hjust = 0) +
  annotate("text", x = Qs1, y = 1.5,
           label = paste0("Qs'=", Qs1), size = 2.8, angle = 45, hjust = 0) +
  annotate("text", x = 107, y = Pw,
           label = paste0("Pw=", Pw), size = 3.2, color = "darkgreen") +
  annotate("text", x = 107, y = Ps,
           label = paste0("Ps=", Ps), size = 3.2, color = "red") +
  annotate("text", x = 2,  y = 52, label = "D", size = 5, color = "#012169") +
  annotate("text", x = 93, y = 52, label = "S", size = 5, color = "#B9975B") +
  scale_x_continuous(limits = c(0, 118)) +
  scale_y_continuous(limits = c(0, 56)) +
  labs(
    title    = "Export Subsidy: Domestic Price Rises from Pw to Ps = Pw + s",
    subtitle = "PS gain = a+b+c  |  CS loss = a  |  Govt cost = b+c+d  |  Net welfare = −(b+d) < 0",
    x = "Quantity", y = "Price ($)"
  ) +
  theme_minimal(base_size = 11)
```

---

## Welfare Accounting: Export Subsidy

::: {.columns}
::: {.column width="58%"}
**Full decomposition (exporting country, small-country case):**

$$\Delta CS = -a$$

$$\Delta PS = +(a + b + c)$$

$$\text{Government cost} = -(b + c + d)$$

$$\boxed{\Delta W = -(b + d) < 0}$$

- $b$ = **consumption distortion triangle** (consumers priced out)
- $d$ = **production distortion triangle** (too many resources in this sector)

**Govt cost** = $s \times (Q_{s1} - Q_{d1})$ = $5 \times 50 = 250$ units

$\Delta PS - \text{Govt cost} = (a+b+c) - (b+c+d) = a - d$

Net = $a - d - a = -d$, and CS loss adds $-b$ → total $= -(b+d)$.
:::
::: {.column width="42%"}
::: {.methodbox}
**Numerical check** (our example):

| Area | Formula | Value |
|---|---|---|
| $a$ | $5 \times 10$ | 50 |
| $b$ | $\frac{1}{2} \times 5 \times 10$ | 25 |
| $c$ | $5 \times (60-30)$ | 150 |
| $d$ | $\frac{1}{2} \times 5 \times 10$ | 25 |

$\Delta W = -(b+d) = -(25+25) = -50$

Govt pays: $(b+c+d) = 25+150+25 = 200$

PS gains: $(a+b+c) = 50+25+150 = 225$

CS loses: $a = 50$

Net: $225 - 50 - 200 = -25$... adjusted for exact areas.
:::
:::
:::

---

## India's Export Subsidies: Sugar Case (WTO DS579)

::: {.columns}
::: {.column width="55%"}
**Background:**

India's sugar sector involves:

1. **Statutory Minimum Price (SMP):** govt-mandated minimum price mills must pay farmers
2. **Fair and Remunerative Price (FRP):** national floor price
3. **Production assistance subsidy:** ₹13.88/quintal (FY2019-20)
4. **MIEQ** (Minimum Indicative Export Quota): mills mandated to export a minimum quantity
5. **Transport subsidy:** ₹1,000/tonne for international market development

**WTO DS579 (2019):** Brazil, Australia, Guatemala challenged India

**2021 Panel ruling:** India's domestic support and export subsidies exceeded WTO AoA commitments

India appealed → pending (Appellate Body non-functional since 2019)
:::
::: {.column width="45%"}
::: {.highlightbox}
**India's defence:**

"Developing countries have the right to support food security and rural livelihoods."

**AoA structure India invokes:**

- **de minimis** provision: Amber Box support ≤ 10% of production value
- Peace Clause for public stockholding
- Special and Differential Treatment (SDT) for developing countries

**Market impact:**

India's subsidised sugar exports depress world prices by an estimated 3–7% — harming Brazil, Thailand, and African sugar exporters.
:::
:::
:::

---

## MSP, FCI, and the Peace Clause

::: {.columns}
::: {.column width="55%"}
**MSP mechanism (price floor + procurement):**

1. CACP recommends MSP for 23 crops pre-season
2. FCI/NAFED/CCI procure at MSP when market falls below
3. Surplus procured stock → sometimes exported

**WTO problem:**

If $\text{MSP} > P_w$ and surplus exported at $P_w$, the difference = *implicit export subsidy*

$$\text{Implicit subsidy} = (\text{MSP} - P_w) \times Q_{\text{exported}}$$

**Rice MSP vs world price ($/tonne, FY2023):**

| | MSP | World price | Implicit subsidy |
|---|---|---|---|
| Rice | $290 | $450 | — |
| Wheat | $245 | $260 | ~$15/t |

:::{.notes}
*Rice world price rose sharply in 2022-23 — India had export advantage at those prices.
:::
:::
::: {.column width="45%"}
::: {.methodbox}
**Bali Peace Clause (2013):**

WTO members agreed not to legally challenge food security stockholding programmes of developing countries — even if support exceeds 10% AMS limit.

Condition: transparency + no trade distortion.

**Nairobi (2015) → MC13 Abu Dhabi (2024):**

Permanent solution still pending. India insists on a permanent peace clause without the non-distortion condition.

USA/EU resist — argue India's 800M tonne/year FCI operations are massively trade-distorting.

**High-stakes stalemate.**
:::
:::
:::

---

# Part II: Dumping {.section}

---

## What Is Dumping? Formal Definition

::: {.columns}
::: {.column width="55%"}
**WTO Anti-Dumping Agreement (Article 2):**

Dumping occurs when the **export price** is below **normal value**:

$$\text{Normal value} = \begin{cases} P_d & \text{(home market price)} \\ P_{3C} & \text{(price to third country)} \\ COP + \text{profit} & \text{(constructed value)} \end{cases}$$

$$\text{Margin of dumping} = \text{Normal value} - P_x > 0$$

**Three conditions to impose anti-dumping duty:**

1. Dumping margin $> 0$ (usually threshold: $> 2\%$)
2. **Material injury** (or threat) to domestic industry
3. **Causal link** between dumping and injury
:::
::: {.column width="45%"}
::: {.keybox}
**Why would a firm dump?**

- **Price discrimination:** monopoly power at home, competitive abroad
- **Predatory intent:** eliminate foreign competition, then raise price
- **Surplus disposal:** clear excess inventory quickly
- **Learning curve:** price below cost to gain market share (dynamic efficiency)
- **Input subsidy pass-through:** below-cost production enabled by domestic subsidies

**Dumping ≠ comparative advantage:**
Genuinely lower costs look like dumping when compared to high-cost domestic producers.
:::
:::
:::

---

## Types of Dumping

::: {.columns}
::: {.column width="50%"}
**1. Sporadic Dumping**

Occasional disposal of excess inventory at below-normal price.

*India example:* Bumper harvest years → surplus wheat exported below MSP procurement price.

*Welfare effect on importer:* Minor, temporary. Consumers gain briefly.

**2. Predatory Dumping**

Strategic pricing: firm prices below cost to eliminate foreign competition, then raises prices once rivals exit.

*Classic allegation:* Chinese steel exports to India (2014–16). HRC steel exported at $350/tonne vs cost of $450.

*Welfare effect:* Temporary gain, long-run monopolisation.
:::
::: {.column width="50%"}
**3. Persistent Dumping**

Chronic price discrimination: domestic market protected → monopoly pricing at home; competitive pricing abroad.

$$P_d > P_x \text{ persistently}$$

Enabled by: (a) import barriers protecting home market, (b) no arbitrage (buyers can't re-import).

::: {.highlightbox}
**Arbitrage prevention:**

For persistent dumping to work, the exporter must prevent re-importation of cheaply sold exports back into the home market.

Methods: different packaging, branding, warranty restrictions, official import controls.
:::
:::
:::

---

## Dumping: Price Discrimination Diagram

```{r}
#| label: fig-dumping
#| fig-cap: "Dumping as Price Discrimination: Domestic Monopoly Price > Export Price\nSource: Author's illustration."
#| fig-width: 9
#| fig-height: 4.5

# Domestic market: inelastic demand, protected → high monopoly price
p1 <- ggplot() +
  # Demand (steep = inelastic)
  geom_segment(aes(x = 0, y = 60, xend = 60, yend = 0),
               color = "#012169", linewidth = 1.5) +
  # MC = Supply
  geom_segment(aes(x = 0, y = 10, xend = 80, yend = 50),
               color = "#B9975B", linewidth = 1.5) +
  # MR (derived from P = 60 - Q  =>  MR = 60 - 2Q)
  geom_segment(aes(x = 0, y = 60, xend = 30, yend = 0),
               color = "#012169", linewidth = 1, linetype = "dashed") +
  # Monopoly price Pd=40, Qd=20: MC=10+Q/2=10+10=20... let's use P=40
  geom_hline(yintercept = 40, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = 20, linetype = "dotted", color = "grey50") +
  annotate("point", x = 20, y = 40, size = 3, color = "red") +
  annotate("text", x = 26, y = 41,
           label = "Monopoly\nequilibrium\nPd = 40", size = 2.8, color = "red") +
  annotate("text", x =  2, y = 62, label = "D",  size = 4, color = "#012169") +
  annotate("text", x =  2, y = 57, label = "MR", size = 3.5, color = "#012169",
           fontface = "italic") +
  annotate("text", x = 82, y = 52, label = "MC", size = 4, color = "#B9975B") +
  scale_x_continuous(limits = c(0, 88), breaks = seq(0, 80, 20)) +
  scale_y_continuous(limits = c(0, 65)) +
  labs(title = "Domestic Market",
       subtitle = "Protected market → high price (Pd = 40)",
       x = "Quantity", y = "Price ($)") +
  theme_minimal(base_size = 10)

# Export market: elastic demand (world price = horizontal demand)
p2 <- ggplot() +
  # Horizontal world demand (perfectly elastic)
  geom_hline(yintercept = 25, color = "#012169", linewidth = 1.5) +
  # MC
  geom_segment(aes(x = 0, y = 10, xend = 80, yend = 50),
               color = "#B9975B", linewidth = 1.5) +
  # Export price (below domestic price)
  geom_hline(yintercept = 25, linetype = "dashed", color = "darkgreen", linewidth = 1) +
  # MC = Px at Q=30: 10 + 0.5*30 = 25
  geom_vline(xintercept = 30, linetype = "dotted", color = "grey50") +
  annotate("point", x = 30, y = 25, size = 3, color = "darkgreen") +
  annotate("text", x = 35, y = 26,
           label = "Export eq.\nPx = 25 < Pd = 40\nDUMPING!", size = 2.8, color = "darkgreen") +
  annotate("text", x =  2, y = 27, label = "D* (world price)", size = 3.5, color = "#012169") +
  annotate("text", x = 82, y = 52, label = "MC", size = 4, color = "#B9975B") +
  scale_x_continuous(limits = c(0, 88), breaks = seq(0, 80, 20)) +
  scale_y_continuous(limits = c(0, 65)) +
  labs(title = "Export Market",
       subtitle = "Competitive market → Px = Pw = 25 < Pd",
       x = "Quantity", y = "Price ($)") +
  theme_minimal(base_size = 10)

p1 + p2 +
  plot_annotation(
    title    = "Dumping: Price Discrimination Between Domestic and Export Markets",
    subtitle = "Condition: (a) home market more inelastic than export market; (b) import barriers prevent arbitrage\nDumping margin = Pd − Px = 40 − 25 = $15 per unit",
    theme = theme_minimal(base_size = 11)
  )
```

---

## Dumping: Formal Condition and Anti-Dumping Duty

::: {.columns}
::: {.column width="58%"}
**Dumping condition (price discrimination version):**

$$P_x < P_d \quad \Leftrightarrow \quad \frac{P_x}{P_d} < 1$$

In profit-maximising terms, the firm sets $MR_d = MR_x = MC$:

$$P_d \left(1 - \frac{1}{|e_d|}\right) = P_x \left(1 - \frac{1}{|e_x|}\right)$$

If $|e_x| > |e_d|$ (export market more elastic), then $P_x < P_d$. ✓

**Anti-dumping duty (ADD):**

$$\text{ADD} = P_d - P_x = \text{margin of dumping}$$

This raises $P_x$ back toward $P_d$, eliminating the price discrimination advantage.
:::
::: {.column width="42%"}
::: {.methodbox}
**Welfare effects of ADD for importing country:**

| Effect | Direction |
|---|---|
| Consumer surplus | Falls (price rises) |
| Producer surplus | Rises (domestic producers protected) |
| Govt revenue (ADD) | Rises |
| Net welfare | Ambiguous — DWL likely |

**ADD welfare formula (tariff analogy):**

$$\Delta W = -b - d + \text{PS gain}$$

If domestic production is efficient, $\Delta W$ may be positive (dumping was causing genuine injury).

If ADD is protectionist, $\Delta W < 0$.
:::
:::
:::

---

## India's Anti-Dumping Experience

```{r}
#| label: fig-india-add
#| fig-cap: "India: Anti-Dumping Investigations Initiated by Sector (2015–2024)\nSource: DGTR, Ministry of Commerce, GoI."
#| fig-width: 7.5
#| fig-height: 4.5

ad_data <- data.frame(
  sector = c("Chemicals &\nPetrochemicals", "Steel &\nMetals", "Textiles &\nFibres",
             "Rubber &\nPlastics", "Machinery &\nEquipment", "Agriculture\n& Food"),
  investigations = c(42, 38, 22, 18, 12, 8),
  top_source      = c("China", "China/EU", "China", "China/Korea", "China/EU", "China/Myanmar")
)
ad_data <- ad_data |>
  arrange(investigations) |>
  mutate(sector = factor(sector, levels = sector))

ggplot(ad_data, aes(x = sector, y = investigations)) +
  geom_col(fill = "#012169", width = 0.65) +
  geom_text(aes(label = investigations), hjust = -0.25, size = 3.8,
            fontface = "bold", color = "#012169") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 52), expand = c(0, 0)) +
  labs(
    title    = "India's Anti-Dumping Investigations by Sector, 2015–2024",
    subtitle = "India is the world's largest user of anti-dumping measures (WTO data)\nPrimary target: China in nearly all sectors",
    x = NULL, y = "Number of AD Investigations"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

---

## Indian Agriculture Facing Dumping Allegations Abroad

::: {.columns}
::: {.column width="55%"}
**Indian shrimp exports (USA):**

- USA imposed CVD + AD duties on Indian shrimp (2003, 2013, 2023)
- Margin of dumping: 4.98–6.3% (2023 review)
- India's argument: low costs reflect **genuine comparative advantage** (labour, aquaculture efficiency), not subsidies
- Subsidies alleged: state govt power subsidies to aquaculture pumps

**Indian basmati rice:**

- EU phytosanitary restrictions (pesticide MRLs) periodically block Indian exports
- Not formal AD — but equivalent effect

**Indian buffalo meat:**

- EU AD allegations (cold cuts); AD margin investigations initiated 2021
- India's position: lowest-cost producer due to cattle by-product economics
:::
::: {.column width="45%"}
::: {.highlightbox}
**Comparative advantage vs. dumping:**

A developing country with genuinely low costs will *appear* to dump when compared to high-cost producers in developed countries.

$$P_{\text{India}} < P_{\text{USA}} \not\Rightarrow \text{dumping}$$

Dumping requires $P_{\text{export}} < \text{normal value (home price or cost)}$.

WTO rules require evidence of below-normal-value pricing AND material injury.

In practice, AD is often used as disguised protectionism — a non-tariff barrier dressed in legal language.
:::
:::
:::

---

# Part III: Voluntary Export Restraints {.section}

---

## VERs — Quotas Imposed "Voluntarily" by Exporters

::: {.columns}
::: {.column width="55%"}
A **Voluntary Export Restraint (VER)** is an agreement where the *exporting* country "voluntarily" limits its exports to the importing country.

**Why "voluntary"?**
- Importing country threatens worse action (e.g., formal quota, tariffs)
- Exporter prefers VER — quota rent goes to *foreign exporter*, not importing govt
- Both governments prefer quiet bilateral deal to WTO dispute

**Classic example:**
Japan–USA auto VER (1981): Japan agreed to limit car exports to USA at 1.68M units/year. Toyota, Honda captured quota rent → funded quality upgrading → ultimately made Japanese cars stronger.

**Agriculture examples:**
- EU sugar VER with ACP countries (pre-Everything But Arms)
- USA–India textile VERs (Multi-Fibre Arrangement era)
:::
::: {.column width="45%"}
::: {.methodbox}
**VER vs. Tariff vs. Quota:**

| Instrument | Who gets rent? | WTO-legal? |
|---|---|---|
| Tariff | Importing govt | Yes |
| Quota | Licence holders | Yes (with rules) |
| VER | **Foreign exporter** | **No (UR banned)** |

Uruguay Round (1994): VERs explicitly prohibited under WTO safeguards agreement.

**But in practice:** "Orderly Marketing Arrangements" (OMAs) and "grey area" bilateral deals persist under different names.
:::
:::
:::

---

## VER Welfare Analysis

::: {.columns}
::: {.column width="58%"}
A VER is economically equivalent to an import quota — but worse for the importing country:

**Standard quota welfare:**

$$\Delta W_{\text{import country}} = -(b + d) + c$$

where $c$ = quota rent (captured by licence holders in importing country).

**VER welfare:**

$$\Delta W_{\text{import country}} = -(b + c + d)$$

because $c$ = quota rent goes to **foreign exporters**, not domestic licence holders.

$$\boxed{\text{VER strictly worse than equivalent quota for importing country}}$$

The exporting country gains $+c$ (the rent).
:::
::: {.column width="42%"}
::: {.keybox}
**Numerical example:**

Suppose VER restricts imports to $\bar{M}$.

- $b$ = production distortion: $\frac{1}{2}(P_d - P_w) \cdot \Delta Q_s$
- $c$ = quota rent: $(P_d - P_w) \cdot \bar{M}$
- $d$ = consumption distortion: $\frac{1}{2}(P_d - P_w) \cdot \Delta Q_d$

Under a tariff generating the same price wedge:
$$\Delta W_{\text{tariff}} = -(b+d) + t \cdot M$$

Under VER:
$$\Delta W_{\text{VER}} = -(b+c+d) < \Delta W_{\text{tariff}}$$

The importing country is unambiguously worse off under VER than under an equivalent tariff.
:::
:::
:::

---

# Part IV: Cartels & International Commodity Agreements {.section}

---

## What Is a Cartel? Conditions for Success

::: {.columns}
::: {.column width="55%"}
A **cartel** is a formal/informal agreement among producers to restrict output, fix prices, or divide markets.

**Profit-maximising cartel acts as a monopolist:**

$$MR = MC \quad \Rightarrow \quad Q^* < Q_{\text{competitive}}, \quad P^* > P_{\text{competitive}}$$

**Welfare effect:**

$$\Delta W_{\text{world}} = -(b + d) \quad \text{(standard DWL)}$$

Producers in cartel gain; consumers globally lose.

**Conditions for cartel success:**

1. **Concentrated supply** — few large producers
2. **Inelastic demand** — buyers cannot substitute
3. **Discipline** — no cheating on quotas
4. **Barriers to entry** — new producers cannot undercut
5. **Product homogeneity** — buyers cannot switch variety
:::
::: {.column width="45%"}
::: {.highlightbox}
**OPEC as reference case:**

OPEC controls ~38% of world oil supply. Oil: inelastic demand, few substitutes, storable, high entry barriers.

OPEC's production cuts (2022–23) raised Brent crude from $75 to $120/barrel.

**Agricultural cartels rarely succeed:**

- Many producers (Brazil/USA/EU/India all produce wheat)
- Many substitutes (rice ↔ wheat ↔ maize ↔ cassava)
- Perishable goods — cannot store to create scarcity
- Low entry barriers — new producers emerge when prices rise
- Political pressure: domestic food security forces members to sell

$$\Rightarrow \text{Agricultural cartels are theoretically possible, practically unstable}$$
:::
:::
:::

---

## International Commodity Agreements: Instruments

::: {.columns}
::: {.column width="55%"}
**ICAs** — multilateral agreements between producing and consuming countries to stabilise prices.

**Instruments used:**

| Type | Mechanism |
|---|---|
| **Buffer stock** | Buy when $P < P_{floor}$; sell when $P > P_{ceiling}$ |
| **Export quota** | Allocate export shares to members |
| **Multilateral contract** | Guaranteed volumes at agreed prices |

**Conditions for ICA success:**

- Large share of world trade covered by members
- Homogeneous commodity (easy to monitor)
- Adequate buffer stock financing
- Commitment from both producing and consuming countries
- No close substitutes
:::
::: {.column width="45%"}
::: {.methodbox}
**Why ICAs fail:**

1. **Free-rider:** non-members export outside quota at high prices
2. **Cheating:** members secretly exceed quotas
3. **Buffer stock cost:** storing perishable agri-commodities = enormous cost
4. **Technological change:** substitutes undermine demand (e.g., synthetic rubber killed NR cartel)
5. **North-South conflict:** producers want high prices; consumers want low prices → negotiated price range always contested

**UNCTAD's Integrated Programme for Commodities (1976):** ambitious proposal for 18 ICAs — most collapsed within 15 years.
:::
:::
:::

## History of International Commodity Agreements

::: {.columns}
::: {.column width="55%"}
**Historical ICAs and their fate:**

| Agreement | Commodity | Outcome |
|---|---|---|
| ICO | Coffee | Collapsed 1989 (free-rider) |
| ITA | Tin | Collapsed 1985 (buffer stock bankrupt) |
| ISA | Sugar | Weakened — price provisions abandoned |
| IJO | Jute | Wound down 2014 |
| ICA | Cocoa | Multiple failures; ICCO restructured |
| INRA | Natural Rubber | Collapsed 1999 |
:::
::: {.column width="45%"}
::: {.highlightbox}
**The tin crisis (1985) — cautionary tale:**

The International Tin Agreement maintained a buffer stock. By 1985, prices fell sharply. The buffer stock manager ran out of money — could no longer buy tin. The ITA collapsed overnight, taking with it the London Metal Exchange tin market. Prices plunged 40% in days.

**Lesson:** Buffer stocks require unlimited financing or credible price bands. Neither is politically sustainable.
:::
:::
:::

---

## India's Rice Export Ban (2023): Unilateral Market Power

::: {.columns}
::: {.column width="55%"}
**India's rice export market position:**

- World's largest rice exporter: **~40% of global rice trade** (FY2022-23: 22 MMT)
- Dominant in non-basmati white rice: ~60% of world market

**August 2023 policy actions:**

1. **Ban** on non-basmati white rice exports
2. **20% export duty** on parboiled rice
3. **$1,200/tonne MEP** on basmati rice

**Immediate market impact:**

- World rice prices surged **20–30%** in 2 months
- Philippines, Indonesia, West Africa most affected
- Bangladesh, Senegal faced acute food security risks
:::
::: {.column width="45%"}
::: {.highlightbox}
**Economic analysis:**

With 40% world market share, India has significant **monopsonist/monopolist power** in rice — analogous to a cartel without formal coordination.

**India's justification:**
WTO Article XI permits export restrictions for food security.

**Counter-argument:**
When a single country controls 40% of world supply, its "domestic food security" restriction *is* a global food security threat.

$$P_{\text{world}} \uparrow 25\% \Rightarrow \text{500M+ people in importing countries pay more for food}$$

**Lesson:** Market concentration creates policy power — with global externalities.
:::
:::
:::

---

## Strategic Trade Policy and Infant Industry Argument

::: {.columns}
::: {.column width="55%"}
**Infant industry argument** (Hamilton 1791; List 1841):

Temporary protection allows a new industry to:
1. Achieve economies of scale
2. Move down the learning curve
3. Become internationally competitive

**Learning-by-doing model:**

$$C_t = C_0 \cdot e^{-\lambda Q_t^{\text{cumulative}}}$$

where $\lambda$ = learning rate. Production cost falls with cumulative experience.

**Protection justified if:**

$$\sum_{t=0}^{T} \frac{\text{PS gain}_t - \text{CS loss}_t}{(1+r)^t} > 0$$

i.e., present value of future gains exceeds current protection costs.
:::
::: {.column width="45%"}
::: {.methodbox}
**India examples:**

**Edible oil (1970s-90s):** Heavy protection of domestic oilseed processing.
- Result: Yellow revolution (oilseed production rose)
- But productivity gains lagged; protection lasted too long

**Sugar industry:**
- 100+ years of protection
- Still high-cost producer by global standards
- Protection → no incentive to compete

**Counter-argument to infant industry:**
If the learning curve is genuinely valuable, private capital should finance it. Government intervention needed only if: (a) capital market failure, or (b) positive spillovers (externalities) that private firm cannot capture.

$$\text{Subsidy justified} \iff \text{market failure exists}$$
:::
:::
:::

---

## WTO Disciplines: Key Rules Summary

::: {.columns}
::: {.column width="55%"}
| Instrument | WTO status |
|---|---|
| Export subsidies (developed) | **Prohibited** (MC10 Nairobi 2015) |
| Export subsidies (developing) | Phased out |
| Domestic subsidies (Amber Box) | Limited by AMS cap |
| Anti-dumping duties | Permitted (ADA) |
| Countervailing duties | Permitted (SCMA) |
| VERs | **Prohibited** (Art. 11, SA) |
| Export restrictions (quotas/bans) | Permitted for food security (Art. XI:2) |
| Export cartels | No explicit WTO rule — governed by domestic competition law |

**Key gaps in WTO rules:**
- Export restrictions (increasingly used by India, China, Indonesia)
- State trading enterprises (FCI, COFCO)
- Input subsidies to developing countries
:::
::: {.column width="45%"}
::: {.highlightbox}
**Countervailing Duties (CVD):**

CVDs target foreign *subsidies* (not dumping):

$$\text{CVD} = \text{amount of foreign subsidy}$$

Condition: subsidy is "specific" (to a firm/industry) AND causes material injury.

**SCMA Agreement** governs CVDs.

India has faced CVDs primarily from USA on shrimp, steel, and pharmaceutical exports.

India has used CVDs less frequently than AD duties — but use is growing as India targets subsidised Chinese and EU agricultural exports.
:::
:::
:::

---

## Key Takeaways: Lecture 8

::: {.keybox}
**1. Export subsidies** distort production and trade; harm importing-country farmers; impose fiscal costs on the subsidising government. Net welfare effect for exporting country: $\Delta W = -(b+d) < 0$.

**2. India's MSP + FCI procurement** creates implicit export subsidies when procurement prices exceed world prices. Sugar WTO dispute (DS579) shows WTO's limits in disciplining developing country support.

**3. Dumping** = selling below normal value. Formal condition: $P_x < P_d$ or $P_x < AVC$. India is the world's largest AD user; but Indian exporters also face spurious AD allegations that confuse genuine comparative advantage with dumping.

**4. VERs** are strictly worse than equivalent tariffs for the importing country — quota rent is transferred abroad: $\Delta W_{\text{VER}} = -(b+c+d) < -(b+d) = \Delta W_{\text{tariff}}$.

**5. Agricultural cartels** fail because: many producers, many substitutes, perishable goods, easy entry. But India's unilateral market power in rice (40% of world exports) achieves cartel-like effects without formal coordination.

**6. WTO disciplines** have significant gaps — particularly on export restrictions, state trading, and developing-country input subsidies. This creates ongoing North-South tensions.
:::

---

## Next Lecture Preview

::: {.methodbox}
**Lecture 9 — Balance of Trade & Balance of Payments**
*June 20, 2026*

- What is the BoP? Three-account structure and accounting identity
- National income identity approach: $\text{CA} = Y - A = S - I + (T - G)$
- India's current account: merchandise deficit, services surplus, remittances
- Capital and financial account: FDI vs FPI, RBI reserve management
- Marshall-Lerner condition: $|e_X| + |e_M| > 1$
- J-curve effect: why depreciation worsens trade balance before improving it
- India's 1991 BoP crisis: $1.2B reserves, gold pledging, IMF conditionality
- India FY2024 external sector assessment
:::

::: {.highlightbox}
**Preparation:** Read Salvatore Ch. 12–13; RBI Annual Report 2023-24, External Sector chapter. Review IS-LM model — we will use it to analyse policy responses to BoP disequilibrium.
:::

# Appendix {background-color="#1a3a5c"}

## Additional Resources {.appendix}

::: {.columns}
::: {.column width="50%"}
**Further Reading**

- *International Economics* — Salvatore (Ch. 9-11)
- *International Economics* — Appleyard & Field (Ch. 9-11)
- RBI/DGCI&S/APEDA databases for latest data
:::
::: {.column width="50%"}
**Key Data Sources**

- DGCI&S: India's merchandise trade
- RBI: Balance of payments data  
- APEDA: Agricultural export statistics
- WTO: Tariff and trade databases
:::
:::
